### 2022-07-02 Preprint-Differentiation.Rmd
Added .Rmd file
Added replication .Rmd file plus a folder including all figures.

---
title: "Does differentiation beget differentiation?"
subtitle: Examining the link between exposure to and support for differentiated integration
author: "Martin Moland^[ARENA, Centre for European Studies]"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
abstract: "Questions about the public legitimacy of the EU's current flexible integration are becoming increasingly important in the public opinion literature. However, we still know little about how exposure to past differentiation shapes public preferences for a more differentiated EU. Combining survey data from 2020-21 with data detailing EU countries’ differentiations from the EU treaties I show that citizens of countries previously exposed to differentiations from the EU’s legal order are more likely to want a less uniform EU integration in the future. This has implications both for the debate about the shape of the future EU and the broader literature investigating public support for transnational cooperation."
csl: european-journal-of-political-research.csl
bibliography: bib_file.bib
header-includes: 
  - \usepackage{setspace}\doublespacing
  - \usepackage{float}
fontsize: 12pt

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction

Even if the EU is among the world’s most advanced political cooperations, one of the main stories of the EU’s recent development is its increasingly fragmented integration of particularly salient areas [@leuffen_differentiated_2013;@schimmelfennig_ever_2020]. The EU’s political integration has typically been differentiated for two reasons: First, some member states have opted out of cooperation to protect their own sovereignty. This is called “constitutional differentiation”. Second, through a process known as “instrumental differentiation”, countries have integrated into policy cooperation according to different timetables. Both modes of differentiation can either be chosen by the member states themselves or, as was the case for Central and Eastern Europe, externally imposed [@schimmelfennig_instrumental_2014]. Because of these forms of differentiation, the EU’s current shape is best characterised as one of ‘variable geometry’ [@stubb_categorization_1996] rather than ‘ever closer Union’.

While we increasingly know a lot about why people support such deviations from uniform integration [@telle_elite-mass_2022-1; @leuffen_public_2020], we do not know whether living in a country that has had a diverging path of integration from that taken by the other EU member states shapes current preferences for a less uniform EU. While one of the few existing contributions examining the causal effect of differentiation on support for EU democracy finds a positive correlation between the two [@schraff_does_2020], two knowledge gaps remain: First, we know too little about how past experiences with differentiation changes public preferences for a future differentiated EU. Second, we do not know whether the mode of implementation of differentiation affects these preferences. This paper seeks to fill those gaps.

By combining survey data from 2020-2021 on citizens’ attitudes towards constitutional differentiated integration with data detailing their countries’ differentiations from EU treaties, this paper finds that support for differentiated integration is greater among citizens whose countries have had a greater number of differentiations from EU treaties. This is especially the case for Eurosceptics. This holds both where the differentiation was voluntary, as in Scandinavia [@adler-nissen_opting_2014] or externally imposed by the EU’s institutions, as in Central and Eastern Europe [@schimmelfennig_eu_2014]. However, there is one exception: While Eurosceptics living in countries with much previous exposure to past treaty differentiation are broadly more likely to support further differentiation, this does not apply to Central and Eastern Europe. Here Euroscepticism and exposure to differentiated integration seems to produce less support for a more differentiated future EU. I hypothesize that this is either driven by how differentiation has been framed by Central and Eastern European politicians as a threat to the influence of these countries within the relevant states or a historically driven scepticism of a differentiation that has, in these countries, primarily been externally imposed. Thus, my results offer contradictory evidence for a self-reinforcing feedback loop [@busemeyer_beyond_2021] between exposure to differentiation and future support for it: While the main thrust of evidence suggests that greater exposure to differentiation strengthens popular assent, there are also signs of the opposite. This shows that preferences are likely to be determined by a complex interplay of individual characteristics and regional histories of integration.

My findings are relevant to the discussion about the EU's future that Brexit gave rise to [@european_commission_white_2017]: As differentiated integration has become a more important tool of EU governance, especially given the EU’s increasing integration of for instance monetary and fiscal policy [@schimmelfennig_ever_2020; @genschel_differentiated_2013], greater exposure to it among member state citizens may make it more difficult for the EU’s institutions to consolidate support for the project of further integration. The article also makes an important contribution to the emerging literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009], by showing that implementation of differentiations from international law can potentially increase public demand for future divergences from the same legal order. Lastly the paper shows that differentiation, favoured as a tool for mitigating the contestation provoked by the EU’s increasing integration of particularly salient policy areas [@genschel_market_2018], may complicate the EU’s progress towards its goal of “ever closer Union”.

# Public preferences for European integration - what we know and what we don't

A large literature investigates individual preferences for European integration [@basile_sharing_2020; @bremer_junckers_2020; @de_vries_variable_2013; @eichenberg_post-maastricht_2007; @gabel_public_1998; @hobolt_public_2015; @hooghe_calculation_2005]. The literature identifies three mechanisms that shape preferences for European integration. The first is utilitarian, with support for the EU contingent on whether the individual or social groups with which they identify potentially gain from integration [@foster_economic_2021; @gabel_public_1998]. The second is cue-based, and posits that citizens use informational shortcuts such as elite and media cues to make up their minds about European integration, particularly in less salient policy areas [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007]. The literature also shows that strong identification with the nation-state makes citizens more likely to oppose EU integration [@borzel_identity_2020; @hooghe_calculation_2005; @luedtke_european_2005].

However, we know much less about why people support a less uniform European Union. This is despite the fact that a large literature investigates differentiated integration, as this form of flexible integration is commonly called, conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014; @schimmelfennig_ever_2020], empirically [@ganzle_differentiated_2019; @malang_political_2020; @schimmelfennig_eastern_2017; @winzen_government_2020; @ott_between_2017] and normatively [@bellamy_ever_2013; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004]. The few existing contributions probing the question of individual support for differentiated integration points to either liberal economic values [@leuffen_public_2020], Euroscepticism [@de_blok_blessing_2020] and party cueing [@telle_elite-mass_2022-1] as important predictors.

This literature, in other words, mostly focuses on predictors found at the level of individuals. However, popular attitudes towards both the EU and international cooperation more broadly are often shaped by elite cues and context [@dellmuth_elite_2021; @edwards_public_2009; @harteveld_europe_2013; @telle_elite-mass_2022-1]. One under-explored source of such support is the large variation in differentiated integration found across the EU [@schimmelfennig_instrumental_2014]: As either externally imposed or voluntarily chosen differentiation may be either in or out of step with popular attitudes towards how the EU should look, past differentiation may face a popular response in the form of lesser or greater support for differentiation [@busemeyer_beyond_2021; @wlezien_public_1995]. While previous literature has investigated how this feedback mechanism plays out in the case of EU policies [@bolstad_dynamics_2015; @de_bruycker_democratically_2020], this paper will instead investigate how differentiations from the EU's primary legal framework shapes demand for a differently configured polity.

The effect of exposure to differentiation may also be compounded by who originally initiated differentiations, and the circumstances under which it happened. An interesting test of this assumption is support for differentiation in Central and Eastern Europe. Here citizens were exempt from the right to move across EU borders to seek work for several years after their countries acceded to the EU. Differentiation was thus externally imposed as a precondition for EU membership [@cianciara_differentiated_2014; @schimmelfennig_eu_2014]. This was done because existing member states wanted to shield themselves from the economic destabilization that they feared would follow from the EU accession of countries like Poland, Estonia, and Hungary [@schimmelfennig_eu_2014]. This could easily be seen by those excluded from these policies as a cartelization of the benefits of EU integration [@lord_autonomy_2021 547]. Citizens of Central and Eastern Europe could thus respond to the involuntary nature of the exposure they have had to differentiated integration by expressing lower support for it in the future.

However, another causal mechanism could explain lower levels of support for differentiated integration in Central and Eastern Europe without relying on a path dependency argument: Central and Eastern European politicians have framed differentiated integration as a path towards an EU consisting of a powerful core and a less powerful periphery. This frame was prominent during the sovereign debt crisis [@cianciara_differentiated_2014 178-179]. For example, Polish politicians opposed further integration of the Eurozone, a clear example of differentiated integration, because they feared that the Eastern member states would become less influential in economic policy-making due to being non-Eurozone countries. Such cues have been known to shape popular attitudes towards integration, particularly where citizens have little knowledge of the relevant policies. It may thus help explain possible opposition to differentiated integration [@steenbergen_whos_2007; @telle_elite-mass_2022-1].

However, increasingly Eurosceptic elite attitudes in Central and Eastern Europe [@borzel_identity_2020; @brack_towards_2020; @kriesi_politicization_2016] could produce the opposite outcome: Greater support for differentiated integration. The reason is once again elite cueing: Where European integration is increasingly framed as a problem, citizens may see less uniform integration as a solution and thus increase their demand for more differentiation. Here, in other words, the causal mechanism would not necessarily be exposure to differentiated integration, but rather how that same differentiation is framed. This yields two competing hypotheses:  
*H1: That support for constitutional differentiated integration will be lower in Central and Eastern countries with more extensive exposure to differentiated integration compared to countries with lower levels of exposure.*  
*H2: That support for constitutional differentiated integration will be higher in Central and Eastern European countries with more extensive exposure to differentiated integration compared to countries with lower levels of exposure.*

Normative literature on differentiated integration [@bellamy_republican_2019; @cheneval_case_2013] has argued that differentiated integration may allow for greater national autonomy. The same argument has been used by Swedish and Danish governments to justify their decisions to opt out of for instance the economic and monetary union [@adler-nissen_opting_2014; @leruth_operationalizing_2015; @spendzharova_banking_2016] Exposure to such cues could manifest itself in greater popular support for constitutional differentiated integration in these countries, and produce the opposite result from that seen in Central and Eastern Europe: As differentiation through opt-outs has explicitly and continuously been framed as a tool for expanding the sovereignty of both Sweden and Denmark, it may enjoy greater support among Scandinavian citizens than what is found in other regions.

Evidence of this is found in how the renewed Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy on the part of those identifying as Eurosceptic [@schraff_does_2020]. This suggests that citizens may respond positively to differentiation, and could indicate that they would be favourably inclined towards further differentiation. This trend might also be particularly strong among Eurosceptics in these countries: These citizens may, because of their countries’ relatively small sizes, be more likely to see differentiation as a viable second-best alternative to the European disintegration they would otherwise have favoured. This could make them more likely than supporters of EU membership to favour further differentiation. I formulate the following hypotheses:  
*H3: The positive effect of exposure to differentiated integration will be greater for citizens of Sweden and Denmark than in the entire EU.*  
*H3a: The increase in support attributable to increased exposure to differentiated integration will be greater for Scandinavian Eurosceptics than supporters of EU membership.*

It is also likely that the Eurosceptic sentiments found to be important for predicting support for differentiated integration [@de_blok_blessing_2020] will compound the effect of exposure to differentiation. This may be because Eurosceptics are arguably more likely than supporters of EU membership to see differentiated integration as a preferable alternative to current uniform integration. This yields the following hypothesis:  
*H4: Support for constitutional differentiated integration will increase more strongly as a function of previous exposure to differentiated integration among Eurosceptics than among those who support EU membership.*

My paper contributes to the emerging literature on public support for differentiated integration by moving it beyond the current focus on individual-level characteristics [@de_blok_blessing_2020; @leuffen_public_2020]. It does so by shifting the focus to how a key contextual factor, the actual level of differentiation, impacts support for constitutional differentiated integration under different modes of implementation. It thus sheds an important light on whether support for a more differentiated EU is impacted by the actual levels of differentiation that citizens have historically been exposed to. This is also important for explaining how path dependencies brought by past international cooperation helps shape support for international organizations' future choices. It thus answers the question of whether differentiation begets further differentiation by solidifying support for a less uniform EU.

# Data and methods
The hypotheses will be investigated using multilevel linear models and data from Yougov surveys fielded in 2020-21 [@hemerijck_eui-yougov_2021]. I combine these data with the expanded EUDIFF1 dataset [@schimmelfennig_eudiff1_2022]. This dataset features all distinct differentiations from the EU treaties, showing for each country the number of articles in which differentiation was given, the years they were in force and the policy areas covered. The surveys, on the other hand, ask a range of questions related to European integration. Two of these ask about preferences for differentiated integration. The first question asks people whether they want an EU where countries may opt out of certain policy integration. This may be termed a constitutionally differentiated EU. The second asks whether the EU should allow countries to arrive at the same level of integration at different speeds when they are ready. This would be an instrumentally differentiated EU. This article focuses on support for constitutional differentiated integration, whose permanent segmentation represents arguably the strongest departure from the EU's current goal of continuously closer integration. Because it allows for both fragmentation with implications for each country's power within the union as well as greater space for national autonomy, attitudes towards  differentiation are likely to be shaped by both sovereigntist discourses and concerns about the relative strength of one's country within the EU.

## Dependent variable
My dependent variable measures support for constitutional differentiated integration. The variable has a six-unit response scale and asks respondents whether they support a constitutionally differentiated EU (“Please tell us how far you agree or disagree with the following statement: Member states should be allowed to opt out of specific areas of European integration…”). The response categories are ordered from 1-5, with 1 being Strongly agree and 5 being strongly disagree. I reverse the coding, so that the scale goes from full disagreement to full agreement. By not asking about “constitutional differentiated integration”, the question becomes easier to understand. It also frames differentiation as a tool for strengthening autonomy and clarifies that countries themselves are able to opt out of policy areas if they so choose.

## Variables of key interest
To operationalize each country's exposure to EU differentiation I multiply, for each distinct differentiation, the number of articles from which a differentiation has been obtained with the number of years in which the differentiation was in force. This creates a variable for article-years. I then log-transform this variable to reduce the results’ sensitivity to outliers.

This dataset contains data on differentiations from the EU's treaties, rather than from the directives and regulations that constitute the EU’s main legislative output [@duttle_opting_2017]. This has one key benefit: The issue of differentiations from EU treaty provisions has mostly dealt with differentiation from broad and highly salient policy areas, such as monetary policy or the freedom of movement for workers. It is thus more likely that citizens will be aware of these differentiations, when compared to differentiation from ordinary legislation and technical regulation. By extensions differentiations from treaty law are thus more likely to be part of the backdrop against which citizens make up their mind about the future structure of the EU.

To analyse the regionally specific effects of differentiation, I use dummies for Scandinavian and Central and Eastern European citizenship. I code all Swedish and Danish citizens with a 1 on the Scandinavian dummy, and everyone from Hungary, Lithuania, Poland, and Romania with a 1 on the Central and Eastern European dummy. I then use these dummies to construct subsets of respondents that allow me to analyse the regionally specific effects assumed by *H1*, *H2* and *H3*.

## Control variables
I operationalize support for liberal economic policies, a relevant control variable [@leuffen_public_2020], through a question that asks whether respondents want to live in a Europe that “stresses economic integration, market competition and fiscal discipline”. This is contrasted to “a global Europe that acts as a leader on climate, human rights and global peace” or “a protective Europe that defends the European way of life and welfare against internal and external threats”. I code those who say that they prefer to live in a primarily market-oriented Europe as 1 and others 0 on this variable. While the question is multi-faceted, by asking about both fiscal discipline that may be described as austerity policies and more general market integration, I believe it still measures a general affinity for economic liberalism.

I include an equally weighted index of questions measuring the evaluations of the economy. These questions are found by a factor analysis, in contrast to a question about the respondent’s perceptions of the local job market, to measure the same underlying dimension (see table A3). Previous literature shows that positive evaluations of the national economy correlate with support for both uniform and differentiated integration (see for instance @gabel_public_1998; @harteveld_europe_2013; @hooghe_calculation_2005; @leuffen_public_2020 [12]). Egotropic evaluations, on the other hand, have been shown to have little relevance for explaining support for differentiated integration [@leuffen_public_2020].

My second set of variables measures support for the national government and underlying Eurosceptic attitudes. Both are shown to explain support for uniform and differentiated integration [@armingeon_loss_2014; @de_blok_blessing_2020; @harteveld_europe_2013). I operationalize Euroscepticism as a dummy that asks people whether they would vote to leave the EU in a hypothetical referendum or not. I code all those answering yes as 1 and everyone else as 0. Support for the national government is operationalized by a dummy asking respondents whether they believe their nation's government is doing a good job.

I operationalize identity through a widely used question that asks respondents to rank the inclusiveness of their identities. I construct a dummy where those answering that they identify solely with their nation-states are coded as 1 and everyone else as 0. While nested identities combining a national and European component are most common [@risse_euro_2003; @starke_concept_2021], previous literature finds that the distinction between exclusively national and territorially partially inclusive identities is more important for producing Euroscepticism than variations between different configurations of nested identities [@hooghe_calculation_2005].

I then use ideology and gender as socio-political control variables meant to isolate the effect of identity. These are established predictors of support for both general and specific integration (see for example @carrubba_decision_2004; @hobolt_public_2015; @hooghe_calculation_2005; @schoen_identity_2008). I operationalise ideology through a 7-unit scale, with 0 representing far-left and 6 far-right positions. Previous literature also finds evidence that a curvilinear correlation between ideology and support for the EU may exist, with left- and right wing non-centrist ideology both correlating with Euroscepticism [@toshkov_beyond_2022; @van_elsas_changing_2015]. I therefore add a squared indicator of left-right ideology.

## Limitations
A limitation of this study is that both surveys were fielded during the COVID-19 pandemic, which could impact support for differentiation. However, @cicchi_eu_2020 show that solidarity, as is common, seemed to still reside predominantly at the national level in the early phase of the crisis. This makes the results generalizable even beyond the period in which the survey was fielded. To account for the different baseline levels of support for constitutional differentiation, as well as how it is impacted by the COVID-19 pandemic, I use multilevel models with random country intercepts.

There are also large amounts of missingness on several of the variables (see table 1). Missing data may bias the results if they are correlated with individual-level characteristics like income, ideology, or gender. I test this bias through multiple imputation [@buuren_mice_2011] and show the results in the section Robustness checks.

A further limitation, general to all studies using observational survey data, is the possibility of omitted variable bias. In this study, omitted variable bias may stem from the absence of data on the respondents' knowledge of the EU and education [@karp_know_2003; @marquart_knowing_2019]. I use sensitivity analysis [@cinelli_making_2020] to tentatively quantify the size of this bias, and show the results of these analyses in the section Robustness checks.

```{r descriptive table 1, results='hide', message=FALSE, warning=FALSE}

load("EUI_merged.Rdata")
library(dplyr)
library(modelsummary)

data_descriptives <- EUI_final_merged %>%
  dplyr::select(article_log, opt_out_supp, market_europe, alternative_econ, 
                gov_trust, left_right,
                eurosceptic, left_right_sqr, gender, 
                excl_identity)

data_descriptives$excl_identity <- as.numeric(data_descriptives$excl_identity)

data_descriptives$eurosceptic <- as.numeric(data_descriptives$eurosceptic)

#Recode gender
data_descriptives$eurosceptic[data_descriptives$eurosceptic == 1] <- 0
data_descriptives$eurosceptic[data_descriptives$eurosceptic == 2] <- 1

#Recode exclusively national
data_descriptives$excl_identity[data_descriptives$excl_identity == 1] <- 0
data_descriptives$excl_identity[data_descriptives$excl_identity == 2] <- 1

data_descriptives$gender[data_descriptives$gender == 1] <- 0
data_descriptives$gender[data_descriptives$gender == 2] <- 1

#Create factors
data_descriptives <- data_descriptives %>%
  rename(`Support constitutional DI` = opt_out_supp) %>%
  rename(`Article-years (logged)` = article_log) %>%
  rename(`Liberal economic values` = market_europe) %>%
  rename(`Perception of economy` = alternative_econ) %>%
  rename(`Support national government` = gov_trust) %>%
  rename(`Left-right` = left_right) %>%
  rename(`Left-right (sqr.)` = left_right_sqr) %>%
  rename(`Eurosceptic` = eurosceptic) %>%
  rename(`Gender` = gender) %>%
  rename(`Exclusive identity` = excl_identity)
```

```{r descriptive table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
datasummary_skim(data_descriptives, title = "Descriptive statistics", 
                 histogram = FALSE)

```

# Results and discussion
I  first show the effect of the interaction between the log-transformed variable for article-years and Euroscepticism on support for constitutional differentiated integration. I then discuss the effect of differentiation in support for differentiation in both Scandinavia and Central and Eastern Europe.  
```{r model 1, results='hide', message=FALSE, warning=FALSE}
library(lme4)
opt_out_eurosceptic <- lmer(opt_out_supp ~ article_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                              eurosceptic + gender + excl_identity + article_log*eurosceptic + (1|country), weights = EUI_final_merged$weight, 
                            data = EUI_final_merged)

summary(opt_out_eurosceptic)

```

```{r table 2, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(stargazer)

stargazer(opt_out_eurosceptic,
          dep.var.caption	= c("Support for constitutional differentiated integration"), dep.var.labels.include = FALSE,
          covariate.labels = c("Constant","Article-years (logged)", "Support for free-market Europe", "Perception of economy", "Support for national government",
                               "Left-right","Left-right (sqr.)", "Eurosceptic", "Gender", "Exclusively national identity",
                               "Article-years (logged) X Eurosceptic"),
          keep.stat = c("n","aic"),
          digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001), intercept.top = TRUE, intercept.bottom = FALSE,
          type = "latex", notes.align = "l", notes.append = TRUE, no.space = TRUE,
          font.size = "small", column.sep.width = "-100pt", header = FALSE,
          title = "Support for constitutional differentiated integration with interaction between Euroscepticism and exposure to differentiation")
```

```{r figure 1, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.cap="Predicted probability of support for constitutional differentiation as a function of the interaction between article-years (logged) and Euroscepticism. 95% CIs.", fig.pos= "H"}
library(sjPlot)
library(ggplot2)

pred_plot_eurosceptic <- plot_model(opt_out_eurosceptic, type = "pred", terms = c("article_log", "eurosceptic"), 
                                    colors = "bw", legend.title = "Eurosceptic")

pred_plot_eurosceptic_final <-  pred_plot_eurosceptic + ggtitle("Support for constitutional differentiation") + labs(x = "Differentiation (article-years (logged))",
                                                                                                                     y = "Predicted value")
pred_plot_eurosceptic_final

```

Figure 1 indicates clear support for *H4*, that Eurosceptics will express stronger support for differentiation as a result of exposure to differentiation than what is the case for supporters of EU membership. This pattern repeats itself for most levels of differentiated integration, with insignificant differences between the two groups found only where citizens have been least exposed.

This may, as previously stated, be because Eurosceptic citizens in countries with greater exposure to differentiated integration are likelier to treat differentiated integration as a viable alternative to the European disintegration they might otherwise favour. In this sense, differentiation may be treated as a "second-best" alternative. However, one should not underestimate the fact that exposure to differentiated integration also increases support for a more differentiated EU among those who do not actively reject their country’s membership of the EU. It thus appears likely that future demands for differentiated integration will not be restricted only to Eurosceptic segments of the population

An alternative explanation for the correlation between exposure to differentiation and support for future differentiated integration could be that differentiation was primarily chosen where underlying Euroscepticism, and thus a demand for less integration, was already high. Figure A10 may offer some support for this, as Denmark and Sweden are among the countries in which the very high levels of differentiated integration can be explained by attempts to mitigate popular Euroscepticism [@leruth_operationalizing_2015; @spendzharova_banking_2016] However, countries as diverse as Lithuania, Hungary, Italy, and the Netherlands have also experienced differentiations from EU treaties. Differentiated integration is thus not restricted to what @bolstad_dynamics_2015 identifies as a Eurosceptic periphery. This strengthens the case that the relationship is not driven by strongly Eurosceptic outliers.

I next analyse how support for constitutional differentiation is shaped by the fact that differentiation is voluntarily chosen by testing whether support for constitutional differentiated integration is greater in Denmark and Sweden, two countries that have a long history of using differentiated integration to strengthen national autonomy [@adler-nissen_opting_2014; @leruth_operationalizing_2015].

```{r model 2, results='hide', message=FALSE, warning=FALSE}
library(lme4)
EUI_final_nordic <- EUI_final_merged %>%
  dplyr::filter(nordic_dummy == 1)

#Run analysis
opt_out_nordics <- lmer(opt_out_supp ~ article_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                          eurosceptic + gender + excl_identity + (1|country), weights = EUI_final_nordic$weight, 
                        data = EUI_final_nordic)

summary(opt_out_nordics)
```

```{r table 3, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(stargazer)

stargazer(opt_out_nordics,
          dep.var.caption	= c("Support for constitutional differentiated integration"), dep.var.labels.include = FALSE,
          covariate.labels = c("Constant","Article-years (logged)", "Support for free-market Europe", "Perception of economy", "Support for national government",
                               "Left-right","Left-right (sqr.)", "Eurosceptic", "Gender", "Exclusively national identity"),
          keep.stat = c("n","aic"),
          digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001), intercept.top = TRUE, intercept.bottom = FALSE,
          type = "latex", notes.align = "l", notes.append = TRUE, no.space = TRUE,
          font.size = "small", column.sep.width = "-50pt", header = FALSE,
          title = "Support for constitutional differentiated integration in Scandinavia")
```

Table 3 shows that the effect of exposure to differentiated integration in Scandinavia is large, but statistically insignificant. We thus find little conclusive support for the hypothesis that support for differentiated integration is greater where it is voluntarily chosen, rather than externally imposed.

```{r figure 2, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.cap="Predicted probability of support for constitutional differentiation in Sweden and Denmark. 95% CIs.", fig.pos= "H"}
library(sjPlot)
library(ggplot2)

pred_plot_nordics <- plot_model(opt_out_nordics, type = "pred", terms = "article_log")

pred_plot_nordics_final <-  pred_plot_nordics + ggtitle("Support for constitutional differentiation (Scandinavia)") + labs(x = "Differentiation (article-years (logged)",
                                                                                                                           y = "Predicted value")
pred_plot_nordics_final

```

Figure 2 corroborates this picture: We find only small increases in support for constitutional differentiated integration among Scandinavian citizens as exposure to differentiation increases. The very large confidence intervals furthermore underline the uncertainty of these predicted values. The appendix shows that this pattern is found among both Eurosceptics and supporters of EU membership in the Scandinavian countries.

One last question is whether exposure to largely involuntary differentiated integration lowers support for differentiated integration in the longer run. To test this, I show the effect of exposure to differentiated integration in Central and Eastern Europe.

```{r r model 3, results='hide', message=FALSE, warning=FALSE}
library(lme4)
EUI_final_CEE <- EUI_final_merged %>%
  dplyr::filter(CEE_dummy == 1)

#Run opt-out for CEE
opt_out_CEE <- lmer(opt_out_supp ~ article_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                      eurosceptic + gender + excl_identity + (1|country), weights = EUI_final_CEE$weight, 
                    data = EUI_final_CEE)

summary(opt_out_CEE)
```

```{r table 4, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(stargazer)

stargazer(opt_out_CEE,
          dep.var.caption	= c("Support for constitutional differentiated integration"), dep.var.labels.include = FALSE,
          covariate.labels = c("Constant","Article-years (logged)", "Support for free-market Europe", "Perception of economy", "Support for national government",
                               "Left-right","Left-right (sqr.)", "Eurosceptic", "Gender", "Exclusively national identity"),
          keep.stat = c("n","aic"),
          digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001), intercept.top = TRUE, intercept.bottom = FALSE,
          type = "latex", notes.align = "l", notes.append = TRUE, no.space = TRUE,
          font.size = "small", column.sep.width = "-50pt", header = FALSE,
          title = "Support for constitutional differentiated integration in Central and Eastern Europe")
```

The table shows that there is a small effect of exposure to differentiated integration in Central and Eastern Europe, but that the effect is statistically insignificant. This suggests that the effects of exposure to differentiation, whether positive or negative, are limited.

```{r figure 3, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.cap="Predicted probability of support for constitutional differentiation in Central and Eastern Europe. 95% CIs.", fig.pos= "H"}
library(sjPlot)
library(ggplot2)

pred_plot_CEE <- plot_model(opt_out_CEE, type = "pred", terms = "article_log")

pred_plot_CEE_final <-  pred_plot_CEE + ggtitle("Support for constitutional differentiation (Central and Eastern Europe)") + labs(x = "Differentiation (article-years (logged)",
                                                                                                                                  y = "Predicted value")
pred_plot_CEE_final

```

The predicted support shown by figure 3 largely corroborates this result. This weakens support for *H2*, that support will decrease as a function of exposure to differentiation, while offering tentative support for *H1*, stating the opposite. However, the lack of statistical significance means that the results should not be treated as definitive proof of a positive correlation. The true direction of the feedback loop between exposure to and support for differentiation is thus unclear.

The interaction between Euroscepticism and exposure to differentiated integration in Central and Eastern Europe (shown by figure A1) nuances the broader story of this paper: Eurosceptic individuals in Central and Eastern Europe are less likely to support differentiated integration if they have been exposed to greater amounts of it in the past. The first explanation for this may of course be that these citizens' primary exposure to differentiated integration was as something involuntary and externally imposed. However, another reason may be that their greater exposure to differentiated integration could have made them more responsive to elite cueing of differentiated integration as a potential threat to the nation's influence within the EU. Such frames have been especially prevalent in Polish debates about the future shape of the EU [@cianciara_differentiated_2014]. The fact that regional circumstances and cues potentially play a greater role than historical trajectories of integration is also evident in how increasing opposition to European differentiation among Eurosceptics is also found in those countries with the highest levels of exposure to differentiated integration in Southern Europe (see figure A11).

The fact that the surveys were fielded at the height of the COVID-19 pandemic means that we may be seeing lower levels of support for differentiated integration than we would at other times: As @genschel_postfunctionalism_2021 argue, the COVID-19 pandemic “debordered” solidarity, making the prospect of mutual aid across member state borders more attractive to EU citizens. In times where EU solidarity is positively framed, support for differentiation integration may be lower than it otherwise would have been. More research is thus needed to establish whether the patterns established by this paper are consistent also in the future.

# Robustness checks

I use an approach outlined in @cinelli_making_2020 test the sensitivity of all models to omitted variable bias (see figures A3-8). I thus test the variance that a potential confounder would need to account for to nullify the effect of theoretical interest, exposure to differentiated integration. This shows that all models, to varying degrees, are sensitive to such bias. For instance, the test of the model showing the effect of article-years on support for constitutional differentiated integration in Scandinavia (figure A5) shows that a variable explaining 50% of the remaining variance of support for constitutional differentiation would need to explain less than 10% of the remaining variance in real exposure to differentiation for the effect of exposure to reach zero. This is in line with what is found for the other models. One important omitted variable is potentially knowledge of the EU [@leuffen_public_2020]. While knowledge of the EU is unlikely to be strongly connected with actual levels of differentiation, the tests show that omitted variable bias remains a concern for this study

I next use chained equation multiple imputation [@buuren_mice_2011; @fjelstul_explaining_2022] to test how sensitive the results are to bias induced by missing data. Multiple imputation uses each unit's observed data to derive probable values for all covariates with missing values. Each model is imputed 25 times, before the imputations are pooled to provide multiply imputed parameter estimates and standard errors. This better accounts for the inherent uncertainty associated with data imputation. As appendix tables A4-6 show, the multiply imputed models are similar to those from tables 2-4. One important exception is the effect of exposure to differentiated integration in Central and Eastern Europe: While there is a weak, and insignificant, positive effect of exposure in the main model, the multiply imputed model shows an insignificant, weakly negative, effect. Despite this, the tests suggest that the effects established in this paper are not primarily driven by systematic data missingness.

Lastly, I test the robustness of the results to an alternative specification of Euroscepticism: satisfaction with EU democracy. Dissatisfaction with the state of EU democracy can be seen as a less stringent form of Euroscepticism than wanting one's country to leave the EU. Figure A9 shows the same trend as the main results: Those with the lowest levels of satisfaction with EU democracy, and who arguably have the strongest Eurosceptic sentiments, express the greatest support for constitutional differentiated integration. Their support for differentiation also increases more strongly as a function of exposure to previous differentiation than what is the case for less Eurosceptic groups. Both facts strengthen my original conclusions.

# Concluding discussion
This paper establishes, through both recent survey data and data outlining all differentiations from the EU treaties, that there is a positive feedback loop between exposure to differentiated integration and support for more future EU differentiation. This may be because, as citizens become more and more accustomed to such differentiated membership of the EU, it may be perceived as a more favourable way of combining the benefits of national membership of the EU with extended national autonomy. I also show that this particularly applies to those sceptical of their countries' EU membership, who may see differentiation as a "second-best" alternative to European disintegration

There is one exception to this rule: Eurosceptic citizens of the Central and Eastern European countries are less favourably inclined towards differentiated integration if they live in countries with greater exposure to differentiated integration. I point to two competing explanations for why this may be the case: First, the largely involuntary and externally imposed differentiation that Central and Eastern European citizens have primarily been exposed to could shape Eurosceptic attitudes towards differentiation in this region. The fact that these citizens were excluded from for instance the freedom of movement for workers, a highly salient area that would be beneficial to individual citizens [schimmelfennig_eu_2014], could be one source of the negative attitudes found among Eurosceptics. Secondly, these individuals may be more concerned with maintaining national power than those identifying with support for a membership of the EU that necessarily implies ceding some sovereignty to a European political level. Eurosceptics may thus be more attentive to elite's discursive framing of differentiated integration as a path towards a second-rate membership for the Central and Eastern European member states and adjust their attitudes accordingly.

My results open two avenues of future research: It is first important to establish whether cues regarding European integration in general shape views of differentiated integration, and what kind of cues are most important for developing preferences for such differentiation. It may for instance be the case that those who are exposed to Eurosceptic news content may respond not with explicit Euroscepticism, but by developing preferences for a more differentiated Europe. The true impact of cues on such preferences, how and when Eurosceptic cues may produce for differentiation rather than disintegration, and what the circumstances of implementation means for public support are important questions that should be studied using both large-*N* approaches and experimental methods.

The second avenue for future research deals with the question of whether the public is equally likely to prefer differentiation in all policy areas. Existing research [de_blok_blessing_2020; @leuffen_public_2020] has largely focused on individual support for polity-level differentiation of the EU. It is both possible and probable that even those favouring an “ever closer union” may have diverging preferences for integration at the level of individual policies. Testing this proposition may allow us to develop models for understanding when people will express ambivalent, rather than overtly critical, attitudes towards European integration [@de_vries_variable_2013; @goldberg_one_2021].

My results also have implications for the literatures on differentiated integration, the post-crisis EU and the drivers of support for transnational cooperation: They suggest that the increasing use of differentiation may prove problematic for an EU that is still normatively committed to the idea of “ever closer Union”. The clear connection between exposure to differentiation and preference for it as a way organizing a European political community may also offer insights into the possible consequences of allowing for greater legal flexibility for non-European transnational cooperations seeking greater political integration in the future.

# References
<div id="refs"></div>

\newpage
# Appendix

```{r model A1, results='hide', message=FALSE, warning=FALSE}
CEE_eurosceptic <- lmer(opt_out_supp ~ article_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                      eurosceptic + gender + excl_identity + article_log*eurosceptic + (1|country), weights = EUI_final_CEE$weight, 
                    data = EUI_final_CEE)

summary(CEE_eurosceptic)
```

\setcounter{figure}{0}
\renewcommand*{\thefigure}{A\arabic{figure}}
```{r figure A1, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.cap="Support for differentiated integration in Central and Eastern Europe as an interaction between Euroscepticism and exposure to differentiated integration. 95% CIs.", fig.pos= "H"}

pred_plot_CEE_eurosceptic <- plot_model(CEE_eurosceptic, type = "pred", terms = c("article_log", "eurosceptic"), 
                                        legend.title = "Eurosceptic", colors = "bw")

pred_plot_CEE_eurosceptic_final <-  pred_plot_CEE_eurosceptic + ggtitle("Support for constitutional differentiation (Central and Eastern Europe)") + labs(x = "Differentiation (article-years (logged)",
                                                                                                                                                          y = "Predicted value")
pred_plot_CEE_eurosceptic_final
```

\setcounter{table}{0}
\renewcommand*{\thetable}{A\arabic{table}}
```{r table A1, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
stargazer(CEE_eurosceptic,
          dep.var.caption	= c("Support for constitutional differentiated integration (Central and Eastern Europe)"), dep.var.labels.include = FALSE,
          covariate.labels = c("Constant","Article-years (logged)", "Support for free-market Europe", "Perception of economy", "Support for national government",
                               "Left-right","Left-right (sqr.)", "Eurosceptic", "Gender", "Exclusively national identity", 
                               "Article-years (logged) X Euroscepticism"),
          keep.stat = c("n","aic"),
          digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001), intercept.top = TRUE, intercept.bottom = FALSE,
          type = "latex", title = "Table showing interaction between exposure to differentiated integration and Euroscepticism in Central and Eastern Europe", header = FALSE, column.sep.width = "-150pt")
```

```{r model A3, results='hide', message=FALSE, warning=FALSE}
opt_out_nordics_eurosceptic <- lmer(opt_out_supp ~ article_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                                      eurosceptic + gender + excl_identity + article_log*eurosceptic + (1|country), weights = EUI_final_nordic$weight, 
                                    data = EUI_final_nordic)

summary(opt_out_nordics_eurosceptic)
```

```{r figure A2, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.cap="Support for differentiated integration in Scandinavia as an interaction between Euroscepticism and exposure to differentiated integration. 95% CIs.", fig.pos= "H"}
pred_plot_nordics_eurosceptic <- plot_model(opt_out_nordics_eurosceptic, type = "pred", terms = c("article_log", "eurosceptic"),
                                            legend.title = "Eurosceptic", colors = "bw")

pred_plot_nordics_eurosceptic_final <-  pred_plot_nordics_eurosceptic + ggtitle("Support for constitutional differentiation (Scandinavia)") + labs(x = "Differentiation (article-years (logged)",
                                                                                                                                                   y = "Predicted value")

pred_plot_nordics_eurosceptic_final

```

\renewcommand*{\thetable}{A\arabic{table}}
```{r table A2, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
stargazer(opt_out_nordics_eurosceptic,
          dep.var.caption	= c("Support for constitutional differentiated integration (Scandinavia)"), dep.var.labels.include = FALSE,
          covariate.labels = c("Constant","Article-years (logged)", "Support for free-market Europe", "Perception of economy", "Support for national government",
                               "Left-right","Left-right (sqr.)", "Eurosceptic", "Gender", "Exclusively national identity", 
                               "Article-years (logged) X Eurosceptic"),
          keep.stat = c("n","aic"),
          digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001), intercept.top = TRUE, intercept.bottom = FALSE,
          type = "latex", header = FALSE, title = "Table showing interaction between exposure to differentiated integration and Euroscepticism in Scandinavia", column.sep.width = "-100pt")
```

```{r model table A3, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(sensemakr)

EUI_final_merged$diffsupp = EUI_final_merged$opt_out_supp
EUI_final_merged$diffsum = EUI_final_merged$article_log

optout_nointeract <- lm(diffsupp ~ diffsum + alternative_econ + gov_trust + left_right + left_right_sqr +
eurosceptic + gender + excl_identity + factor(country), weights = EUI_final_merged$weight, 
data = EUI_final_merged)

summary(optout_nointeract)
```

```{r table A3, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}
#Sensitivity analyses
library(sensemakr)

sensitivity_optout <- sensemakr(optout_nointeract, 
                                           treatment = "diffsum")

summary(sensitivity_optout)
```

\renewcommand*{\thefigure}{A\arabic{figure}}
```{r figure A3, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.cap="Omitted variable bias in un-interacted model. Analysis of size of confounding effect and its impact on effect size."}
library(sensemakr)

plot(sensitivity_optout, type = "extreme")
```

```{r model table A4, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(sensemakr)
optout_interact <- lm(diffsupp ~ diffsum + alternative_econ + gov_trust + left_right + left_right_sqr +
                        eurosceptic + gender + excl_identity + diffsum*eurosceptic + factor(country), weights = EUI_final_merged$weight, 
                      data = EUI_final_merged)

summary(optout_interact)

sensitivity_interact <- sensemakr(optout_interact, 
                                  treatment = "diffsum:eurosceptic1")

summary(sensitivity_interact)
```

```{r figure A4, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.cap="Omitted variable bias in interacted model. Analysis of size of confounding effect and its impact on effect size"}
plot(sensitivity_interact, type = "extreme")
```

```{r model figure A5, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}

optout_nordic_fixeff <- lm(opt_out_supp ~ article_log + alternative_econ + gov_trust + left_right + left_right_sqr +
                             eurosceptic + gender + excl_identity + factor(country), weights = EUI_final_nordic$weight, 
                           data = EUI_final_nordic)

summary(optout_nordic_fixeff)

#Run sensitivity analysis
sensitivity_optout_nordic <- sensemakr(optout_nordic_fixeff, 
                                       treatment = "article_log")

summary(sensitivity_optout_nordic)
```

```{r figure A5, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.cap="Omitted variable bias in un-interacted model of Scandinavian data. Analysis of size of confounding effect and its impact on effect size"}

plot(sensitivity_optout_nordic, type = "extreme")
```

```{r model figure A6, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}

optout_nordic_fixeff_interact <- lm(two_speed_supp ~ article_log + alternative_econ + gov_trust + left_right + left_right_sqr +
                                      eurosceptic + gender + excl_identity + article_log*eurosceptic + factor(country), weights = EUI_final_nordic$weight, 
                                    data = EUI_final_nordic)

summary(optout_nordic_fixeff_interact)

sensitivity_interact_nordic <- sensemakr(optout_nordic_fixeff_interact, 
                                         treatment = "article_log:eurosceptic1")

summary(sensitivity_interact_nordic)
```

```{r figure A6, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.cap="Omitted variable bias in interacted model of Scandinavian data. Analysis of size of confounding effect and its impact on effect size"}
plot(sensitivity_interact_nordic, type = "extreme")
```

```{r model figure A7, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
optout_CEE_fixeff <- lm(opt_out_supp ~ article_log + alternative_econ + gov_trust + left_right + left_right_sqr +
                          eurosceptic + gender + excl_identity + factor(country), weights = EUI_final_CEE$weight, 
                        data = EUI_final_CEE)

summary(optout_CEE_fixeff)

#Run sensitivity analysis
sensitivity_optout_CEE <- sensemakr(optout_CEE_fixeff, 
                                    treatment = "article_log")

summary(sensitivity_optout_CEE)
```

```{r figure A7, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.cap="Omitted variable bias in un-interacted model of Central and Eastern European data. Analysis of size of confounding effect and its impact on effect size"}
plot(sensitivity_optout_CEE, type = "extreme")
```

```{r model figure A8, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}

optout_CEE_fixeff_interact <- lm(two_speed_supp ~ article_log + alternative_econ + gov_trust + left_right + left_right_sqr +
                                   eurosceptic + gender + excl_identity + article_log*eurosceptic + factor(country), weights = EUI_final_CEE$weight, 
                                 data = EUI_final_CEE)

summary(optout_CEE_fixeff_interact)

#Run sensitivity analysis
sensitivity_interact_CEE <- sensemakr(optout_CEE_fixeff_interact, 
                                      treatment = "article_log:eurosceptic1")

summary(sensitivity_interact_CEE)
```

```{r figure A8, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.cap="Omitted variable bias in interacted model of Central and Eastern data. Analysis of size of confounding effect and its impact on effect size"}
plot(sensitivity_interact_CEE, type = "extreme")
```

```{r fa, echo=FALSE, warning=FALSE, message=FALSE, results='hide', include=FALSE}
library(psychTools)
library(psych)
economy_index_EUI <- EUI_final_merged %>%
  dplyr::select(personal_financial, economic_situation_nation, 
                employment_area) %>%
    rename(`Personal economic situation` = personal_financial) %>%
  rename(`National economic situation` = economic_situation_nation) %>%
  rename(`Employment opportunities of local areas` = employment_area) %>%
  na.omit()

economy_index_EUI <- as.data.frame(economy_index_EUI)

econ_fa <- psych::fa(economy_index_EUI, nfactors = 2)
```

```{r table_fa, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
cat(fa2latex(econ_fa, caption="Factor analysis of perception of economy",
         heading= "Factor loading cutoff: > 0.5", silent = TRUE))

```

```{r mi_complete, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(sandwich)
library(mice)
library(miceadds)
library(dplyr)
library(broom)
library(broom.mixed)
library(kableExtra)

load("multiple_impute.Rdata")

mice_optout <- with(multiple_impute_EUI, lmer(opt_out_supp ~ article_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                                                eurosceptic + gender + excl_identity + article_log*eurosceptic + (1|country), weights = multiple_impute_EUI$weight))

summary(pool(mice_optout))

result_data <- data.frame(summary(pool(mice_optout)), 
                             stringsAsFactors = FALSE)

#Rename
result_data <- result_data %>%
  rename(`Variable` = term) %>%
  rename(`Estimate` = estimate) %>%
  rename(`Std. error` = std.error) %>%
  rename(`T-statistic` = statistic) %>%
  rename(`DFs` = df) %>%
  rename(`p-value` = p.value)

result_data$Variable <- as.character(result_data$Variable)

result_data$Variable[result_data$Variable == "(Intercept)"] <- "Constant"
result_data$Variable[result_data$Variable == "article_log"] <- "Article-years (logged)"
result_data$Variable[result_data$Variable == "market_europe"] <- "Liberal economic values"
result_data$Variable[result_data$Variable == "alternative_econ"] <- "Perception of economic situation"
result_data$Variable[result_data$Variable == "gov_trust"] <- "Support national government"
result_data$Variable[result_data$Variable == "left_right"] <- "Left-right"
result_data$Variable[result_data$Variable == "left_right_sqr"] <- "Left-right (sqr.)"
result_data$Variable[result_data$Variable == "eurosceptic1"] <- "Eurosceptic"
result_data$Variable[result_data$Variable == "gender"] <- "Gender"
result_data$Variable[result_data$Variable == "excl_identity1"] <- "Exclusively national identity"
result_data$Variable[result_data$Variable == "article_log:eurosceptic1"] <- "Article-years (logged) X Eurosceptic"

```

```{r texreg, results='asis', warning=FALSE, echo=FALSE, warning=FALSE}
knitr::kable(result_data, digits = 3, format = "latex",
             caption = "Multiply imputed multilevel model with interaction")
```

```{r mi_complete_nordic, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(sandwich)
library(mice)
library(miceadds)
library(dplyr)
library(broom)
library(broom.mixed)

load("multiple_impute_nordic.Rdata")
mice_optout_nordic <- with(multiple_impute_nordic, lmer(opt_out_supp ~ article_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                                                eurosceptic + gender + excl_identity + (1|country), weights = EUI_final_nordic$weight))

summary(pool(mice_optout_nordic))

result_data_nordic <- data.frame(summary(pool(mice_optout_nordic)), 
                             stringsAsFactors = FALSE)

#Rename
result_data_nordic <- result_data_nordic %>%
  rename(`Variable` = term) %>%
  rename(`Estimate` = estimate) %>%
  rename(`Std. error` = std.error) %>%
  rename(`T-statistic` = statistic) %>%
  rename(`DFs` = df) %>%
  rename(`p-value` = p.value)

result_data_nordic$Variable <- as.character(result_data_nordic$Variable)

result_data_nordic$Variable[result_data_nordic$Variable == "(Intercept)"] <- "Constant"
result_data_nordic$Variable[result_data_nordic$Variable == "article_log"] <- "Article-years (logged)"
result_data_nordic$Variable[result_data_nordic$Variable == "market_europe"] <- "Liberal economic values"
result_data_nordic$Variable[result_data_nordic$Variable == "alternative_econ"] <- "Perception of economic situation"
result_data_nordic$Variable[result_data_nordic$Variable == "gov_trust"] <- "Support national government"
result_data_nordic$Variable[result_data_nordic$Variable == "left_right"] <- "Left-right"
result_data_nordic$Variable[result_data_nordic$Variable == "left_right_sqr"] <- "Left-right (sqr.)"
result_data_nordic$Variable[result_data_nordic$Variable == "eurosceptic1"] <- "Eurosceptic"
result_data_nordic$Variable[result_data_nordic$Variable == "gender"] <- "Gender"
result_data_nordic$Variable[result_data_nordic$Variable == "excl_identity1"] <- "Exclusively national identity"
result_data_nordic$Variable[result_data_nordic$Variable == "article_log:eurosceptic1"] <- "Article-years (logged) X Eurosceptic"

```

```{r kable_complete_nordic, results='asis', warning=FALSE, echo=FALSE, warning=FALSE}
library(kableExtra)
knitr::kable(result_data_nordic, digits = 3, format = "latex",
             caption = "Multiply imputed multilevel model (Scandinavia)")
```

```{r mi_complete_CEE, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(sandwich)
library(mice)
library(miceadds)
library(dplyr)
library(broom)
library(broom.mixed)

load("multiple_impute_CEE.Rdata")

mice_optout_CEE <- with(multiple_impute_CEE, lmer(opt_out_supp ~ article_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                                                eurosceptic + gender + excl_identity + (1|country), weights = EUI_final_CEE$weight))

summary(pool(mice_optout_CEE))

result_data_CEE <- data.frame(summary(pool(mice_optout_CEE)), 
                             stringsAsFactors = FALSE)

#Rename
result_data_CEE <- result_data_CEE %>%
  rename(`Variable` = term) %>%
  rename(`Estimate` = estimate) %>%
  rename(`Std. error` = std.error) %>%
  rename(`T-statistic` = statistic) %>%
  rename(`DFs` = df) %>%
  rename(`p-value` = p.value)

result_data_CEE$Variable <- as.character(result_data_CEE$Variable)

result_data_CEE$Variable[result_data_CEE$Variable == "(Intercept)"] <- "Constant"
result_data_CEE$Variable[result_data_CEE$Variable == "article_log"] <- "Article-years (logged)"
result_data_CEE$Variable[result_data_CEE$Variable == "market_europe"] <- "Liberal economic values"
result_data_CEE$Variable[result_data_CEE$Variable == "alternative_econ"] <- "Perception of economic situation"
result_data_CEE$Variable[result_data_CEE$Variable == "gov_trust"] <- "Support national government"
result_data_CEE$Variable[result_data_CEE$Variable == "left_right"] <- "Left-right"
result_data_CEE$Variable[result_data_CEE$Variable == "left_right_sqr"] <- "Left-right (sqr.)"
result_data_CEE$Variable[result_data_CEE$Variable == "eurosceptic1"] <- "Eurosceptic"
result_data_CEE$Variable[result_data_CEE$Variable == "gender"] <- "Gender"
result_data_CEE$Variable[result_data_CEE$Variable == "excl_identity1"] <- "Exclusively national identity"
result_data_CEE$Variable[result_data_CEE$Variable == "article_log:eurosceptic1"] <- "Article-years (logged) X Eurosceptic"

```

```{r kable_complete_CEE, results='asis', warning=FALSE, echo=FALSE, warning=FALSE}
library(kableExtra)
knitr::kable(result_data_CEE, digits = 3, format = "latex",
             caption = "Multiply imputed multilevel model (Central and Eastern Europe)")
```

```{r alt_eurosceptic, results='hide', warning=FALSE, echo=FALSE, message=FALSE}
opt_out_main_interact <-lmer(opt_out_supp ~ article_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                               eu_democ_satisf + gender + excl_identity + article_log*eu_democ_satisf + 
                               (1|country), weights = EUI_final_merged$weight, 
                             data = EUI_final_merged)

summary(opt_out_main_interact)
```

```{r alt_eurosceptic stargazer, results='asis', warning=FALSE,echo=FALSE, message=FALSE}
stargazer(opt_out_main_interact,
          dep.var.caption	= c("Support for constitutional differentiated integration"), dep.var.labels.include = FALSE,
          covariate.labels = c("Constant","Article-years (logged)", "Support for free-market Europe", "Perception of economy", "Support for national government",
                               "Left-right","Left-right (sqr.)", "EU democracy satisfaction", "Gender", "Exclusively national identity", 
                               "Article-years (logged) X EU democracy satisfaction"),
          keep.stat = c("n","aic"),
          digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001), intercept.top = TRUE, intercept.bottom = FALSE, column.sep.width = "-100pt",
          type = "latex", header = FALSE, title = "Main model with alternative specification of Euroscepticism")
```

```{r alt_euro fig, message=FALSE, warning=FALSE, echo=FALSE, results='asis', fig.cap= "Predicted values of support for constitutional differentiated integration as a function of an interaction between exposure to differentiated integration and satisfaction with EU democracy. Slopes for minimum, median and maximum values of satisfaction with EU democracy are shown. 95% CIs.", fig.pos= "H"}
pred_plot_interact_main <- plot_model(opt_out_main_interact, type = "pred", terms = c("article_log", "eu_democ_satisf"), 
                                      colors = "bw", legend.title = "EU democracy satisfaction")

pred_plot_interact_main_final <-  pred_plot_interact_main + ggtitle("Support for constitutional differentiation", 
                                                                    subtitle = "Interaction between EU democracy satisfaction and differentiation") + labs(x = "Differentiation (article-years (logged)",
                                                                                                                                                           y = "Predicted value")
pred_plot_interact_main_final
```

```{r diff_by_country, message=FALSE, warning=FALSE, echo=FALSE, results='asis', fig.cap= "Number of differentiations from EU primary law 1952-2020 (log-transformed article-years) by EU member state.", fig.pos= "H"}
library(ggplot2)
bar_chart_articleyear <- ggplot(EUI_final_merged, aes(x = factor(country), y = article_log)) +
  geom_col(stat = "identity", position = "dodge")

bar_chart_articleyear + labs(x = "Country", y = "Article-years (log-transformed)") +
  scale_x_discrete(breaks = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14),
                   labels = c("UK", "DK", "FI", "FR", "DE", "SE", "GR", "HU", "IT",
                              "LT", "NL", "PL", "RO", "ES")) +
  ggtitle(label = "Differentiated integration in the EU", 
          subtitle = "Primary law differentiations (article-years log-transformed) by country")
```

```{r southern_fig_model, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}

EUI_final_southern <- EUI_final_merged %>%
  dplyr::filter(south_dummy == 1)

##Create regression
opt_out_southern <- lmer(opt_out_supp ~ article_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                          eurosceptic + gender + excl_identity + (1|country), weights = EUI_final_southern$weight, 
                        data = EUI_final_southern)

summary(opt_out_southern)
```

```{r southern_fig, fig.cap="Support for differentiated integration in Southern Europe as an interaction between Euroscepticism and exposure to differentiated integration. 95% CIs.", fig.pos= "H", results='asis', warning=FALSE, message=FALSE, warning=FALSE, echo=FALSE}

pred_plot_southern <- plot_model(opt_out_southern, type = "pred", terms = "article_log")

pred_plot_southern_final <-  pred_plot_southern + ggtitle("Support for constitutional differentiation (Southern Europe)") + labs(x = "Differentiation (article-years (logged)",
                                                                                                                           y = "Predicted value")
pred_plot_southern_final
```

```{r southern_stargazer, results='asis', warning=FALSE, message=FALSE, echo=FALSE}

stargazer(opt_out_southern,
          dep.var.caption	= c("Support for constitutional differentiated integration in Southern Europe"), dep.var.labels.include = FALSE,
          covariate.labels = c("Constant","Article-years (logged)", "Support for free-market Europe", "Perception of economy", "Support for national government",
                               "Left-right","Left-right (sqr.)", "Eurosceptic", "Gender", "Exclusively national identity"),
          keep.stat = c("n","aic"),
          digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001), intercept.top = TRUE, intercept.bottom = FALSE,
          type = "latex", header = FALSE, column.sep.width = "-100pt", 
          title = "Table showing interaction between exposure to differentiated integration and support for differentiated integration in Southern Europe")
```

### 2022-07-02 Preprint-Differentiation.Rmd
Uploaded the newest version of .Rmd
This includes the newest version of the .Rmd file necessary for replication.

@@
 date: "`r format(Sys.Date(), '%d %B %Y')`"
-output: pdf_document
+output: 
+  pdf_document:
+    keep_md: true
 abstract: "Questions about the public legitimacy of the EU's current flexible integration are becoming increasingly important in the public opinion literature.
@@
 
-My findings are relevant to the discussion about the EU's future that Brexit gave rise to [@european_commission_white_2017]: As differentiated integration has become a more important tool of EU governance, especially given the EU’s increasing integration of for instance monetary and fiscal policy [@schimmelfennig_ever_2020; @genschel_differentiated_2013], greater exposure to it among member state citizens may make it more difficult for the EU’s institutions to consolidate support for the project of further integration.
+My findings are relevant to the discussion about the EU's future spawned by Brexit [@european_commission_white_2017]: As differentiated integration has become a more important tool of EU governance, especially given the EU’s increasing integration of for instance monetary and fiscal policy [@schimmelfennig_ever_2020; @genschel_differentiated_2013], greater exposure to it among member state citizens may make it more difficult for the EU’s institutions to consolidate support for the project of further integration.
 The article also makes an important contribution to the emerging literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009], by showing that implementation of differentiations from international law can potentially increase public demand for future divergences from the same legal order.
@@
 This literature, in other words, mostly focuses on predictors found at the level of individuals.
-However, popular attitudes towards both the EU and international cooperation more broadly are often shaped by elite cues and context [@dellmuth_elite_2021; @edwards_public_2009; @harteveld_europe_2013; @telle_elite-mass_2022-1].
-One under-explored source of such support is the large variation in differentiated integration found across the EU [@schimmelfennig_instrumental_2014]: As either externally imposed or voluntarily chosen differentiation may be either in or out of step with popular attitudes towards how the EU should look, past differentiation may face a popular response in the form of lesser or greater support for differentiation [@busemeyer_beyond_2021; @wlezien_public_1995].
-While previous literature has investigated how this feedback mechanism plays out in the case of EU policies [@bolstad_dynamics_2015; @de_bruycker_democratically_2020], this paper will instead investigate how differentiations from the EU's primary legal framework shapes demand for a differently configured polity.
+However, popular attitudes towards both the EU and international cooperation in general are often shaped by elite cueing and national contexts [@dellmuth_elite_2021; @edwards_public_2009; @harteveld_europe_2013; @telle_elite-mass_2022-1].
+One under-explored source of such support is the large variation in differentiated integration found across the EU [@schimmelfennig_instrumental_2014].
+Because differentiation may be either externally imposed or voluntarily chosen and more or less comprehensive than what citizens prefer, it may be out of step with popular preferences regarding the shape of the future EU.
+This may lead to a thermostatic response, in the form of either greater or lesser support for future differentiated integration [@busemeyer_beyond_2021; @wlezien_public_1995].
+While previous literature has investigated how this feedback mechanism plays out at the level of policies [@bolstad_dynamics_2015; @de_bruycker_democratically_2020], this paper will instead investigate how differentiations from the EU's primary legal framework shapes demand for a differently configured polity.
 
-The effect of exposure to differentiation may also be compounded by who originally initiated differentiations, and the circumstances under which it happened.
-An interesting test of this assumption is support for differentiation in Central and Eastern Europe.
+The effect of exposure to differentiation may also be compounded by who originally initiated it.
+An interesting test of this is to compare support for differentiation in Central and Eastern Europe to that found in the rest of the EU.
 Here citizens were exempt from the right to move across EU borders to seek work for several years after their countries acceded to the EU.
@@
 
-However, another causal mechanism could explain lower levels of support for differentiated integration in Central and Eastern Europe without relying on a path dependency argument: Central and Eastern European politicians have framed differentiated integration as a path towards an EU consisting of a powerful core and a less powerful periphery.
+However, another causal mechanism could explain lower levels of support for differentiated integration in Central and Eastern Europe: Central and Eastern European politicians have framed differentiated integration as a path towards an EU consisting of a powerful core and a less powerful periphery.
 This frame was prominent during the sovereign debt crisis [@cianciara_differentiated_2014 178-179].
-For example, Polish politicians opposed further integration of the Eurozone, a clear example of differentiated integration, because they feared that the Eastern member states would become less influential in economic policy-making due to being non-Eurozone countries.
+For example, Polish politicians opposed further integration of the Eurozone, a prominent example of differentiated integration, because they feared that the Central and Eastern member states would become less influential in economic policy-making due to being non-Eurozone countries.
 Such cues have been known to shape popular attitudes towards integration, particularly where citizens have little knowledge of the relevant policies.
@@
 
-Evidence of this is found in how the renewed Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy on the part of those identifying as Eurosceptic [@schraff_does_2020].
+This is evident in how renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy on the part of those identifying as Eurosceptic [@schraff_does_2020].
 This suggests that citizens may respond positively to differentiation, and could indicate that they would be favourably inclined towards further differentiation.
@@
 I combine these data with the expanded EUDIFF1 dataset [@schimmelfennig_eudiff1_2022].
-This dataset features all distinct differentiations from the EU treaties, showing for each country the number of articles in which differentiation was given, the years they were in force and the policy areas covered.
+This dataset features all distinct differentiations from the EU treaties, showing for each country the number of articles to which differentiation was applied, the years they were in force and the policy areas covered.
 The surveys, on the other hand, ask a range of questions related to European integration.

### 2022-08-30 Manuscript_3_draft1.Rmd
Committed new version of preprint.

---
title: "Does differentiation beget differentiation?"
subtitle: Examining the link between implementation of and support for differentiation
thanks: "This is a very early draft. All comments are welcome at martin.moland@arena.uio.no. This version was finalized at `r format(Sys.time(), '%B %d, %Y')`."
author:
- name: Martin Moland
  affiliation: ARENA, Centre for European Studies
# bibliography: "`r paste0(Sys.getenv('HOME'),'/Dropbox/master.bib')`" # This would be more general
bibliography: inst/bib_paper_3.bib
csl: inst/european-journal-of-political-research.csl
abstract: "Questions about the public legitimacy of the EU’s increasingly flexible integration are attracting increasing scholarly interest. However, we still know little about whether exposure to past differentiation shapes public preferences for a more differentiated EU, and whether this varies according to how people think about EU membership. Combining recent survey data with data showing country-level exposure to differentiated integration, I show that those who have been exposed to more of it are likelier to want a future EU that allows countries to opt out of undesirable policy integration. This effect is particularly strong among Eurosceptics. My results suggest that exposure to the differentiated integration often used to integrate highly salient policy areas can solidify popular support for a less integrated EU, complicating the EU’s move towards an ‘ever closer Union’."
keywords: "differentiated integration, public opinion, Euroscepticism, European integration, contestation of the EU"
geometry: margin=1in
mainfont: cochineal
#sansfont: Linux Biolinum O
fontsize: 12pt
endnote: no
# pandocparas: TRUE
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: ""
  removetitleabstract: ""
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
header-includes:
  - \usepackage{longtable}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{setspace}\doublespacing
---

```{r setup, include=FALSE}
is_docx <- knitr::pandoc_to("docx")
is_latex <- knitr::pandoc_to("latex")

# I don't know how Texas Instruments smart this is, but p-sure default DPI is 96.
# That's not a problem for LaTeX, but it looks not-so-great for Word.
# For Word, let's up that to 600. This should allow for cross-referencing 
# in {bookdown} while allowing for conditional DPI

if (is_latex) {
  conditional_dpi <- 300
  return(conditional_dpi)
} else {
  conditional_dpi <- 600
  return(conditional_dpi)
}

knitr::opts_chunk$set(echo = FALSE,
                      message=FALSE, warning=FALSE,
                      fig.path='doc/figs/',
                      cache.path = 'doc/_cache/',
                      fig.width = 8.5, dpi = conditional_dpi,
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })

library(dplyr)
library(magrittr)
# library(tidyverse)
library(stevemisc)
library(modelsummary)
library(huxtable)
library(kableExtra)

allrds <- list.files(pattern = ".rds", path = "data") %>%
  stringr::str_remove(., ".rds")

for(i in allrds) {
      filepath <- file.path(paste0("data/",i,".rds"))
      assign(i, readRDS(filepath))
    }

```

# Introduction
Despite being among the world’s most advanced political cooperations, the expansion of the EU’s policy remit after the ratification of the Maastricht Treaty has made its integration ever-more differentiated [@leuffen_differentiated_2013; @schimmelfennig_ever_2020]. Internal EU differentiation typically takes two forms: First, some member states have opted out of cooperation to protect their own sovereignty. This is called “constitutional differentiation”. Second, through a process known as “instrumental differentiation”, countries with a diminished capacity for integration have been allowed to integrate at different schedules. This can happen either voluntarily or as a result of EU institutions imposing it as a precondition for membership [@schimmelfennig_instrumental_2014]. The differentiation of current European integration makes ‘variable geography’ [@stubb_categorization_1996] a better description of today’s EU than ‘ever closer Union’.

While increasing attention is being paid to why people want a differentiated EU [@leuffen_public_2022; @telle_elite_2022], we know little about how exposure to it has impacted support for a more differentiated future EU. While one of the few existing contributions examining the effect of differentiation on public opinion shows that it may strengthen the perceptions of the EU’s input legitimacy [@schraff_does_2020], two knowledge gaps remain: We still know too little about how past exposure to differentiation shapes support for it. Second, our understanding of how the relationship between exposure to differentiation and support for it is conditioned by attitudes towards the EU is still incomplete. Whether there is such a conditioning effect, either in the broader EU population, or where differentiation has been implemented in either a dominating or voluntary fashion is also unknown.

By combining survey data on citizens’ attitudes towards constitutional differentiated integration from 2020-2021 with data detailing differentiations from the EU’s treaties, I find that support for constitutional differentiation increases with exposure to differentiations from EU law, and that the extent to which this is the case strongly depends on your past attitudes towards the EU. However, I find that both where differentiation has been voluntary chosen, as in Scandinavia, or externally imposed in a discriminatory fashion, as in Central and Eastern Europe. While the reason for the generally greater support for differentiated integration among Eurosceptics in the countries that have been exposed to it may be that Eurosceptics are more likely than others to believe that differentiated integration is a path towards a less integrated Europe, more research is needed to understand why there is no real effect of pre-existing attitudes towards the EU in two regions with very different experiences of differentiation.

My findings are important to the discussion about the EU’s future spawned by Brexit [@european_commission_white_2017]: Because EU integration has expanded into increasingly salient areas, differentiation may be called upon to dampen public contestation of it [@schimmelfennig_ever_2020]. By providing greater numbers of citizens with exposure to differentiation, the EU’s institutions may solidify opposition to further integration among the Eurosceptics that have on several instances been able to put the brakes on further integration [@hooghe_postfunctionalist_2009].

Furthermore, the article contributes to the broader emerging literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009] While the EU’s legal and functional reach is much greater than that of other international organizations, there has been a turn towards expanding both the political and legal powers of many of the organizations constituting the global governance system [@zurn_theory_2018]. Even if more research is needed to understand whether my results travel to contexts other than the EU, my article helps shine a light on how public support for deeper political integration is impacted by non-uniform implementations of this emerging legal order.

# Public support for European integration - what we know and what we don't
A large literature investigates individual preferences for European integration [@basile_sharing_2020; @bremer_junckers_2020; @de_vries_variable_2013; @eichenberg_post-maastricht_2007; @gabel_public_1998; @hobolt_public_2015; @hooghe_calculation_2005]. The literature identifies three mechanisms that help shape attitudes towards the EU. The first is utilitarian, with support for the EU contingent on whether individuals or social groups with which they identify stand to gain from integration [@foster_economic_2021; @gabel_public_1998]. The second is cue-based, and posits that citizens use heuristics such as elite discourses, media messages, and national benchmarks to form opinions about European integration  [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007; @vries_euroscepticism_2018]. The literature also shows that strong identification with the nation-state makes citizens more likely to oppose EU integration [@borzel_identity_2020; @hooghe_calculation_2005; @luedtke_european_2005].

An under-investigated issue is why some people support a more differentiated EU. This gap is particularly pronounced in light of the large that literature already investigates differentiated integration conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004]. The few contributions investigating individual support for differentiated integration has focused on liberal economic values [@leuffen_public_2022], Euroscepticism [de_blok_blessing_2021], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite_2022].

Thus, with one notable exception, national contextual factors are under-studied in this literature. However, these contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; harteveld_europe_2013]. One contextual source of support for differentiated integration may be the exposure that citizens of EU countries have had to it: A country’s number of differentiations from EU law may bring its level of integration in and out of alignment with what a citizen would prefer. If there is a misalignment between the actual and desired levels of integration, citizens may respond by adjusting their stated preference for future differentiated integration [@busemeyer_beyond_2021; @wlezien_public_1995]. While previous literature investigates the potential feedback loops between EU policy-making and public opinion [@bolstad_dynamics_2015; @de_bruycker_democratically_2020] this paper shifts the focus to both how support for EU differentiation is impacted by exposure to differentiated integration generally and whether this effect is conditioned by attitudes towards European integration.

I believe that greater exposure to differentiated integration will increase support for it as a model of future integration. This is because such exposure is likely to familiarize citizens with EU differentiation, thus increasing their knowledge of how differentiated integration would work in the future. Greater knowledge is often, though not always, associated with greater support for European integration [@marquart_knowing_2019; @verhaegen_does_2015]. I furthermore find it theoretically plausible that whether this is the case will depend on whether one identifies with Eurosceptic sentiments or not: Because constitutional differentiated integration allows for permanent opt-outs from specific policy areas, Eurosceptics are more likely to see differentiated integration as a viable path towards the less differentiated Europe that they want to see in the future. This yields two complementary hypotheses:

*H1a: Exposure to differentiated integration increases support for future differentiated integration.*

```{r h1a, eval=!is_latex, results="asis"}
cat('> **H1a**: Exposure to differentiated integration increases support for future differentiated integration. ')
```

*H1b: The effect of exposure is larger among Eurosceptics than supporters of EU membership.*

```{r h1b, eval=!is_latex, results="asis"}
cat('> **H1b**: The effect of exposure is larger among Eurosceptics than supporters of EU membership.')
```

Responses to differentiated integration may further vary based on whether differentiation was imposed by the EU’s institutions or voluntarily chosen by member state government. An important reason for this is that citizens may perceive such differentiation as the EU dominating their countries, by shutting them out of economically beneficial integration, for instance in the field of freedom of movement [@cianciara_differentiated_2014; @schimmelfennig_instrumental_2014]. As @schimmelfennig_eu_2014 shows, the countries of Central and Eastern Europe were among those exposed to the greatest amount of discriminatory, rather than preferential, differentiated integration. The reason was that existing member states wanted to shield themselves from the economic destabilization believed to follow from the EU accession of countries like Poland, Estonia, and Hungary. As figure A6 shows, citizens of Central and Eastern Europe are among those in this sample exposed to the greatest amount of differentiated integration.

It is furthermore likely that Eurosceptics in these countries will have a more adverse reaction to differentiated integration than those who support EU membership: As differentiated integration in these countries has been predominantly discriminatory, they may be more likely to identify a non-uniformly integrated EU, in which countries can choose to opt out of integration if they so choose, with the potential for domination. Given the importance this group places on autonomy they are more likely to have adverse reactions to the discriminatory imposition of differentiated integration.

Elites in Central and Eastern Europe, especially in Poland, have also framed differentiated integration as a path towards a divided EU, with a powerful core and a less powerful periphery [@cianciara_differentiated_2014 178-179]. For example, Polish politicians opposed further integration of the Eurozone, a prominent example of differentiated integration, because they feared that the Central and Eastern member states, as non-members, would lose their influence over economic policy-making by not being inside the EU’s economic policy-making structures. Because elite cues are particularly important for shaping attitudes towards less salient issues, such as differentiated integration [@steenbergen_whos_2007; @telle_elite_2022], a reasonable assumption is that citizens would respond to these discourses by adjusting their preferences for integration downwards. Thus, two different causal mechanisms could produce the same outcome.

However, this history might be less relevant if it not being used by current actors. Instead, the increasingly Eurosceptic elite attitudes in Central and Eastern Europe [@borzel_identity_2020; @brack_towards_2020; @kriesi_politicization_2016] might form a more potent backdrop against which Central and Eastern European citizens might be expected to make up their minds. Exposure to such cues might, in contrast to what was stated before, make citizens more likely to express an increased demand for differentiated integration. This yields two competing hypotheses:

*H2a: Exposure to differentiated integration will lead to lower levels of support for future differentiated integration in Central and Eastern Europe than in the rest of the EU.*

```{r h2a, eval=!is_latex, results="asis"}
cat('> **H2a**: Exposure to differentiated integration in Central and Eastern will lead to lower levels of support for future differentiated integration than in the rest of the EU.')
```

*H2b: Exposure to differentiated integration in Central and Eastern Europe will lead to lower levels of support for it among Eurosceptics in these countries compared to supporters of EU membership.*

```{r h2b, eval=!is_latex, results="asis"}
cat('> **H2b**: Exposure to differentiated integration in Central and Eastern Europe will lead to lower levels of support for it among Eurosceptics in these countries compared to supporters of EU membership.')
```

*H3: Exposure to differentiated integration in Central and Eastern Europe will lead to higher levels of support for it than in the rest of the EU.*

```{r h3, eval=!is_latex, results="asis"}
cat('> **H3**: Exposure to differentiated integration in Central and Eastern Europe will lead to higher levels of support for it than in the rest of the EU.')
```

If exposure to discriminatorily imposed differentiation is assumed to lead to less support for differentiated integration, especially among Eurosceptics, the opposite is likely to be the case for countries with a great deal of voluntarily chosen differentiation. Scandinavia offers an interesting case for testing this assumption. Here the mode of implementation has been drastically different from that found in Central and Eastern Europe, as differentiation was both desired by national governments and justified through the autonomy-enhancing aspects of differentiated integration [@adler-nissen_opting_2014; @leruth_operationalizing_2015; @spendzharova_banking_2016]. Because of the previously explained cueing mechanism one might expect this pro-autonomy discourse to produce greater support for differentiated integration in Scandinavia than elsewhere in the EU.

Previous literature has found that the renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy among Danish Eurosceptics [@schraff_does_2020]. This suggests both that citizens may respond positively to differentiation and that past exposure could make them more favourably inclined to it in the future. This is particularly likely to be true for Scandinavian Eurosceptics: These citizens may, because their countries are both relatively small and highly dependent on European market access, be more likely to see differentiation as a viable second-best alternative to the European disintegration they would otherwise favour. More generally it may also be the case that, because they are sceptical of EU integration, they are more likely to see differentiated integration as a goal in itself as it represents the kind of less integrated Europe that they would be likely to favour. I hypothesize the following:

*H4a: The positive effect of exposure to differentiated integration will be greater for citizens of Sweden and Denmark than in the EU as a whole.*

```{r h4a, eval=!is_latex, results="asis"}
cat('> **H4a**: The positive effect of exposure to differentiated integration will be greater for citizens of Sweden and Denmark than in the EU as a whole')
```

*H4b: The positive effect of exposure to differentiated integration will be greater for Scandinavian Eurosceptics than supporters of EU membership.*

```{r h4b, eval=!is_latex, results="asis"}
cat('> **H4b**: The positive effect of exposure to differentiated integration will be greater for Scandinavian Eurosceptics than supporters of EU membership.')
```

My paper thus investigates the impact of exposure to differentiation on two levels: I first investigate the general effect of exposure to differentiated integration in the EU as a whole, as well as in Scandinavia and Central and Eastern Europe. I then test whether this relationship is moderated by pre-existing Eurosceptic sentiments. Such a moderating effect is particularly likely to exist in either of the two regions because of the way that voluntary or discriminatory differentiation may be perceived as either an expansion or diminishing of the autonomy that this group is particularly likely to value.

# Data and methods
```{r load data}
EUI_final_DI_disc <- readRDS("data/EUI_final_DI_disc.Rds")
nordic_data <- readRDS("data/nordic_data.Rds")

```

I investigate these hypotheses using multilevel linear models and data from Yougov surveys fielded in 2020-21 [@hemerijck_eui-yougov_2021]. The sample includes respondents from Denmark, Finland, France, Germany, Sweden, Greece, Hungary, Italy, Lithuania, the Netherlands, Poland, Romania, and Spain and has an *N* of `r nrow(EUI_final_DI_disc)`. The surveys include two questions about differentiated integration. The first asks whether respondents want an EU where countries may choose to opt out of certain policy integration. This is called constitutional differentiation. The second asks whether the EU should allow countries to integrate at different speeds, while maintaining the goal of uniform integration in the long run. This is an instrumentally differentiated EU. This article focuses on popular support for the former, as its allowance for an EU that is permanently segmented into multiple policy communities represents arguably the sharpest departure from the EU’s normative goal of an “ever closer Union”.

I combine these data with a novel dataset [@schimmelfennig_eudiff1_2022] that features all differentiations from the EU treaties between 1952-2020. The data shows each instance of differentiation a country has been exposed to, the number of articles involved, and how long it was enforced. The dataset contains data on differentiations from EU treaties, or primary law, rather than differentiations from the less salient secondary law that constitutes the EU’s main legislative output [@duttle_opting_2017]. I use treaty differentiations because these are typically more publicly salient and subject to greater debate. Citizens are thus more likely to be aware of them than differentiations in secondary law.

## Dependent variable
My dependent variable measures support for constitutional differentiated integration. The variable uses a five-unit Likert scale and asks respondents whether they support a constitutionally differentiated EU (“Please tell us how far you agree or disagree with the following statement: Member states should be allowed to opt out of specific areas of European integration. This means that a member state can negotiate exceptions ("opt-out") for areas in which it does not wish to
cooperate. For example, Denmark has opted out of the common currency, and Poland has
opted out of the EU's Charter of Fundamental Rights.)". The response categories are ordered from 1-5, with 1 being Strongly agree and 5 being strongly disagree. I reverse the variable coding, so that higher values indicate agreement. By not asking about “constitutional differentiated integration”, the question becomes easier to understand. It also frames differentiation as a tool for strengthening autonomy and clarifies that countries can choose to opt out.

## Variables of key interest
To operationalize each country’s exposure to EU differentiation I multiply, for each differentiation, the number of articles that a country was exempt from with the number of years the differentiation was in force. This is then weighted based on the length of time that has passed since the differentiation ceased to be enforced, and finally log-transformed.

Operationalizing exposure to differentiation in this way has three main benefits: First, it captures both the temporal and functional scope of a country’s exposure to differentiation. Second, the weighting scheme means that those instances of differentiation that are still in force are given the most weight when investigating differentiation’s impact on support for differentiated integration in the future. This is likely to correspond to how respondents relate their own experience of differentiation to support for it: While many respondents from Central and Eastern Europe may still remember the particulars of their countries’ exposure to externally imposed differentiation, possibly producing a cohort effect among these individuals [@lauterbach_europe_2020], Southern European respondents are less likely to remember that they were exposed to externally imposed differentiation several decades ago and use this as an informational backdrop for their attitude-formation. Finally, the log-transformation reduces the influence of outliers.

To analyse the regionally specific effects of differentiation, I create dummies for Scandinavian and Central and Eastern European citizenship. I code all Swedish and Danish citizens with a 1 on the Scandinavian dummy. To create a Central and Eastern European dummy I do the same for everyone from Hungary, Lithuania, Poland, and Romania. These are used to investigate the interaction effects assumed by H2-H4. I also create a dummy for Euroscepticism, coding those who answer that they would vote yes to leaving the European Union in a hypothetical referendum with a 1 and those answering no with a 0.

I operationalize support for liberal economic policies, which has been found by previous literature to be a key determinant of support for differentiated integration [@leuffen_public_2022], through a question that asks whether respondents want to live in a Europe that “stresses economic integration, market competition and fiscal discipline”. This is contrasted to “a global Europe that acts as a leader on climate, human rights and global peace” or “a protective Europe that defends the European way of life and welfare against internal and external threats”. I code those who say that they prefer to live in a primarily market-oriented Europe as 1 and others 0 on this variable. While the question is multi-faceted, by asking about both fiscal discipline that may be described as austerity policies and more general market integration, I believe it still measures a general affinity for economic liberalism.

I include an equally weighted index of questions measuring respondents’ evaluation of the national economy. A factor analysis finds that these questions, in contrast to one asking about the respondent’s perceptions of the local job market, measure the same underlying dimension (see table A1). Previous literature shows that positive sociotropic economic evaluations correlate with support for both uniform and differentiated integration (see for instance @gabel_public_1998; @harteveld_europe_2013; @hooghe_calculation_2005; @leuffen_public_2022). Egotropic evaluations have, however, been shown to have little relevance for explaining support for differentiated integration [@leuffen_public_2022]. I measure support for national governments, relevant for explaining support for differentiated integration [@armingeon_loss_2014; @de_blok_blessing_2021; @harteveld_europe_2013], through a dummy asking respondents whether they believe their nation’s government is doing a good job.

I operationalize identity through a widely used question that asks respondents to rank the inclusiveness of their identities. I construct a dummy where those answering that they identify solely with their nation-states are coded as 1 and everyone else as 0. While nested identities combining a national and European component are common [@risse_euro_2003; @starke_concept_2021], previous literature finds the distinction between exclusively national and partially inclusive identities to be more important for producing Eurosceptic sentiments than variations between different nested identities [@hooghe_calculation_2005].

I also use ideology and gender as socio-political control variables. These are established predictors of support for both general and specific integration (see for example @carrubba_decision_2004; @hobolt_public_2015; @hooghe_calculation_2005; @schoen_identity_2008). I operationalise ideology through a 7-unit scale, with 0 representing far-left and 6 far-right positions. As previous literature also finds evidence of a curvilinear correlation between ideology and support for the EU [@toshkov_beyond_2022; @van_elsas_changing_2015], I add a squared indicator of left-right ideology.

```{r create descriptive data}
data_descriptives <- readRDS("data/Data_descriptives.Rds")

data_descriptives <- data_descriptives %>%
  rename(`Support constitutional DI` = opt_out_supp) %>%
  rename(`Article-years (logged)` = article_disc_log) %>%
  rename(`Liberal economic values` = market_europe) %>%
  rename(`Perception of economy` = alternative_econ) %>%
  rename(`Support national government` = gov_trust) %>%
  rename(`Left-right` = left_right) %>%
  rename(`Left-right (sqr.)` = left_right_sqr) %>%
  rename(`Eurosceptic` = eurosceptic) %>%
  rename(`Gender` = gender) %>%
  rename(`Exclusive identity` = excl_identity)
```

```{r}
library(modelsummary)

table_format <- ifelse(is_docx, "huxtable", 'default')

modelsummary::datasummary_skim(data_descriptives, fmt = 3, histogram = FALSE, 
                 title = "Descriptive statistics", output = "latex")

```

## Limitations
A limitation of the study is that both surveys were fielded during the COVID-19 pandemic, which could impact support for differentiation. However, @cicchi_eu_2020 show that solidarity, as is common, still seemed to reside at the national level in the early phase of the crisis. This means that my results may generalize to periods beyond those in which the survey was fielded. To account for how different baseline levels of support for differentiation are likely to exist in each country, I use multilevel models with random country intercepts.

There are also large amounts of missingness on certain variables (see table 1). Missing data may bias the results if they are correlated with individual-level characteristics like ideology or gender. I test the effect of such bias by multiply imputing the data [@buuren_mice_2011]. I show the results of this test in the section Robustness checks.

A further limitation, general to all studies using observational survey data, is the challenge of omitted variable bias. In this study omitted variable bias is particularly likely to stem from the absence of data on the respondents’ knowledge of the EU and education [@karp_know_2003; @marquart_knowing_2019]. I use sensitivity analysis [@cinelli_making_2020] to tentatively quantify the size of this bias and show the results of the analyses in the section Robustness checks.

# Results and discussion
I first show the effect of exposure to differentiation in the broader population, and especially how this interacts with already pre-existing Euroscepticism. I then discuss the effect of differentiation on support for it in both Scandinavia and Central and Eastern Europe.

```{r analysis 1}
library(lme4)
library(merTools)

```

```{r }

main_mod = list()

main_mod$"Euroscepticism" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                                                         eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + (1|country), weights = EUI_final_DI_disc$weight, 
                                                         data = EUI_final_DI_disc)

coef_map_eurosceptic <- c('(Intercept)' = 'Constant',
                          'article_disc_logeurosceptic1' = 'Article-years (logged) X Euroscepticism',
                          "article_disc_lognordic_dummy1" = "Article-years (logged) X Scandinavia", 
                          "article_disc_logCEE_dummy1" = "Article-years (logged) X CEE",
                          "nordic_dummy1" = "Scandinavia",
                          "CEE_dummy1" = "Central and Eastern Europe",
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'excl_identity' = 'Exclusively national identity')

modelsummary(main_mod, fmt = 3, coef_map = coef_map_eurosceptic, output = "latex", 
             stars = TRUE, title = "Model showing interaction between Euroscepticism and exposure to DI")
```

Both table 2 and figure 1 strongly support *H1a* and *H1b*, by confirming that exposure to differentiated integration is associated with greater support for differentiation as a future model for the EU. The effect of exposure is also significantly greater among Eurosceptics. The reason may be that Eurosceptic citizens in countries with greater exposure to differentiated integration are likelier to be familiar with the structure of a differentiated EU membership than those in countries not exposed to any differentiation. They may also see it as a more viable “second-best” alternative to full disintegration. However, even if the effect is larger for Eurosceptic than non-Eurosceptic citizens, the results suggest that future demands for differentiated integration among those most strongly exposed to it are unlikely to be restricted only to Eurosceptics.

```{r, fig.pos="H", fig.cap="Predicted probability of support for constitutional differentiation as a function of the interaction between article-years (logged) and Euroscepticism. 95% CIs."}

library(sjPlot)
library(ggplot2)
pred_plot_eurosceptic <- plot_model(main_mod$`Euroscepticism`, type = "pred", terms = c("article_disc_log", "eurosceptic"), 
                                    colors = "bw", legend.title = "Eurosceptic")

pred_plot_eurosceptic_final <-  pred_plot_eurosceptic + ggtitle("Support for constitutional differentiation", 
                                                                subtitle = "Support for differentiation among Eurosceptics and non-Eurosceptics") + labs(x = "Differentiation (article-years (logged))",
                                                                                                                     y = "Predicted value") + scale_linetype_manual(name = "Eurosceptic",
                           values = c("dashed", "solid", "dotted"),
                           labels = c("Non-Eurosceptic", "Eurosceptic"))

pred_plot_eurosceptic_final <- pred_plot_eurosceptic_final + font_size(title = 15, labels.x = 15, labels.y = 15)

pred_plot_eurosceptic_final
```

A plausible counter-argument to my hypothesized causal mechanism could be that differentiation was primarily chosen where underlying Euroscepticism, and thus a demand for differentiation, was already high. Figure A6 offers some support for such an assumption, as we find high levels of differentiated integration in Sweden and Denmark. These are countries where opt-outs from integration have been deployed as a response to a Euroscepticism shared by both political parties and voters [@leruth_operationalizing_2015; @spendzharova_banking_2016]. However, countries as diverse as Lithuania, Hungary, Italy, and the Netherlands have also seen differentiations from EU treaties. Differentiated integration is thus not a phenomenon restricted to the countries found in what @bolstad_dynamics_2015 identifies as a Eurosceptic periphery. This strengthens the case that the effect is not only driven by an endogenous relationship between Euroscepticism and differentiation.

I next analyse how support for constitutional differentiation is shaped by whether differentiation is voluntarily chosen or involuntarily imposed. I do this by comparing the interaction effects of Scandinavian and Central and Eastern European citizenship.

```{r }

main_regional = list()

main_regional$"Regional dummies" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                                               eurosceptic + gender + excl_identity + nordic_dummy + CEE_dummy + article_disc_log*nordic_dummy + article_disc_log*CEE_dummy +  (1|country), 
                                               weights = EUI_final_DI_disc$weight, data = EUI_final_DI_disc)

library(modelsummary)
library(dplyr)
coef_map_regions <- c('(Intercept)' = 'Constant',
                          'article_disc_lognordic_dummy1' = 'Article-years (logged) X Scandinavia',
                      'article_disc_logCEE_dummy1' = 'Article-years (logged) X CEE',
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'excl_identity' = 'Exclusively national identity', 
                      "article_disc_logeurosceptic1" = "Article-years (logged) X Eurosceptic")

modelsummary(main_regional, fmt = 3, output = "latex", coef_map = coef_map_regions, 
             stars = TRUE, title = "Model showing regional interactions")

```

Table 3 shows a positive and statistically significant effect of the interaction between exposure to differentiated integration and Scandinavian citizenship. However, it finds no statistically significant effect of the interaction between Central and Eastern European citizenship and exposure to differentiated integration.

```{r, create plot}
library(sjPlot)
library(ggplot2)
#Scandinavia
pred_plot_nordics <- plot_model(main_regional$`Regional dummies`, type = "pred", terms = c("article_disc_log", "nordic_dummy"), 
                                legend.title = "Scandinavia", colors = "bw")

pred_plot_nordics_final <-  pred_plot_nordics + ggtitle("Scandinavia") + labs(x = "Differentiation (article-years (logged)",
                                                                                                                           y = "Predicted value")
pred_plot_nordics_final <- pred_plot_nordics_final + font_size(title = 15, labels.x = 15, labels.y = 15) + scale_linetype_manual(name = "Scandinavia",
                           values = c("dashed", "solid"),
                           labels = c("Non-Scandinavian", "Scandinavian"))

##CEE
pred_plot_CEE <- plot_model(main_regional$`Regional dummies`, type = "pred", terms = c("article_disc_log", "CEE_dummy"), 
                                    colors = "bw", legend.title = "CEE")

pred_plot_CEE_final <-  pred_plot_CEE + ggtitle("Central and Eastern Europe") + labs(x = "Differentiation (article-years (logged))",
                                                                                                                     y = "Predicted value") + scale_linetype_manual(name = "CEE",
                           values = c("dashed", "solid"),
                           labels = c("Non-CEE", "CEE"))

pred_plot_CEE_final <- pred_plot_CEE_final + font_size(title = 15, labels.x = 15, labels.y = 15)
```

```{r, fig.pos="H", fig.cap="Variations in support for differentiation as a function of exposure to DI (A = Scandinavia and B = CEE)", fig.height=8.5}
library(patchwork)

regional_final <- pred_plot_nordics_final/pred_plot_CEE_final + plot_annotation(tag_levels = "A", title = "Support for constitutional differentiation", subtitle = "Predicted values for regional interactions of exposure to differentiation in Scandinavia and Central and Eastern Europe", theme = theme(plot.title = element_text(size = 16)))

regional_final
```

Figure 2 supports this picture: Scandinavian citizens appear to become more favourably inclined towards differentiated integration as their exposure to it increases, compared to those from other regions. However, this effect is mainly present at the lower end of exposure to differentiation. Thus,I am able to confirm *H4a*, as distinct differences do exist within certain ranges of the variable. However, because of the large overlap for the largest range of the exposure variable this statistical significance must not be taken as indicative of a substantively large interaction effect.

Similarly, the predicted values show no significant difference between support for differentiation in Central and Eastern Europe and the rest of the EU. This leads me to reject *H2a*. I do, however, find some support for *H3*, whose underlying assumption is that exposure to Eurosceptic cues can lead to stronger support for differentiation even if it was imposed in a discriminatory fashion.

```{r}
region_mod <- readRDS("data/region_mod.Rds")

modelsummary::modelsummary(region_mod, fmt = 3, output = "latex", coef_map = coef_map_eurosceptic, 
             stars = TRUE, title = "Models showing interactions of Euroscepticism and region")
```

```{r create plot2}
library(sjPlot)
library(ggplot2)
#Scandinavia
pred_plot_nordics_eurosceptic <- plot_model(region_mod$Scandinavia, type = "pred", terms = c("article_disc_log", "eurosceptic"), 
                                legend.title = "Eurosceptic", colors = "bw")

pred_plot_nordics_eurosceptic_final <-  pred_plot_nordics_eurosceptic + ggtitle("Scandinavia") + labs(x = "Differentiation (article-years (logged))",
                                                                                                                           y = "Predicted value")
pred_plot_nordics_eurosceptic <- pred_plot_nordics_eurosceptic_final +
scale_linetype_manual(name = "Eurosceptic",
                           values = c("dashed", "solid"),
                           labels = c("Non-Eurosceptic", "Eurosceptic"))

##CEE
pred_plot_CEE_eurosceptic <- plot_model(region_mod$CEE, type = "pred", terms = c("article_disc_log", "eurosceptic"), 
                                legend.title = "Eurosceptic", colors = "bw")

pred_plot_CEE_eurosceptic_final <-  pred_plot_CEE_eurosceptic + ggtitle("Central and Eastern Europe") + labs(x = "Differentiation (article-years (logged))",
                                                                                                                           y = "Predicted value")
pred_plot_CEE_eurosceptic_final <- pred_plot_CEE_eurosceptic_final +  scale_linetype_manual(name = "Eurosceptic",
                           values = c("dashed", "solid"),
                           labels = c("Non-Eurosceptic", "Eurosceptic"))
```

```{r, fig.pos="H", fig.cap="Interaction between Euroscepticism and region. 95% CIs", fig.height=8.5}
library(patchwork)

regional_euroscepticism_final <- pred_plot_nordics_eurosceptic/pred_plot_CEE_eurosceptic_final + plot_annotation(tag_levels = "A", title = "Support for constitutional differentiation", subtitle = "Interaction of Euroscepticism and exposure to differentiation in CEE and Scandinavia", theme = theme(plot.title = element_text(size = 16)))

regional_euroscepticism_final
```

I must also reject *H2b*, as the effect of the interaction between differentiated integration and Euroscepticism is statistically insignificant in the Central and Eastern Europe data. I furthermore reject *H4b*, as there is no statistically significant difference between the effect of exposure for Eurosceptics and supporters of EU membership in Scandinavia. The lack of a lower level of support for differentiated integration among Central and Eastern Europeans, despite the fact that the discriminatory imposition of differentiation has been especially prevalent in these countries, suggests that historical trajectories and past experience with integration may predominantly lead to altered support for differentiation when it is used as a part of elite discourse. This finding is important for our understanding of how attitudes towards greater differentiation can be shaped by political elites in the future.

The fact that the surveys were fielded at the height of the COVID-19 pandemic means that we may be seeing lower levels of support for differentiated integration than we otherwise would: As @genschel_postfunctionalism_2021 argue, the COVID-19 pandemic “debordered” solidarity, making the prospect of mutual aid across member state borders more attractive to EU citizens. In times where EU solidarity is framed as a net positive for its citizens, support for differentiation may be lower than it otherwise would have been. More research is thus needed to establish whether the patterns established by this paper are consistent also in the future.

# Robustness checks
I use an approach outlined in @cinelli_making_2020 to test the sensitivity of several of my models to omitted variable bias (see figures A1-4). This quantifies the remaining variance that a potential confounder would need to account for to nullify the effect of theoretical interest, exposure to differentiated integration. A test of the sensitivity of the interaction between Euroscepticism and exposure to differentiation in Scandinavia (figure A3) shows that a variable explaining 50% of the remaining variance of support for constitutional differentiation would need to explain less than 10% of the remaining variance in real exposure to differentiation to nullify the effect of exposure. One potentially important omitted variable is knowledge of the EU [@leuffen_public_2022]. While knowledge of the EU is unlikely to be strongly connected with actual levels of differentiation, its strong connection to support for differentiation makes its omission particularly problematic. This suggests that omitted variable bias is likely to be an analytical concern throughout the entire study. Future studies should mitigate these concerns by including knowledge questions about current EU politics, to allow researchers to include similar controls as those commonly found in other EU public opinion research.

I also use multiple imputation through chained equations [@buuren_mice_2011; @fjelstul_explaining_2022] to test how sensitive the results are to bias caused by underlying patterns of data missingness. Multiple imputation uses each unit’s observed data to derive probable values for all covariates where a unit has missing data. This procedure is iterated many times, before the estimates derived from analysis of each imputed dataset are pooled. This procedure better accounts for the inherent uncertainty associated with data imputation and will thus yield more realistic parameter estimates and standard errors than traditional mean or zero imputation. As appendix tables A2-3 show, the multiply imputed models are similar to those found in tables 2-3. One important exception is the effect of exposure to differentiated integration in Central and Eastern Europe: While there is a weak positive effect of exposure in the main model, the multiply imputed model shows a weakly negative effect. However, both are statistically insignificant. Despite this minor deviation, the tests suggest that the effects established in this paper are not primarily driven by systematic data missingness.

I also test whether the substantial overlap between Scandinavian Eurosceptics and non-Eurosceptics is driven by the very low variability, with a standard deviation of `r round(sd(nordic_data$article_disc_log), digits = 2)`,  in the measure of exposure to differentiated integration in the Scandinavian data. I do this by testing a three-way interaction between Scandinavian citizenship, exposure to differentiated integration and Euroscepticism in the full data. Appendix table A6 suggests that the results are substantially similar to the interaction effect of Scandinavian citizenship and Euroscepticism shown in table. The results thus suggest that the estimates are not unduly biased by the very small standard deviation .

Lastly, I test how robust the results are to an alternative operationalization of Euroscepticism: satisfaction with EU democracy. Dissatisfaction with the state of EU democracy may be seen as a form of "regime scepticism", characterized by a dissatisfaction with EU decision-making, than "exit scepticism", as expressed through a desire to leave the union altogether [@vries_euroscepticism_2018]. Figure A5 ( coefficients in table A4) shows similar trends for both operationalizations: Those least satisfied with EU democracy, and who thus express the greatest amount of regime scepticism, also express the greatest support for constitutional differentiated integration. Their support for differentiation also increases more strongly as a function of exposure to previous differentiation than what is the case for those more satisfied with EU democracy. This strengthens the robustness of my original conclusions.

# Concluding discussion

This paper establishes, through a combination of data surveying citizen preferences for differentiated integration and their countries’ actual exposure to differentiation from the EU treaties, that there is a strong association between support for differentiated integration and past exposure to it. This may be because citizens more accustomed to differentiated integration potentially perceive it as a favourable way of combining the benefits of national membership of the EU with greater national autonomy. Importantly, I show that a potential tendency to see the kind of exposure that one has been exposed to as a preferable model for the future EU is, broadly speaking, conditioned by whether one opposes or supports EU membership. However, the results do not suggest that the mode of implementation of the differentiation that the same citizens have been exposed to play any large role in shaping attitudes towards a future differentiated EU.

What this suggests is that even if some differentiated integration could be perceived as a cartelization of the benefits of integration and as a dominating imposition on the political freedom of a polity, this does not necessarily translate to less support for differentiated integration as a policy. While my article does not establish why this is the case, it does strongly suggest that there is only a limited learning effect from exposure to differentiated integration. This is important because it implies that even the dominating imposition of a particular set of policies may not be enough to appreciably alter attitudes towards the same policy even among those who may be most likely to be critical of EU integration. Understanding what kind of discourses are more likely to produce critical attitudes towards differentiated integration in this group is an important question for future research.

Another question is whether the public is equally likely to prefer differentiation in all policy areas. Existing research [@de_blok_blessing_2021] has largely focused on individual support for differentiated integration of the EU as a polity. It is both possible and probable that even those favouring an “ever closer Union” may have diverging preferences for integration at the level of individual policies. Testing where individual citizens may prefer more differentiation may also help theorize the ambivalent attitudes towards EU integration found by an increasingly prevalent strand of the EU public opinion literature [@de_vries_variable_2013; @goldberg_one_2021; @moland_constraining_2022].

My results have policy implications for both the EU and other international organizations seeking to deepen their own political integration. They suggest first that allowing for a greater institutionalization of differentiated integration may undermine the public legitimacy of the EU’s goal of ‘ever closer Union’. However, beyond the EU, the results also suggest that the public legitimacy of deeper political integration may be challenged if citizens gradually become accustomed to more differentiated and non-uniform forms of integration. While one should be wary of generalizing findings from the EU to other international organizations, my results suggest that allowing some countries to opt out of the legal order that is increasingly developing beyond the EU [@zurn_theory_2018] may lead to greater popular opposition to the idea that countries should be members of entire frameworks of political cooperation, rather than more functionally limited partnerships for solving particular problems.

# References
<div id="refs"></div>

# Appendix
\setcounter{table}{0}
\renewcommand*{\thetable}{A\arabic{table}}
```{r, results='asis'}

appendix_items <- readRDS("data/appendix_items.Rds")
library(psychTools)

cat(fa2latex(appendix_items$`Factor analysis of economy`, heading = "Factor analysis of perception of economy", silent = TRUE, caption = "Factor analysis of perceptions of the economy."))
```
\setcounter{figure}{0}
\renewcommand*{\thefigure}{A\arabic{figure}}
```{r, fig.cap="Sensitivity to omitted variable bias for article-years (logged)", fig.height=8.5}
library(sensemakr)
appendix_items$"Fixed effect of non-interacted model" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
                          eurosceptic + gender + excl_identity + factor(country), weights = EUI_final_DI_disc$weight, 
                        data = EUI_final_DI_disc)

#Run sensemakr
appendix_items$"Sensitivity test of non-interacted model" <- sensemakr::sensemakr(appendix_items$`Fixed effect of non-interacted model`, 
                                treatment = "article_disc_log")

plot(appendix_items$"Sensitivity test of non-interacted model", type = "extreme")

```

```{r, fig.cap="Sensitivity to omitted variable bias for article-years (logged) X Euroscepticism", fig.height=8.5}

appendix_items$"Fixed effect of interacted model" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
                        eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + factor(country), weights = EUI_final_DI_disc$weight, 
                      data = EUI_final_DI_disc)

appendix_items$"Sensitivity test of interacted model" <- sensemakr(appendix_items$`Fixed effect of interacted model`, 
                                  treatment = "article_disc_log:eurosceptic1")

plot(appendix_items$`Sensitivity test of interacted model`, type = "extreme")

```

```{r, fig.cap="Sensitivity to omitted variable bias for article-years (logged) X Euroscepticism in Scandinavia", fig.height=8.5}
nordic_data <- readRDS("data/nordic_data.Rds")
CEE_data <- readRDS("data/CEE_data.Rds")

appendix_items$"Fixed effect of interacted model (Nordics)" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
                                                          eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + factor(country), weights = nordic_data$weight, 
                                                        data = nordic_data)

appendix_items$"Sensitivity test of interacted model (Nordics)" <- sensemakr(appendix_items$`Fixed effect of interacted model (Nordics)`, 
                                                                   treatment = "article_disc_log:eurosceptic1")

plot(appendix_items$`Sensitivity test of interacted model (Nordics)`, type = "extreme")
```

```{r, fig.cap="Sensitivity to omitted variable bias for article-years (logged) X Euroscepticism in Central and Eastern Europe", fig.height=8.5}
library(sensemakr)
appendix_items$"Fixed effect of interacted model (CEE)" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
                                                                    eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + factor(country), weights = CEE_data$weight, 
                                                                  data = CEE_data)

appendix_items$"Sensitivity test of interacted model (CEE)" <- sensemakr(appendix_items$`Fixed effect of interacted model (CEE)`, 
                                                                             treatment = "article_disc_log:eurosceptic1")

plot(appendix_items$`Sensitivity test of interacted model (CEE)`, type = "extreme")

```

```{r}

library(mice)
library(miceadds)
library(merTools)
main_impute <- readRDS("data/main_impute.Rds")
```

```{r}
coef_map_regions2 <- c('(Intercept)' = 'Constant',
                          'article_disc_log:nordic_dummy' = 'Article-years (logged) X Scandinavia',
                      'article_disc_log:CEE_dummy' = 'Article-years (logged) X CEE',
                      "article_disc_log:eurosceptic1" = "Article-years (logged) X Eurosceptic",
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'excl_identity' = 'Exclusively national identity', 
                      "nordic_dummy" = "Scandinavia",
                      "CEE_dummy" = "Central and Eastern Europe")

modelsummary(main_impute, fmt = 3, output = "latex", coef_map = coef_map_regions2, 
             stars = TRUE, title = "Imputed model of interaction between Euroscepticism and article-years (logged) and model with regional dummies")
```

```{r}
regional_impute <- readRDS("data/regional_impute.Rds")

coef_map_eurosceptic2 <- c('(Intercept)' = 'Constant',
                          'article_disc_log:eurosceptic1' = 'Article-years (logged) X Euroscepticism',
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'excl_identity' = 'Exclusively national identity')

modelsummary::modelsummary(regional_impute, fmt = 3, output = "latex", coef_map = coef_map_eurosceptic2, 
             stars = TRUE, title = "Imputed models showing interactions of Euroscepticism and article-years (logged) by region")
```

```{r}

coef_map_alteurosceptic <- c('(Intercept)' = 'Constant',
                          'article_disc_logeu_democ_satisf' = 'Article-years (logged) X Euroscepticism',
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eu_democ_satisf' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'excl_identity' = 'Exclusively national identity')

modelsummary::modelsummary(appendix_items$`Alternative conceptualization of Euroscepticism`, fmt = 3, output = "latex", coef_map = coef_map_alteurosceptic, 
             stars = TRUE, title = "Model showing alternative conceptualization of Euroscepticism")
```

```{r, fig.cap="Predicted values of support for constitutional differentiated integration as a function of an interaction between exposure to differentiated integration and satisfaction with EU democracy. Slopes for minimum, median and maximum values of satisfaction with EU democracy are shown. 95% CIs", fig.height=8.5}

pred_plot_interact_main <- plot_model(appendix_items$`Alternative conceptualization of Euroscepticism`, type = "pred", terms = c("article_disc_log", "eu_democ_satisf"), 
                                      colors = "bw", legend.title = "EU democracy satisfaction")

pred_plot_interact_main_final <-  pred_plot_interact_main + ggtitle("Support for constitutional differentiation", 
                                                                    subtitle = "Interaction between EU democracy satisfaction and differentiation") + labs(x = "Differentiation (article-years (logged))",
                                                                                                                                                           y = "Predicted value") + scale_linetype_manual(name = "EU democracy satisfaction",
                           values = c("dashed", "solid", "dotted"),
                           labels = c("Minimum", "Median", "Max")) + font_size(title = 15, labels.x = 15, labels.y = 15)

pred_plot_interact_main_final
```

```{r, fig.cap="Number of differentiations from EU primary law 1952-2020 (log-transformed and recency-weighted article-years) by EU member state.", fig.height=8.5}
bar_chart_articleyear <- ggplot(EUI_final_DI_disc, aes(x = factor(country), y = article_disc_log)) +
  geom_col(stat = "identity", position = "dodge")

bar_chart_articleyear + labs(x = "Country", y = "Discounted article-years (log-transformed)") +
  scale_x_discrete(breaks = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14),
                   labels = c("UK", "DK", "FI", "FR", "DE", "SE", "GR", "HU", "IT",
                              "LT", "NL", "PL", "RO", "ES")) +
  ggtitle(label = "Differentiated integration in the EU", 
          subtitle = "Primary law differentiations (article-years log-transformed and weighted)") + font_size(title = 15, labels.x = 15, labels.y = 15)
```

```{r}
southern_mod = list()

southern_mod$"Southern Europe" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                           eurosceptic + gender + excl_identity + south_dummy + south_dummy*article_disc_log + (1|country), weights = EUI_final_DI_disc$weight, 
                         data = EUI_final_DI_disc)

coef_map_southern <- c('(Intercept)' = 'Constant',
                          'article_disc_logsouthern_dummy1' = 'Article-years (logged) X Southern Europe',
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'excl_identity' = 'Exclusively national identity', 
                      "south_dummy1" = "Southern Europe")

modelsummary::modelsummary(southern_mod, fmt = 3, output = "latex", coef_map = coef_map_alteurosceptic, 
             stars = TRUE, title = "Model showing effect of exposure in Southern Europe")

```

```{r, fig.pos="H", fig.cap= "Support for differentiated integration in Southern Europe as an interaction between Euroscepticism and exposure to differentiated integration. 95% CIs.", fig.height=8.5}

pred_plot_southern <- plot_model(southern_mod$`Southern Europe`, type = "pred", terms = c("article_disc_log", "south_dummy"), 
                                 colors = "bw", legend.title = "Southern Europe")

pred_plot_southern_final <-  pred_plot_southern + ggtitle("Support for constitutional differentiation (Southern Europe)", subtitle = "Support for differentiation among Southern and non-Southern Europeans") + labs(x = "Differentiation (article-years (logged))",
                                                                                                                                 y = "Predicted value")
pred_plot_southern_final + font_size(title = 15, labels.x = 15, labels.y = 15) + scale_linetype_manual(name = "Southern Europe",
                           values = c("dashed", "solid"),
                           labels = c("Non-Southern", "Southern"))
```

```{r }
scandi_eurosceptic = list()

scandi_eurosceptic$"Scandinavia (three-way interactions)" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                             eurosceptic + gender + excl_identity + nordic_dummy + article_disc_log*eurosceptic*nordic_dummy + (1|country), weights = EUI_final_DI_disc$weight, 
                           data = EUI_final_DI_disc)

coef_map_nordic <- c('(Intercept)' = 'Constant',
                          'article_disc_logeurosceptic1nordic_dummy1' = 'Article-years (logged) X Eurosceptic X Scandinavia',
                          'article_disc_log' = 'Article-years (logged)',
                          'nordic_dummy1' = "Scandinavia",
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'excl_identity' = 'Exclusively national identity')

modelsummary::modelsummary(scandi_eurosceptic, fmt = 3, output = "latex", coef_map = coef_map_nordic,
             stars = TRUE, title = "Model showing three-way interaction between Scandinavian dummy, Eurosceptic and article-years (logged)")
```

### 2022-08-30 R/1-prep.R

@@ -1,0 +1,28 @@
+# load libraries ----
+# load the functions I need to process the data
+library(stevedata) # has the raw data
+
+# "load" the data ----
+# Note: in your case, you can load raw data from wherever, even outside the project directory
+# For example, if you're analyzing ANES data, you're not ask to stick it in your project directory
+# If you can share the raw data, create a directory called "data-raw" in the main project directory
+# Then, load it from there.
+data(ESS9GB, package="stevedata")
+
+# "prep" the data ----
+# I can do anything I want here. I can recode things, transform variables, or whatever
+# Here, let's just create a stupid noise variable
+
+Data <- ESS9GB
+
+set.seed(8675309)
+Data$noise <- rnorm(nrow(Data))
+
+# in {tidyverse}
+# set.seed(8675309)
+# ESS9GB %>%
+#   mutate(noise = rnorm(nrow(.))) -> Data
+
+# ^ Notice that I "finished" my data prep into a new object, titled "Data"
+# Now: I save it to data/Data.rds
+saveRDS(Data, "data/Data.rds")

### 2022-08-30 R/3-sims.R

@@ -1,0 +1,32 @@
+# Libraries first
+library(stevemisc)
+# library(tidyverse)
+library(modelr)
+
+# Load stuff
+Mods <- readRDS("data/Mods.rds")
+Data <- readRDS("data/Data.rds")
+
+# in {tidyverse}
+# Data %>%
+#   data_grid(lrscale = unique(lrscale), .model = Mods[[1]],
+#             immigsent = 0) %>% na.omit -> newdat
+
+
+Sims <- list()
+
+na.omit(data_grid(Data, lrscale = unique(lrscale), .model = Mods[[1]], immigsent = 0)) -> newdat
+
+Sims$"SQI (Ideology)" <- get_sims(Mods[[1]], newdat = newdat, 1000, 8675309)
+
+Sims[[1]]$lrscale <- rep(seq(0:10), 1000)
+
+
+
+# in {tidyverse}
+# newdat %>%
+#   # repeat this data frame how many times we did simulations
+#   dplyr::slice(rep(row_number(), 1000)) %>%
+#   bind_cols(get_sims(Mods[[1]], newdata = newdat, 1000, 8675309), .) -> Sims$"SQI (Ideology)"
+
+saveRDS(Sims, "data/Sims.rds")

### 2022-08-30 src/get_citations.R

@@ -1,0 +1,53 @@
+
+if_not_install <- function(packages) {
+  new_pack <- packages[!(packages %in% installed.packages()[,"Package"])]
+  if(length(new_pack)) install.packages(new_pack)
+}
+
+if_not_install(c("stringr", "dplyr", "magrittr",
+                 "rmarkdown", "bib2df"))
+
+
+library(stringr)
+library(dplyr)
+library(magrittr)
+library(rmarkdown)
+library(bib2df)
+
+doc <- readLines("Manuscript_3_draft1.Rmd")
+the_entry <- yaml_front_matter("ms.Rmd")$bibliography
+
+read_bib <- function(x) {
+  if(str_starts(x, "`r ") == TRUE) {
+    ht <- str_remove_all(the_entry, "`")
+    ht <- str_remove(ht, "r ")
+
+    ht <- bib2df(eval(parse(text=ht)))
+    return(ht)
+  } else {
+    ht <- bib2df(the_entry)
+    return(ht)
+  }
+}
+
+the_bib <- read_bib(the_entry)
+
+
+cites <- c()
+
+for(i in 1:length(doc)) {
+  hold_this <- as.data.frame(str_match_all(doc[i], "@([[:alnum:]:&!~=_+-]+)")[[1]])
+  cites <- bind_rows(cites, hold_this)
+}
+
+cites %>% select(-V1) -> cites
+
+cites %>%
+  rename(BIBTEXKEY = V2) %>%
+  left_join(., the_bib) %>%
+  distinct(BIBTEXKEY, .keep_all = TRUE) %>%
+  # For now, this will tease out those wrong matches from the YAML or truly anything else
+  # It assumes, hopefully, you don't have a "gmail" entry in your .bib file or something to that effect
+  # In this application, that's where that's happening.
+  filter(!is.na(CATEGORY)) %>%
+  bib2df::df2bib(., file = "inst/refs.bib")

### 2022-08-30 src/render_abstract.R

@@ -1,0 +1,32 @@
+library(rmarkdown)
+library(ymlthis)
+library(magrittr)
+
+YAML <- yaml_front_matter("Manuscript_3_draft1.Rmd")
+
+YAML[names(YAML) %in% c("title","author","abstract","keywords","geometry","mainfont")] %>%
+  as_yml() -> A
+
+A %>%
+  yml_rticles_opts(header_includes = "\\pagenumbering{gobble}") -> A
+
+yaml::write_yaml(A, file="doc/tmp.yml")
+
+cat(paste0("---\n",readr::read_file("doc/tmp.yml"), "\n---"), file="doc/tmp.Rmd")
+
+render("doc/tmp.Rmd",
+       output_file="abstract.pdf",
+       rmarkdown::pdf_document(template = stevetemplates::templ_article2(),
+                               latex_engine = "xelatex", dev="cairo_pdf"))
+
+delfiles <- dir(path="doc/" ,pattern="tmp")
+file.remove(file.path("doc/", delfiles))
+
+# disabled for now...
+
+#tmp <- use_index_rmd(.yml = print(A), path='_dross', open_doc = FALSE)
+# render(tmp,
+#        output_file="abstract.pdf",
+#        rmarkdown::pdf_document(template = stevetemplates::templ_article2(),
+#                                           latex_engine = "xelatex", dev="cairo_pdf"))
+# file.remove(tmp)

### 2022-08-30 src/render_docx.R

@@ -1,0 +1,8 @@
+# rmarkdown::render("ms.Rmd", output_file="doc/ms.docx", rmarkdown::word_document(reference_docx = stevetemplates::templ_word()))
+# ^ old, legacy, here if you want to revert to it.
+
+rmarkdown::render("Manuscript_3_draft1.Rmd", output_file="doc/ms.docx",
+                  bookdown::word_document2(reference_docx = stevetemplates::templ_word(),
+                                           toc = FALSE, number_sections = FALSE))
+
+# ^ {bookdown} has table/figure cross-ref ability.

### 2022-08-30 src/render_html.R

@@ -1,0 +1,4 @@
+rmarkdown::render("Manuscript_3_draft1.Rmd",
+                  output_file="doc/ms.html",
+                  bookdown::html_document2(template = stevetemplates::templ_html_article(),
+                                           toc = FALSE, number_sections = FALSE))

### 2022-08-30 src/render_pdf-anon.R

@@ -1,0 +1,14 @@
+# rmarkdown::render("ms.Rmd", output_file="doc/ms-anon.pdf",
+#                   params=list(anonymous=TRUE,doublespacing=TRUE,removetitleabstract=TRUE),
+#                   rmarkdown::pdf_document(template = stevetemplates::templ_article2(),
+#                                           latex_engine = "xelatex", dev= "cairo_pdf"))
+# ^ old, legacy, here if you want to revert to it.
+
+
+rmarkdown::render("Manuscript_3_draft1.Rmd", output_file="doc/ms-anon.pdf",
+                  params=list(anonymous=TRUE,doublespacing=TRUE,removetitleabstract=TRUE),
+                  bookdown::pdf_document2(template = stevetemplates::templ_article2(),
+                                          latex_engine = "xelatex", dev= "cairo_pdf",
+                                          toc = FALSE, number_sections = FALSE))
+
+# ^ {bookdown} has table/figure cross-ref ability.

### 2022-08-30 src/render_pdf.R

@@ -1,0 +1,14 @@
+# rmarkdown::render("Manuscript_3_draft1.Rmd",
+#                   output_file="doc/ms.pdf",
+#                   rmarkdown::pdf_document(template = stevetemplates::templ_article2(),
+#                                           latex_engine = "xelatex", dev="cairo_pdf"))
+# ^ old, legacy, here if you want to revert to it.
+
+
+rmarkdown::render("Manuscript_3_draft1.Rmd",
+                  output_file="doc/ms.pdf",
+                  bookdown::pdf_document2(template = stevetemplates::templ_article2(),
+                                          latex_engine = "xelatex", dev="cairo_pdf",
+                                          toc = FALSE, number_sections = FALSE))
+
+# ^ {bookdown} has table/figure cross-ref ability.

### 2022-09-02 Manuscript_3_draft1.Rmd
Extensive revisions to Markdown manuscript, both in terms of code and content.

@@
 However, we still know little about whether exposure to past differentiation shapes public preferences for a more differentiated EU, and whether this varies according to how people think about EU membership.
-Combining recent survey data with data showing country-level exposure to differentiated integration, I show that those who have been exposed to more of it are likelier to want a future EU that allows countries to opt out of undesirable policy integration.
+Combining recent survey data with data showing country-level exposure to differentiated integration, I show that those who have been exposed to more differentiation are likelier to want a future EU that allows countries to opt out of undesirable policy integration.
 This effect is particularly strong among Eurosceptics.
@@
 Second, through a process known as “instrumental differentiation”, countries with a diminished capacity for integration have been allowed to integrate at different schedules.
-This can happen either voluntarily or as a result of EU institutions imposing it as a precondition for membership [@schimmelfennig_instrumental_2014].
+This can either happen voluntarily or as a result of EU institutions imposing it as a precondition for membership [@schimmelfennig_instrumental_2014].
 The differentiation of current European integration makes ‘variable geography’ [@stubb_categorization_1996] a better description of today’s EU than ‘ever closer Union’.
 
-While increasing attention is being paid to why people want a differentiated EU [@leuffen_public_2022; @telle_elite_2022], we know little about how exposure to it has impacted support for a more differentiated future EU.
+While increasing attention is being paid to why people want a differentiated EU [@leuffen_public_2022; @telle_elite_2022], we know little about how exposure to it has impacted support for a differentiated future EU.
 While one of the few existing contributions examining the effect of differentiation on public opinion shows that it may strengthen the perceptions of the EU’s input legitimacy [@schraff_does_2020], two knowledge gaps remain: We still know too little about how past exposure to differentiation shapes support for it.
 Second, our understanding of how the relationship between exposure to differentiation and support for it is conditioned by attitudes towards the EU is still incomplete.
-Whether there is such a conditioning effect, either in the broader EU population, or where differentiation has been implemented in either a dominating or voluntary fashion is also unknown.
+Whether there is a conditioning effect, either in the broader EU population, or where differentiation has been implemented in either a dominating or voluntary fashion is also unknown.
 
 By combining survey data on citizens’ attitudes towards constitutional differentiated integration from 2020-2021 with data detailing differentiations from the EU’s treaties, I find that support for constitutional differentiation increases with exposure to differentiations from EU law, and that the extent to which this is the case strongly depends on your past attitudes towards the EU.
-However, I find that both where differentiation has been voluntary chosen, as in Scandinavia, or externally imposed in a discriminatory fashion, as in Central and Eastern Europe.
-While the reason for the generally greater support for differentiated integration among Eurosceptics in the countries that have been exposed to it may be that Eurosceptics are more likely than others to believe that differentiated integration is a path towards a less integrated Europe, more research is needed to understand why there is no real effect of pre-existing attitudes towards the EU in two regions with very different experiences of differentiation.
+However, I find that both where differentiation has been voluntary chosen, as in Scandinavia, or externally imposed by the EU's institutions, as in Central and Eastern Europe, support for differentiated integration is higher than in the rest of the EU.
+While Eurosceptics may be more likely to support differentiated integration if they have been exposed to greater amounts of it because it would make them more likely than others to believe that differentiated integration is a path towards a less integrated Europe, more research is needed to understand why there is no real effect of past histories of differentiation in two regions with very different historical experiences with differentiated integration.
 
 My findings are important to the discussion about the EU’s future spawned by Brexit [@european_commission_white_2017]: Because EU integration has expanded into increasingly salient areas, differentiation may be called upon to dampen public contestation of it [@schimmelfennig_ever_2020].
-By providing greater numbers of citizens with exposure to differentiation, the EU’s institutions may solidify opposition to further integration among the Eurosceptics that have on several instances been able to put the brakes on further integration [@hooghe_postfunctionalist_2009].
+By providing greater numbers of citizens with exposure to differentiation, the EU’s institutions may solidify opposition to further integration among the same Eurosceptics that have previously been able to put the brakes on further integration [@hooghe_postfunctionalist_2009].
 
@@
 The first is utilitarian, with support for the EU contingent on whether individuals or social groups with which they identify stand to gain from integration [@foster_economic_2021; @gabel_public_1998].
-The second is cue-based, and posits that citizens use heuristics such as elite discourses, media messages, and national benchmarks to form opinions about European integration  [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007; @vries_euroscepticism_2018].
+The second is cue-based, and posits that citizens use informational shortcuts like elite discourses, media messages, and national benchmarks to form an opinion about European integration  [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007; @vries_euroscepticism_2018].
 The literature also shows that strong identification with the nation-state makes citizens more likely to oppose EU integration [@borzel_identity_2020; @hooghe_calculation_2005; @luedtke_european_2005].
 
-An under-investigated issue is why some people support a more differentiated EU.
-This gap is particularly pronounced in light of the large that literature already investigates differentiated integration conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004].
+An under-investigated issue is why some people support a more differentiated EU, in which the EU institutionally permits the divergence of the EU's legal order.
+Such differentiated integration is already comprehensively investigated conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004].
 The few contributions investigating individual support for differentiated integration has focused on liberal economic values [@leuffen_public_2022], Euroscepticism [de_blok_blessing_2021], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite_2022].
 
-Thus, with one notable exception, national contextual factors are under-studied in this literature.
-However, these contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; harteveld_europe_2013].
+Thus, with one notable exception, contextual factors have not been brought to the fore in studies of public support for differentiated integration.
+However, national contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; harteveld_europe_2013].
 One contextual source of support for differentiated integration may be the exposure that citizens of EU countries have had to it: A country’s number of differentiations from EU law may bring its level of integration in and out of alignment with what a citizen would prefer.
 If there is a misalignment between the actual and desired levels of integration, citizens may respond by adjusting their stated preference for future differentiated integration [@busemeyer_beyond_2021; @wlezien_public_1995].
-While previous literature investigates the potential feedback loops between EU policy-making and public opinion [@bolstad_dynamics_2015; @de_bruycker_democratically_2020] this paper shifts the focus to both how support for EU differentiation is impacted by exposure to differentiated integration generally and whether this effect is conditioned by attitudes towards European integration.
+However, previous literature has primarily investigated the potential feedback loops between EU policy-making, such as directives and regulations, and public opinion [@bolstad_dynamics_2015; @de_bruycker_democratically_2020].
+This paper shifts the focus to how alterations to the EU's structure, in the form of negotiated or imposed exemptions from the EU's treaties, shapes attitudes towards future preferences for the very same structure.
 
@@
 Responses to differentiated integration may further vary based on whether differentiation was imposed by the EU’s institutions or voluntarily chosen by member state government.
-An important reason for this is that citizens may perceive such differentiation as the EU dominating their countries, by shutting them out of economically beneficial integration, for instance in the field of freedom of movement [@cianciara_differentiated_2014; @schimmelfennig_instrumental_2014].
+An important reason is that citizens may perceive such differentiation as the EU dominating their countries, by shutting them out of economically beneficial integration.
 As @schimmelfennig_eu_2014 shows, the countries of Central and Eastern Europe were among those exposed to the greatest amount of discriminatory, rather than preferential, differentiated integration.
+Among other things, the new Eastern member states were for several years excluded from the mobility rights afforded to Schengen citizens.
 The reason was that existing member states wanted to shield themselves from the economic destabilization believed to follow from the EU accession of countries like Poland, Estonia, and Hungary.
@@
 
-It is furthermore likely that Eurosceptics in these countries will have a more adverse reaction to differentiated integration than those who support EU membership: As differentiated integration in these countries has been predominantly discriminatory, they may be more likely to identify a non-uniformly integrated EU, in which countries can choose to opt out of integration if they so choose, with the potential for domination.
+Eurosceptics in these countries may also have a more adverse reaction to differentiated integration than those who support EU membership: As differentiated integration in these countries has been predominantly discriminatory, they may be more likely to identify a non-uniformly integrated EU, in which countries can choose to opt out of integration if they so choose, with the potential for domination.
 Given the importance this group places on autonomy they are more likely to have adverse reactions to the discriminatory imposition of differentiated integration.
@@
 Elites in Central and Eastern Europe, especially in Poland, have also framed differentiated integration as a path towards a divided EU, with a powerful core and a less powerful periphery [@cianciara_differentiated_2014 178-179].
-For example, Polish politicians opposed further integration of the Eurozone, a prominent example of differentiated integration, because they feared that the Central and Eastern member states, as non-members, would lose their influence over economic policy-making by not being inside the EU’s economic policy-making structures.
-Because elite cues are particularly important for shaping attitudes towards less salient issues, such as differentiated integration [@steenbergen_whos_2007; @telle_elite_2022], a reasonable assumption is that citizens would respond to these discourses by adjusting their preferences for integration downwards.
+For example, Polish politicians opposed further integration of the Eurozone, a prominent example of differentiated integration, because they feared that the Central and Eastern member states outside the currency area would lose their influence over the EU's economic policy-making.
+Because elite cues are particularly important for shaping attitudes towards less salient issues, such as differentiated integration, [@steenbergen_whos_2007; @telle_elite_2022] citizens could be more likely to respond to these discourses by adjusting their preferences for integration downwards than if they were related to more salient policy outcomes.
 Thus, two different causal mechanisms could produce the same outcome.
 
-However, this history might be less relevant if it not being used by current actors.
+However, this history might be less relevant if it not being discursively mobilized by current actors.
 Instead, the increasingly Eurosceptic elite attitudes in Central and Eastern Europe [@borzel_identity_2020; @brack_towards_2020; @kriesi_politicization_2016] might form a more potent backdrop against which Central and Eastern European citizens might be expected to make up their minds.
@@
 
-```{r h2a, eval=!is_latex, results="asis"}
-cat('> **H2a**: Exposure to differentiated integration in Central and Eastern will lead to lower levels of support for future differentiated integration than in the rest of the EU.')
-```
-
 *H2b: Exposure to differentiated integration in Central and Eastern Europe will lead to lower levels of support for it among Eurosceptics in these countries compared to supporters of EU membership.*  
-
-```{r h2b, eval=!is_latex, results="asis"}
-cat('> **H2b**: Exposure to differentiated integration in Central and Eastern Europe will lead to lower levels of support for it among Eurosceptics in these countries compared to supporters of EU membership.')
-```
-
 *H3: Exposure to differentiated integration in Central and Eastern Europe will lead to higher levels of support for it than in the rest of the EU.*
 
-```{r h3, eval=!is_latex, results="asis"}
-cat('> **H3**: Exposure to differentiated integration in Central and Eastern Europe will lead to higher levels of support for it than in the rest of the EU.')
-```
-
-If exposure to discriminatorily imposed differentiation is assumed to lead to less support for differentiated integration, especially among Eurosceptics, the opposite is likely to be the case for countries with a great deal of voluntarily chosen differentiation.
+If exposure to discriminatory and externally imposed differentiation is assumed to lead to less support for differentiated integration, especially among Eurosceptics, the opposite is likely to be the case for countries where differentiation has primarily been voluntarily chosen.
 Scandinavia offers an interesting case for testing this assumption.
@@
 
-```{r h4a, eval=!is_latex, results="asis"}
-cat('> **H4a**: The positive effect of exposure to differentiated integration will be greater for citizens of Sweden and Denmark than in the EU as a whole')
-```
-
 *H4b: The positive effect of exposure to differentiated integration will be greater for Scandinavian Eurosceptics than supporters of EU membership.*
-
-```{r h4b, eval=!is_latex, results="asis"}
-cat('> **H4b**: The positive effect of exposure to differentiated integration will be greater for Scandinavian Eurosceptics than supporters of EU membership.')
-```
 

### 2022-09-05 Manuscript_3_draft1.Rmd
Uploaded appendix Rmd file and revised main manuscript.

@@
 csl: inst/european-journal-of-political-research.csl
-abstract: "Questions about the public legitimacy of the EU’s increasingly flexible integration are attracting increasing scholarly interest.
-However, we still know little about whether exposure to past differentiation shapes public preferences for a more differentiated EU, and whether this varies according to how people think about EU membership.
+abstract: "Questions about the public legitimacy of the EU’s flexible integration is emerging as an important question for European public opinion research.
+However, we still know very little about how attitudes towards the EU's differentiated integration are shaped by exposure to such differentiation in the past, and how this effect relates to individual Euroscepticism.
 Combining recent survey data with data showing country-level exposure to differentiated integration, I show that those who have been exposed to more differentiation are likelier to want a future EU that allows countries to opt out of undesirable policy integration.
 This effect is particularly strong among Eurosceptics.
-My results suggest that exposure to the differentiated integration often used to integrate highly salient policy areas can solidify popular support for a less integrated EU, complicating the EU’s move towards an ‘ever closer Union’."
+However, whether differentiated integration is voluntarily chosen or discriminatorily imposed seems to be less important.
+Altogether, my results suggest that exposure to the differentiated integration often used to integrate highly salient policy areas can solidify popular support for a less integrated EU.
+This potentially complicates the EU’s move towards an ‘ever closer Union’."
 keywords: "differentiated integration, public opinion, Euroscepticism, European integration, contestation of the EU"
@@
 is_latex <- knitr::pandoc_to("latex")
-
 # I don't know how Texas Instruments smart this is, but p-sure default DPI is 96.
@@
 # in {bookdown} while allowing for conditional DPI
-
 if (is_latex) {
@@
 } 
-
 knitr::opts_chunk$set(echo = FALSE,
@@
                       })
-
 library(dplyr)
@@
 library(kableExtra)
-
 allrds <- list.files(pattern = ".rds", path = "data") %>%
   stringr::str_remove(., ".rds")
-
 for(i in allrds) {
@@
     }
-
 ```
@@
 # Introduction
-Despite being among the world’s most advanced political cooperations, the expansion of the EU’s policy remit after the ratification of the Maastricht Treaty has made its integration ever-more differentiated [@leuffen_differentiated_2013; @schimmelfennig_ever_2020].
-Internal EU differentiation typically takes two forms: First, some member states have opted out of cooperation to protect their own sovereignty.
+Despite the EU being one of the most advanced political cooperations at the global stage, its integration has become increasingly fragmented after the ratification of the 1992 Maastricht Treaty.
+The differentiation of EU integration has primarily been driven by a desire to deepen political cooperation despite contestation of it in groups of countries [@leuffen_differentiated_2013; @schimmelfennig_ever_2020].
+Differentiation has typically taken two forms: First, some member states have opted out of cooperation to protect their own sovereignty.
 This is called “constitutional differentiation”.
-Second, through a process known as “instrumental differentiation”, countries with a diminished capacity for integration have been allowed to integrate at different schedules.
-This can either happen voluntarily or as a result of EU institutions imposing it as a precondition for membership [@schimmelfennig_instrumental_2014].
-The differentiation of current European integration makes ‘variable geography’ [@stubb_categorization_1996] a better description of today’s EU than ‘ever closer Union’.
+Second, “instrumental differentiation” has happened where countries with a diminished capacity for integration have been allowed to integrate at different schedules.
+Such differentiation can either happen voluntarily or because the EU's institutions impose differentiation as a precondition for membership [@schimmelfennig_instrumental_2014].
+Thus, ‘variable geography’ [@stubb_categorization_1996] is in many ways a better description of the EU's political development than the union's own goal of "ever closer Union".
 
-While increasing attention is being paid to why people want a differentiated EU [@leuffen_public_2022; @telle_elite_2022], we know little about how exposure to it has impacted support for a differentiated future EU.
-While one of the few existing contributions examining the effect of differentiation on public opinion shows that it may strengthen the perceptions of the EU’s input legitimacy [@schraff_does_2020], two knowledge gaps remain: We still know too little about how past exposure to differentiation shapes support for it.
-Second, our understanding of how the relationship between exposure to differentiation and support for it is conditioned by attitudes towards the EU is still incomplete.
-Whether there is a conditioning effect, either in the broader EU population, or where differentiation has been implemented in either a dominating or voluntary fashion is also unknown.
+While increasing attention is being paid to what shapes popular preferences for such differentiation [@leuffen_public_2022; @telle_elite_2022], we know little about how actual exposure to it has impacted support for the goal of a uniformly integrated EU.
+While one of the few existing contributions examining the effect of differentiation on public opinion shows that allowing for differentiation may strengthen the perceived democratic legitimacy of the EU [@schraff_does_2020], important knowledge gaps remain: First, as mentioned, we know too little about how past exposure to differentiation shapes support for it.
+Second, our understanding of how this relationship is moderated by Eurosceptic sentiments.
+Lastly, we do not know whether the voluntarily chosen or involuntarily imposed nature of past differentiation impacts citizens' preferences for differentiation as a future model of EU integration.
 
-By combining survey data on citizens’ attitudes towards constitutional differentiated integration from 2020-2021 with data detailing differentiations from the EU’s treaties, I find that support for constitutional differentiation increases with exposure to differentiations from EU law, and that the extent to which this is the case strongly depends on your past attitudes towards the EU.
-However, I find that both where differentiation has been voluntary chosen, as in Scandinavia, or externally imposed by the EU's institutions, as in Central and Eastern Europe, support for differentiated integration is higher than in the rest of the EU.
-While Eurosceptics may be more likely to support differentiated integration if they have been exposed to greater amounts of it because it would make them more likely than others to believe that differentiated integration is a path towards a less integrated Europe, more research is needed to understand why there is no real effect of past histories of differentiation in two regions with very different historical experiences with differentiated integration.
+By combining survey data on citizens’ attitudes towards constitutional differentiated integration from 2020-2021 with data detailing country-level differentiations from the EU’s treaties, I find that support for constitutional differentiation increases with exposure to differentiations from EU law, and that the extent to which this is the case strongly depends on individual-level Euroscepticism.
+However, there does not seem to be a clear connection between how the differentiated integration citizens have been exposed to in the past have been implemented and support for a less uniformly integrated EU.
+The greater support for differentiation among Eurosceptics can be explained by how such differentiation is more likely to bring about the less integrated Europe that such individuals may want.
+However, more research is needed to understand the seeming lack of a diverging response to differentiation even where it has been implemented under very different circumstances.
 
-My findings are important to the discussion about the EU’s future spawned by Brexit [@european_commission_white_2017]: Because EU integration has expanded into increasingly salient areas, differentiation may be called upon to dampen public contestation of it [@schimmelfennig_ever_2020].
-By providing greater numbers of citizens with exposure to differentiation, the EU’s institutions may solidify opposition to further integration among the same Eurosceptics that have previously been able to put the brakes on further integration [@hooghe_postfunctionalist_2009].
+In addition to expanding our knowledge of public support for differentiated EU integration, this article contributes to a broader literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009] While the EU’s legal and functional reach is much greater than that of other international organizations, many of the organizations constituting the global governance system have seen their legal and political powers expand in recent years [@zurn_theory_2018].
+Even if more research is needed to understand whether my results travel to contexts other than the EU, my results can help illuminate how the allowance of differentiations from an emerging international legal order can shape attitudes towards deepening transnational cooperation.
 
-Furthermore, the article contributes to the broader emerging literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009] While the EU’s legal and functional reach is much greater than that of other international organizations, there has been a turn towards expanding both the political and legal powers of many of the organizations constituting the global governance system [@zurn_theory_2018].
-Even if more research is needed to understand whether my results travel to contexts other than the EU, my article helps shine a light on how public support for deeper political integration is impacted by non-uniform implementations of this emerging legal order.
+My findings are also important to the discussion about the EU’s future spawned by Brexit [@european_commission_white_2017]: Because EU integration has expanded into increasingly salient areas, differentiation has often been used to mitigate the challenges caused by public contestation [@schimmelfennig_ever_2020].
+By providing greater numbers of citizens with exposure to differentiation, the EU’s institutions may solidify opposition to further integration among greater numbers of citizens, potentially stalling future attempts at cooperation [@hooghe_postfunctionalist_2009].
 
@@
 The literature identifies three mechanisms that help shape attitudes towards the EU.
-The first is utilitarian, with support for the EU contingent on whether individuals or social groups with which they identify stand to gain from integration [@foster_economic_2021; @gabel_public_1998].
-The second is cue-based, and posits that citizens use informational shortcuts like elite discourses, media messages, and national benchmarks to form an opinion about European integration  [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007; @vries_euroscepticism_2018].
-The literature also shows that strong identification with the nation-state makes citizens more likely to oppose EU integration [@borzel_identity_2020; @hooghe_calculation_2005; @luedtke_european_2005].
+The first is utilitarian, with support for the EU contingent on whether individuals or the social groups with which they identify stand to gain from integration [@foster_economic_2021; @gabel_public_1998].
+The second is cue-based, with citizens relying informational shortcuts like elite discourses, media messages, and national benchmarks to form opinions about European integration [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007; @vries_euroscepticism_2018].
+Lastly, strong identification with the nation-state potentially makes citizens more likely to oppose EU integration [@borzel_identity_2020; @hooghe_calculation_2005; @luedtke_european_2005].
 
@@
 
-Thus, with one notable exception, contextual factors have not been brought to the fore in studies of public support for differentiated integration.
-However, national contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; harteveld_europe_2013].
-One contextual source of support for differentiated integration may be the exposure that citizens of EU countries have had to it: A country’s number of differentiations from EU law may bring its level of integration in and out of alignment with what a citizen would prefer.
-If there is a misalignment between the actual and desired levels of integration, citizens may respond by adjusting their stated preference for future differentiated integration [@busemeyer_beyond_2021; @wlezien_public_1995].
-However, previous literature has primarily investigated the potential feedback loops between EU policy-making, such as directives and regulations, and public opinion [@bolstad_dynamics_2015; @de_bruycker_democratically_2020].
-This paper shifts the focus to how alterations to the EU's structure, in the form of negotiated or imposed exemptions from the EU's treaties, shapes attitudes towards future preferences for the very same structure.
+Thus, with one notable exception, the effect of contextual factors are under-studied.
+Such national contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; harteveld_europe_2013].
+One potential source of support for differentiated integration may be the varying exposure that citizens of EU countries have had to it: A country’s number of differentiations from EU law may bring its level of integration in and out of alignment with what a citizen would prefer.
+If there is a misalignment between the actual and desired levels of integration, citizens may respond by either expressing greater or lesser support for a more differentiated EU that can bring about a political cooperation more aligned with their preferences [@busemeyer_beyond_2021; @wlezien_public_1995].
+Previous literature has primarily investigated the potential feedback loops between EU policy-making, such as directives and regulations, and public opinion [@bolstad_dynamics_2015; @de_bruycker_democratically_2020].
+This paper instead focuses on how alterations to the EU's legal structure, in the form of negotiated or imposed exemptions from the EU's treaties, shapes attitudes towards more or less uniform integration.
 
-I believe that greater exposure to differentiated integration will increase support for it as a model of future integration.
-This is because such exposure is likely to familiarize citizens with EU differentiation, thus increasing their knowledge of how differentiated integration would work in the future.
-Greater knowledge is often, though not always, associated with greater support for European integration [@marquart_knowing_2019; @verhaegen_does_2015].
-I furthermore find it theoretically plausible that whether this is the case will depend on whether one identifies with Eurosceptic sentiments or not: Because constitutional differentiated integration allows for permanent opt-outs from specific policy areas, Eurosceptics are more likely to see differentiated integration as a viable path towards the less differentiated Europe that they want to see in the future.
+Greater exposure to differentiated integration is also likely to make citizens more familiar with EU differentiation, thus increasing their knowledge of how differentiated integration would work in the future.
+Such knowledge of EU policies is often, though not always, associated with greater support for EU integration [@marquart_knowing_2019; @verhaegen_does_2015].
+I also find it theoretically plausible that this relationship will depend on whether citizens support their countries' EU memberships or not: Because constitutional differentiated integration allows for permanent opt-outs from specific policy areas, Eurosceptics who favour leaving the EU are more likely to see differentiated integration as a viable path towards the less differentiated Europe that they primarily prefer.
 This yields two complementary hypotheses:
@@
 
-```{r h1a, eval=!is_latex, results="asis"}
-cat('> **H1a**: Exposure to differentiated integration increases support for future differentiated integration. ')
-```
+*H1b: Exposure will lead to greater support for differentiation among Eurosceptics than supporters of EU membership.*
 
-*H1b: The effect of exposure is larger among Eurosceptics than supporters of EU membership.*
-
-```{r h1b, eval=!is_latex, results="asis"}
-cat('> **H1b**: The effect of exposure is larger among Eurosceptics than supporters of EU membership.')
-```
-
-Responses to differentiated integration may further vary based on whether differentiation was imposed by the EU’s institutions or voluntarily chosen by member state government.
+Responses to differentiated integration may also vary based on whether differentiation was imposed by the EU’s institutions or voluntarily chosen by member state governments.
 An important reason is that citizens may perceive such differentiation as the EU dominating their countries, by shutting them out of economically beneficial integration.
-As @schimmelfennig_eu_2014 shows, the countries of Central and Eastern Europe were among those exposed to the greatest amount of discriminatory, rather than preferential, differentiated integration.
-Among other things, the new Eastern member states were for several years excluded from the mobility rights afforded to Schengen citizens.
+As @schimmelfennig_eu_2014 shows, countries in Central and Eastern Europe were among those exposed to the greatest amount of discriminatory differentiated integration.
+Among other things, the new Eastern member states were for several years excluded from Schengen, and thus the possibility of easily moving for work purposes.
 The reason was that existing member states wanted to shield themselves from the economic destabilization believed to follow from the EU accession of countries like Poland, Estonia, and Hungary.
@@
 
-Eurosceptics in these countries may also have a more adverse reaction to differentiated integration than those who support EU membership: As differentiated integration in these countries has been predominantly discriminatory, they may be more likely to identify a non-uniformly integrated EU, in which countries can choose to opt out of integration if they so choose, with the potential for domination.
-Given the importance this group places on autonomy they are more likely to have adverse reactions to the discriminatory imposition of differentiated integration.
+Eurosceptics in these countries may also have more adverse reactions to differentiated integration than those supporting EU membership: Because citizens in these countries may predominantly be familiar with externally imposed and discriminatory differentiation, Eurosceptics may be more likely to identify a non-uniformly integrated EU that allows countries to opt out of integration with the potential for domination.
+Their concern for national sovereignty may again lead them to develop more strongly negative attitudes towards differentiated integration than what can be found among supporters of EU membership.
 
@@
 For example, Polish politicians opposed further integration of the Eurozone, a prominent example of differentiated integration, because they feared that the Central and Eastern member states outside the currency area would lose their influence over the EU's economic policy-making.
-Because elite cues are particularly important for shaping attitudes towards less salient issues, such as differentiated integration, [@steenbergen_whos_2007; @telle_elite_2022] citizens could be more likely to respond to these discourses by adjusting their preferences for integration downwards than if they were related to more salient policy outcomes.
+As elite cues are particularly important for shaping attitudes towards less salient EU questions [@steenbergen_whos_2007], such as differentiated integration [@telle_elite_2022], citizens would be likely to respond to these discourses by expressing less support for differentiated integration.
 Thus, two different causal mechanisms could produce the same outcome.
 
-However, this history might be less relevant if it not being discursively mobilized by current actors.
-Instead, the increasingly Eurosceptic elite attitudes in Central and Eastern Europe [@borzel_identity_2020; @brack_towards_2020; @kriesi_politicization_2016] might form a more potent backdrop against which Central and Eastern European citizens might be expected to make up their minds.
+However, this history might be less relevant if current political elites do not use it to shape preferences towards differentiated.
+Instead, the increasingly Eurosceptic attitudes expressed by Central and Eastern European politicians [@borzel_identity_2020; @brack_towards_2020; @kriesi_politicization_2016] might form a more potent backdrop against which Central and Eastern European citizens might be expected to make up their minds.
 Exposure to such cues might, in contrast to what was stated before, make citizens more likely to express an increased demand for differentiated integration.
@@
 *H2a: Exposure to differentiated integration will lead to lower levels of support for future differentiated integration in Central and Eastern Europe than in the rest of the EU.*    
-
 *H2b: Exposure to differentiated integration in Central and Eastern Europe will lead to lower levels of support for it among Eurosceptics in these countries compared to supporters of EU membership.*
@@
 
-If exposure to discriminatory and externally imposed differentiation is assumed to lead to less support for differentiated integration, especially among Eurosceptics, the opposite is likely to be the case for countries where differentiation has primarily been voluntarily chosen.
-Scandinavia offers an interesting case for testing this assumption.
-Here the mode of implementation has been drastically different from that found in Central and Eastern Europe, as differentiation was both desired by national governments and justified through the autonomy-enhancing aspects of differentiated integration [@adler-nissen_opting_2014; @leruth_operationalizing_2015; @spendzharova_banking_2016].
-Because of the previously explained cueing mechanism one might expect this pro-autonomy discourse to produce greater support for differentiated integration in Scandinavia than elsewhere in the EU.
+If exposure to discriminatory and externally imposed differentiation leads to less support for differentiated integration, especially among Eurosceptics, we are theoretically likely to find the opposite where differentiation has been voluntarily chosen.
+Scandinavia is an interesting case study of this theoretical assumption: Here differentiation has been chosen by national governments and justified by explicit references to how these differentiations enhanced national autonomy [@adler-nissen_opting_2014; @leruth_operationalizing_2015; @spendzharova_banking_2016].
+Previous literature has also found that the renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy among Danish Eurosceptics [@schraff_does_2020].
+In other words, it is likely that the familiarity that citizens in Scandinavia have with voluntarily chosen differentiation makes could make them more likely to support it as a future model of EU integration.
 
-Previous literature has found that the renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy among Danish Eurosceptics [@schraff_does_2020].
-This suggests both that citizens may respond positively to differentiation and that past exposure could make them more favourably inclined to it in the future.
-This is particularly likely to be true for Scandinavian Eurosceptics: These citizens may, because their countries are both relatively small and highly dependent on European market access, be more likely to see differentiation as a viable second-best alternative to the European disintegration they would otherwise favour.
-More generally it may also be the case that, because they are sceptical of EU integration, they are more likely to see differentiated integration as a goal in itself as it represents the kind of less integrated Europe that they would be likely to favour.
+This is also particularly likely to be true for Scandinavian Eurosceptics: Because their countries are both relatively small and highly dependent on European market access, Eurosceptic Scandinavians could be more likely to see differentiation as a viable second-best alternative to the European disintegration they would otherwise favour.
+It may also be the case that they, because of a generally critical attitude towards EU integration, are more likely to see differentiated integration as a solution that would better facilitate the less integrated Europe they otherwise favour.
 I hypothesize the following:
 
-*H4a: The positive effect of exposure to differentiated integration will be greater for citizens of Sweden and Denmark than in the EU as a whole.*
+*H4a: Exposure to differentiated integration will lead to greater levels of support for it among for citizens of Sweden and Denmark than in the EU as a whole.*
 
-*H4b: The positive effect of exposure to differentiated integration will be greater for Scandinavian Eurosceptics than supporters of EU membership.*
+*H4b: Exposure to differentiated integration will lead to greater support for it among Scandinavian Eurosceptics than supporters of EU membership.*
 
@@
 I then test whether this relationship is moderated by pre-existing Eurosceptic sentiments.
-Such a moderating effect is particularly likely to exist in either of the two regions because of the way that voluntary or discriminatory differentiation may be perceived as either an expansion or diminishing of the autonomy that this group is particularly likely to value.
+This lets me test both the general effect of exposure to differentiation as well as its interaction with pre-existing attitudes to European integration.
 
@@
 nordic_data <- readRDS("data/nordic_data.Rds")
-
 ```
 
-I investigate these hypotheses using multilevel linear models and data from Yougov surveys fielded in 2020-21 [@hemerijck_eui-yougov_2021].
+I investigate support for differentiated integration using multilevel linear models and data from Yougov surveys fielded in 2020-21 [@hemerijck_eui-yougov_2021].
 The sample includes respondents from Denmark, Finland, France, Germany, Sweden, Greece, Hungary, Italy, Lithuania, the Netherlands, Poland, Romania, and Spain and has an *N* of `r nrow(EUI_final_DI_disc)`.
@@
 This is called constitutional differentiation.
-The second asks whether the EU should allow countries to integrate at different speeds, while maintaining the goal of uniform integration in the long run.
+The second asks whether the EU should allow countries to integrate at different speeds, while maintaining the goal of uniform integration.
 This is an instrumentally differentiated EU.
-This article focuses on popular support for the former, as its allowance for an EU that is permanently segmented into multiple policy communities represents arguably the sharpest departure from the EU’s normative goal of an “ever closer Union”.
+This article focuses on popular support for constitutional differentiation, as its allowance for an EU that is permanently segmented into multiple policy communities represents arguably the sharpest departure from the EU’s normative goal of an “ever closer Union”.
 
-I combine these data with a novel dataset [@schimmelfennig_eudiff1_2022] that features all differentiations from the EU treaties between 1952-2020.
-The data shows each instance of differentiation a country has been exposed to, the number of articles involved, and how long it was enforced.
-The dataset contains data on differentiations from EU treaties, or primary law, rather than differentiations from the less salient secondary law that constitutes the EU’s main legislative output [@duttle_opting_2017].
-I use treaty differentiations because these are typically more publicly salient and subject to greater debate.
-Citizens are thus more likely to be aware of them than differentiations in secondary law.
+The survey data are combined with a novel dataset [@schimmelfennig_eudiff1_2022] that features all differentiations from the EU treaties between 1952-2019.
+The data shows each differentiation a country has been exposed to, the number of articles it involved, and how long it was enforced.
+The data contain information about differentiations from EU treaties, or primary law, rather than differentiations from the less salient secondary law that constitutes the EU’s main legislative output [@duttle_opting_2017].
+I believe the greater salience of primary law differentiation in public debates will lead to a greater knowledge of them among EU citizens than what is the case for differentiations from secondary law.
+They are thus more likely to be important for respondents as they make up their minds about the desirability of differentiated integration.
 
@@
 The variable uses a five-unit Likert scale and asks respondents whether they support a constitutionally differentiated EU (“Please tell us how far you agree or disagree with the following statement: Member states should be allowed to opt out of specific areas of European integration.
-This means that a member state can negotiate exceptions ("opt-out") for areas in which it does not wish to
-cooperate. For example, Denmark has opted out of the common currency, and Poland has
-opted out of the EU's Charter of Fundamental Rights.)".
-The response categories are ordered from 1-5, with 1 being Strongly agree and 5 being strongly disagree.
-I reverse the variable coding, so that higher values indicate agreement.
+This means that a member state can negotiate exceptions ("opt-out") for areas in which it does not wish to cooperate.
+For example, Denmark has opted out of the common currency, and Poland has opted out of the EU's Charter of Fundamental Rights.)".
+The response categories are recoded, so that 1 becomes Strongly disagree and 5 becomes Strongly agree.
 By not asking about “constitutional differentiated integration”, the question becomes easier to understand.
@@
 To operationalize each country’s exposure to EU differentiation I multiply, for each differentiation, the number of articles that a country was exempt from with the number of years the differentiation was in force.
-This is then weighted based on the length of time that has passed since the differentiation ceased to be enforced, and finally log-transformed.
+This is then weighted based on the length of time that has passed since the differentiation ceased to be enforced.
+I lastly log-transform the variable.
 
-Operationalizing exposure to differentiation in this way has three main benefits: First, it captures both the temporal and functional scope of a country’s exposure to differentiation.
-Second, the weighting scheme means that those instances of differentiation that are still in force are given the most weight when investigating differentiation’s impact on support for differentiated integration in the future.
-This is likely to correspond to how respondents relate their own experience of differentiation to support for it: While many respondents from Central and Eastern Europe may still remember the particulars of their countries’ exposure to externally imposed differentiation, possibly producing a cohort effect among these individuals [@lauterbach_europe_2020], Southern European respondents are less likely to remember that they were exposed to externally imposed differentiation several decades ago and use this as an informational backdrop for their attitude-formation.
-Finally, the log-transformation reduces the influence of outliers.
+This operationalization has three benefits: First, it captures both the temporal and functional scope of a country’s exposure to differentiation.
+Second, the weighting scheme means that those instances of differentiation that were recently in force are given the most weight when investigating the impact of exposure to differentiated integration.
+This is also likely to correspond to how respondents relate their own experience of differentiation to support for it: While many respondents from Central and Eastern Europe may still remember the particulars of their countries’ exposure to externally imposed differentiation, possibly producing a cohort effect [@lauterbach_europe_2020], Southern European respondents are less likely to both remember that they were exposed to externally imposed differentiation several decades ago and use this as a backdrop against which they shape their attitudes towards it.
+Finally, the log-transformation reduces the influence of extreme values, caused by particularly large numbers of differentiations in certain countries.
 
@@
 To create a Central and Eastern European dummy I do the same for everyone from Hungary, Lithuania, Poland, and Romania.
-These are used to investigate the interaction effects assumed by H2-H4.
+These are used to investigate the interaction effects assumed by *H2-H4*.
 I also create a dummy for Euroscepticism, coding those who answer that they would vote yes to leaving the European Union in a hypothetical referendum with a 1 and those answering no with a 0.
@@
 data_descriptives <- readRDS("data/Data_descriptives.Rds")
-
 data_descriptives <- data_descriptives %>%
@@
 library(modelsummary)
-
 table_format <- ifelse(is_docx, "huxtable", 'default')
-
 modelsummary::datasummary_skim(data_descriptives, fmt = 3, histogram = FALSE, 
                  title = "Descriptive statistics", output = "latex")
-
 ```
@@
 ## Limitations
-A limitation of the study is that both surveys were fielded during the COVID-19 pandemic, which could impact support for differentiation.
+A limitation of the study is that both surveys were fielded during the COVID-19 pandemic.
+This might impact support for differentiation.
 However, @cicchi_eu_2020 show that solidarity, as is common, still seemed to reside at the national level in the early phase of the crisis.
 This means that my results may generalize to periods beyond those in which the survey was fielded.
-To account for how different baseline levels of support for differentiation are likely to exist in each country, I use multilevel models with random country intercepts.
+To account for the different baseline levels of support that are likely to exist in each country, I use multilevel models with random country constants
 
@@
 A further limitation, general to all studies using observational survey data, is the challenge of omitted variable bias.
-In this study omitted variable bias is particularly likely to stem from the absence of data on the respondents’ knowledge of the EU and education [@karp_know_2003; @marquart_knowing_2019].
-I use sensitivity analysis [@cinelli_making_2020] to tentatively quantify the size of this bias and show the results of the analyses in the section Robustness checks.
+Here omitted variable bias is particularly likely to stem from the absence of data on the respondents’ knowledge of the EU and education [@karp_know_2003; @marquart_knowing_2019].
+I use sensitivity analysis [@cinelli_making_2020] to tentatively quantify the size of this bias.
+I show the results of these analyses in the section Robustness checks.
 
 # Results and discussion
-I first show the effect of exposure to differentiation in the broader population, and especially how this interacts with already pre-existing Euroscepticism.
-I then discuss the effect of differentiation on support for it in both Scandinavia and Central and Eastern Europe.
+I first show the effect of exposure to differentiation in the broader population, and especially how this interacts with pre-existing Euroscepticism.
+I then discuss how the effect of exposure is moderated by Scandinavian or Central and Eastern European citizenship.
 
@@
 library(merTools)
-
 ```
@@
 ```{r }
-
 main_mod = list()
-
 main_mod$"Euroscepticism" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
@@
                                                          data = EUI_final_DI_disc)
-
 coef_map_eurosceptic <- c('(Intercept)' = 'Constant',
@@
                           'excl_identity' = 'Exclusively national identity')
-
 modelsummary(main_mod, fmt = 3, coef_map = coef_map_eurosceptic, output = "latex", 
@@
 Both table 2 and figure 1 strongly support *H1a* and *H1b*, by confirming that exposure to differentiated integration is associated with greater support for differentiation as a future model for the EU.
-The effect of exposure is also significantly greater among Eurosceptics.
+We also find that exposure leads to greater levels of support for differentiated integration among Eurosceptics.
 The reason may be that Eurosceptic citizens in countries with greater exposure to differentiated integration are likelier to be familiar with the structure of a differentiated EU membership than those in countries not exposed to any differentiation.
-They may also see it as a more viable “second-best” alternative to full disintegration.
+They may thus be the ones most likely to see such differentiation as a viable solution for achieving the goal of a less integrated EU.
 However, even if the effect is larger for Eurosceptic than non-Eurosceptic citizens, the results suggest that future demands for differentiated integration among those most strongly exposed to it are unlikely to be restricted only to Eurosceptics.
@@
 ```{r, fig.pos="H", fig.cap="Predicted probability of support for constitutional differentiation as a function of the interaction between article-years (logged) and Euroscepticism. 95% CIs."}
-
 library(sjPlot)
@@
                                     colors = "bw", legend.title = "Eurosceptic")
-
 pred_plot_eurosceptic_final <-  pred_plot_eurosceptic + ggtitle("Support for constitutional differentiation", 
@@
                            labels = c("Non-Eurosceptic", "Eurosceptic"))
-
 pred_plot_eurosceptic_final <- pred_plot_eurosceptic_final + font_size(title = 15, labels.x = 15, labels.y = 15)
-
 pred_plot_eurosceptic_final
@@
 
-A plausible counter-argument to my hypothesized causal mechanism could be that differentiation was primarily chosen where underlying Euroscepticism, and thus a demand for differentiation, was already high.
-Figure A6 offers some support for such an assumption, as we find high levels of differentiated integration in Sweden and Denmark.
+A plausible counter-argument to my hypothesized causal mechanism could be that differentiation has primarily been chosen where underlying Euroscepticism, and potentially the demand for differentiation, was already high.
+Figure A6 offers some support for this assumption, as we find high levels of differentiated integration in Sweden and Denmark.
 These are countries where opt-outs from integration have been deployed as a response to a Euroscepticism shared by both political parties and voters [@leruth_operationalizing_2015; @spendzharova_banking_2016].
-However, countries as diverse as Lithuania, Hungary, Italy, and the Netherlands have also seen differentiations from EU treaties.
-Differentiated integration is thus not a phenomenon restricted to the countries found in what @bolstad_dynamics_2015 identifies as a Eurosceptic periphery.
+However, citizens of countries as diverse as Lithuania, Hungary, Italy, and the Netherlands have also been exposed to differentiated integration.
+It is thus not a phenomenon restricted to the countries found in what @bolstad_dynamics_2015 identifies as a Eurosceptic periphery.
 This strengthens the case that the effect is not only driven by an endogenous relationship between Euroscepticism and differentiation.
@@
 ```{r }
-
 main_regional = list()
-
 main_regional$"Regional dummies" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
@@
                                                weights = EUI_final_DI_disc$weight, data = EUI_final_DI_disc)
-
 library(modelsummary)
@@
                       "article_disc_logeurosceptic1" = "Article-years (logged) X Eurosceptic")
-
 modelsummary(main_regional, fmt = 3, output = "latex", coef_map = coef_map_regions, 
@@
 
-Table 3 shows a positive and statistically significant effect of the interaction between exposure to differentiated integration and Scandinavian citizenship.
-However, it finds no statistically significant effect of the interaction between Central and Eastern European citizenship and exposure to differentiated integration.
+Table 3 shows that the effect of the interaction between exposure to differentiated integration and Scandinavian citizenship is positive and statistically significant.
+However, the interaction effect between exposure to differentiated integration and Central and Eastern European citizenship is statistically insignificant.
 
@@
                                 legend.title = "Scandinavia", colors = "bw")
-
 pred_plot_nordics_final <-  pred_plot_nordics + ggtitle("Scandinavia") + labs(x = "Differentiation (article-years (logged)",
@@
                            labels = c("Non-Scandinavian", "Scandinavian"))
-
 ##CEE
@@
                                     colors = "bw", legend.title = "CEE")
-
 pred_plot_CEE_final <-  pred_plot_CEE + ggtitle("Central and Eastern Europe") + labs(x = "Differentiation (article-years (logged))",
@@
                            labels = c("Non-CEE", "CEE"))
-
 pred_plot_CEE_final <- pred_plot_CEE_final + font_size(title = 15, labels.x = 15, labels.y = 15)
@@
 library(patchwork)
-
 regional_final <- pred_plot_nordics_final/pred_plot_CEE_final + plot_annotation(tag_levels = "A", title = "Support for constitutional differentiation", subtitle = "Predicted values for regional interactions of exposure to differentiation in Scandinavia and Central and Eastern Europe", theme = theme(plot.title = element_text(size = 16)))
-
 regional_final
@@
 
-Figure 2 supports this picture: Scandinavian citizens appear to become more favourably inclined towards differentiated integration as their exposure to it increases, compared to those from other regions.
-However, this effect is mainly present at the lower end of exposure to differentiation.
-Thus,I am able to confirm *H4a*, as distinct differences do exist within certain ranges of the variable.
-However, because of the large overlap for the largest range of the exposure variable this statistical significance must not be taken as indicative of a substantively large interaction effect.
+Figure 2 supports this picture: Scandinavian citizens appear to become more favourably inclined towards differentiated integration as their exposure to it increases compared to those from other regions.
+However, this effect is only present for a certain range of the measure of exposure.
+Thus, I can confirm *H4a*.
+However, for a large range of the exposure variable the difference between Scandinavian and non-Scandinavian citizens is statistically indistinguishable from zero.
 
-Similarly, the predicted values show no significant difference between support for differentiation in Central and Eastern Europe and the rest of the EU.
+However, the predicted values show no significant difference between support for differentiation in Central and Eastern Europe and the rest of the EU.
 This leads me to reject *H2a*.
-I do, however, find some support for *H3*, whose underlying assumption is that exposure to Eurosceptic cues can lead to stronger support for differentiation even if it was imposed in a discriminatory fashion.
+Neither do I find support for *H3*, as there is no statistically significant greater positive effect of exposure in Central and Eastern Europe.
 
@@
 region_mod <- readRDS("data/region_mod.Rds")
-
 modelsummary::modelsummary(region_mod, fmt = 3, output = "latex", coef_map = coef_map_eurosceptic, 
@@
                                 legend.title = "Eurosceptic", colors = "bw")
-
 pred_plot_nordics_eurosceptic_final <-  pred_plot_nordics_eurosceptic + ggtitle("Scandinavia") + labs(x = "Differentiation (article-years (logged))",
@@
                            labels = c("Non-Eurosceptic", "Eurosceptic"))
-
 ##CEE
@@
                                 legend.title = "Eurosceptic", colors = "bw")
-
 pred_plot_CEE_eurosceptic_final <-  pred_plot_CEE_eurosceptic + ggtitle("Central and Eastern Europe") + labs(x = "Differentiation (article-years (logged))",
@@
 library(patchwork)
-
 regional_euroscepticism_final <- pred_plot_nordics_eurosceptic/pred_plot_CEE_eurosceptic_final + plot_annotation(tag_levels = "A", title = "Support for constitutional differentiation", subtitle = "Interaction of Euroscepticism and exposure to differentiation in CEE and Scandinavia", theme = theme(plot.title = element_text(size = 16)))
-
 regional_euroscepticism_final
@@
 
-I must also reject *H2b*, as the effect of the interaction between differentiated integration and Euroscepticism is statistically insignificant in the Central and Eastern Europe data.
-I furthermore reject *H4b*, as there is no statistically significant difference between the effect of exposure for Eurosceptics and supporters of EU membership in Scandinavia.
-The lack of a lower level of support for differentiated integration among Central and Eastern Europeans, despite the fact that the discriminatory imposition of differentiation has been especially prevalent in these countries, suggests that historical trajectories and past experience with integration may predominantly lead to altered support for differentiation when it is used as a part of elite discourse.
+I also reject *H2b*, as the effect of the interaction between differentiated integration and Euroscepticism is statistically insignificant in the Central and Eastern Europe data.
+Lastly, I reject *H4b*, as there is no statistically significant difference between the effect of exposure for Eurosceptics and supporters of EU membership in Scandinavia.
+The lack of a lower level of support for differentiated integration among Central and Eastern Europeans, even if the discriminatory imposition of differentiation has been prevalent in these countries, suggests that historical experience with differentiated integration may shape support for it mainly when it is used as a part of elite discourse.
 This finding is important for our understanding of how attitudes towards greater differentiation can be shaped by political elites in the future.
@@
 In times where EU solidarity is framed as a net positive for its citizens, support for differentiation may be lower than it otherwise would have been.
-More research is thus needed to establish whether the patterns established by this paper are consistent also in the future.
+More research is thus needed to establish whether the patterns established by this paper hold also in the future.
 
@@
 I use an approach outlined in @cinelli_making_2020 to test the sensitivity of several of my models to omitted variable bias (see figures A1-4).
-This quantifies the remaining variance that a potential confounder would need to account for to nullify the effect of theoretical interest, exposure to differentiated integration.
-A test of the sensitivity of the interaction between Euroscepticism and exposure to differentiation in Scandinavia (figure A3) shows that a variable explaining 50% of the remaining variance of support for constitutional differentiation would need to explain less than 10% of the remaining variance in real exposure to differentiation to nullify the effect of exposure.
-One potentially important omitted variable is knowledge of the EU [@leuffen_public_2022].
+This quantifies the variance that a potential confounder would need to account for to nullify the effect of exposure to differentiated integration.
+A test of the sensitivity of the interaction between Euroscepticism and exposure to differentiation in Scandinavia (figure A3) shows that a variable explaining 50% of the remaining variance of support for constitutional differentiation would need to explain less than 10% of the remaining variance in real exposure to differentiation to nullify its effect.
+One important omitted variable is knowledge of the EU [@leuffen_public_2022].
 While knowledge of the EU is unlikely to be strongly connected with actual levels of differentiation, its strong connection to support for differentiation makes its omission particularly problematic.
-This suggests that omitted variable bias is likely to be an analytical concern throughout the entire study.
-Future studies should mitigate these concerns by including knowledge questions about current EU politics, to allow researchers to include similar controls as those commonly found in other EU public opinion research.
+Omitted variable bias is thus likely to be an analytical concern throughout the entire study.
+Future studies must mitigate these concerns by including knowledge questions about current EU politics, to allow researchers to include a control similar to that found in other EU public opinion research.
 
-I also use multiple imputation through chained equations [@buuren_mice_2011; @fjelstul_explaining_2022] to test how sensitive the results are to bias caused by underlying patterns of data missingness.
-Multiple imputation uses each unit’s observed data to derive probable values for all covariates where a unit has missing data.
-This procedure is iterated many times, before the estimates derived from analysis of each imputed dataset are pooled.
-This procedure better accounts for the inherent uncertainty associated with data imputation and will thus yield more realistic parameter estimates and standard errors than traditional mean or zero imputation.
-As appendix tables A2-3 show, the multiply imputed models are similar to those found in tables 2-3.
-One important exception is the effect of exposure to differentiated integration in Central and Eastern Europe: While there is a weak positive effect of exposure in the main model, the multiply imputed model shows a weakly negative effect.
-However, both are statistically insignificant.
-Despite this minor deviation, the tests suggest that the effects established in this paper are not primarily driven by systematic data missingness.
+I also use multiple imputation through chained equations [@buuren_mice_2011; @fjelstul_explaining_2022] to test how sensitive the results are to bias caused by potentially systematic data missingness.
+Multiple imputation better accounts for the uncertainty associated with data imputation than traditional mean or zero imputation methods, and provides more realistic standard errors.
+This is particularly beneficial when testing the statistical significance of multiply imputed estimates.
+As appendix tables A2-3 show, the multiply imputed models are substantively similar to those found in tables 2-4.
+This suggests that the effect sizes or directions are not driven by systematic data missingness.
 
-I also test whether the substantial overlap between Scandinavian Eurosceptics and non-Eurosceptics is driven by the very low variability, with a standard deviation of `r round(sd(nordic_data$article_disc_log), digits = 2)`,  in the measure of exposure to differentiated integration in the Scandinavian data.
+I also test whether the substantial overlap between Scandinavian Eurosceptics and non-Eurosceptics is driven by the very low variability, with a standard deviation of `r round(sd(nordic_data$article_disc_log), digits = 2)`,  in the measure of exposure to differentiated integration found in data from this region.
 I do this by testing a three-way interaction between Scandinavian citizenship, exposure to differentiated integration and Euroscepticism in the full data.
-Appendix table A6 suggests that the results are substantially similar to the interaction effect of Scandinavian citizenship and Euroscepticism shown in table.
-The results thus suggest that the estimates are not unduly biased by the very small standard deviation .
+Appendix table A6 suggests that the results are substantially similar to the interaction effect of Scandinavian citizenship and Euroscepticism shown in table 4.
+The results thus suggest that the estimates are not unduly biased by the very small standard deviation found in this particular region.
 
 Lastly, I test how robust the results are to an alternative operationalization of Euroscepticism: satisfaction with EU democracy.
-Dissatisfaction with the state of EU democracy may be seen as a form of "regime scepticism", characterized by a dissatisfaction with EU decision-making, than "exit scepticism", as expressed through a desire to leave the union altogether [@vries_euroscepticism_2018].
-Figure A5 ( coefficients in table A4) shows similar trends for both operationalizations: Those least satisfied with EU democracy, and who thus express the greatest amount of regime scepticism, also express the greatest support for constitutional differentiated integration.
+Dissatisfaction with the state of EU democracy may be seen as a form of "regime scepticism", characterized by a dissatisfaction with EU decision-making, rather than the "exit scepticism" expressed by a desire to leave the union altogether [@vries_euroscepticism_2018].
+Figure A5 (coefficients in table A4) shows similar trends for both operationalizations: Those least satisfied with EU democracy, and who thus express the greatest amount of regime scepticism, also express greater support for constitutional differentiated integration than those more satisfied with EU democracy.
 Their support for differentiation also increases more strongly as a function of exposure to previous differentiation than what is the case for those more satisfied with EU democracy.
@@
 
-This paper establishes, through a combination of data surveying citizen preferences for differentiated integration and their countries’ actual exposure to differentiation from the EU treaties, that there is a strong association between support for differentiated integration and past exposure to it.
-This may be because citizens more accustomed to differentiated integration potentially perceive it as a favourable way of combining the benefits of national membership of the EU with greater national autonomy.
-Importantly, I show that a potential tendency to see the kind of exposure that one has been exposed to as a preferable model for the future EU is, broadly speaking, conditioned by whether one opposes or supports EU membership.
+This paper establishes, by combining data surveying citizen preferences for differentiated integration with data on their countries’ actual exposure to differentiated integration, that there is a strong association between support for differentiated integration and past exposure to it.
+This may be because citizens more accustomed to differentiated integration may see it as a favourable way of combining the benefits of national membership of the EU with greater national autonomy.
+Importantly, this effect appears moderated by whether one opposes or supports EU membership.
 However, the results do not suggest that the mode of implementation of the differentiation that the same citizens have been exposed to play any large role in shaping attitudes towards a future differentiated EU.
 
-What this suggests is that even if some differentiated integration could be perceived as a cartelization of the benefits of integration and as a dominating imposition on the political freedom of a polity, this does not necessarily translate to less support for differentiated integration as a policy.
-While my article does not establish why this is the case, it does strongly suggest that there is only a limited learning effect from exposure to differentiated integration.
-This is important because it implies that even the dominating imposition of a particular set of policies may not be enough to appreciably alter attitudes towards the same policy even among those who may be most likely to be critical of EU integration.
-Understanding what kind of discourses are more likely to produce critical attitudes towards differentiated integration in this group is an important question for future research.
+What this suggests is that even if some differentiated integration could be perceived as a cartelization of the benefits of integration that also constrains the political autonomy of those subject to it, this does not necessarily translate to less support for differentiated integration as a policy.
+While my article does not establish why this is the case, it does strongly suggest that the positive or negative effect of historical exposure to differentiated integration on support for it in the future depends on whether political elites mobilize this discourse as a part of their efforts to shape attitudes towards the question.
+Understanding what kind of discourses are more likely to produce critical attitudes towards differentiated integration is thus an important question for future research.
 
 Another question is whether the public is equally likely to prefer differentiation in all policy areas.
-Existing research [@de_blok_blessing_2021] has largely focused on individual support for differentiated integration of the EU as a polity.
+Existing research [@de_blok_blessing_2021; @telle_elite_2022; @leuffen_public_2022] has largely focused on individual support for differentiated integration of the EU as a polity.
 It is both possible and probable that even those favouring an “ever closer Union” may have diverging preferences for integration at the level of individual policies.
-Testing where individual citizens may prefer more differentiation may also help theorize the ambivalent attitudes towards EU integration found by an increasingly prevalent strand of the EU public opinion literature [@de_vries_variable_2013; @goldberg_one_2021; @moland_constraining_2022].
+Testing where individual citizens may prefer more differentiation may also help theorize the ambivalent attitudes towards EU integration found in previous EU public opinion literature [@de_vries_variable_2013; @goldberg_one_2021; @moland_constraining_2022].
 
@@
 However, beyond the EU, the results also suggest that the public legitimacy of deeper political integration may be challenged if citizens gradually become accustomed to more differentiated and non-uniform forms of integration.
-While one should be wary of generalizing findings from the EU to other international organizations, my results suggest that allowing some countries to opt out of the legal order that is increasingly developing beyond the EU [@zurn_theory_2018] may lead to greater popular opposition to the idea that countries should be members of entire frameworks of political cooperation, rather than more functionally limited partnerships for solving particular problems.
+While one should be wary of generalizing findings from the EU to other international organizations, my results suggest that allowing some countries to opt out of the legal order that is increasingly developing also beyond the EU [@zurn_theory_2018] may lead to greater popular opposition to the idea that countries should be members of entire frameworks of political cooperation, rather than more functionally limited partnerships dedicated to solving particular problems.
 
@@
 <div id="refs"></div>
-
-# Appendix
-\setcounter{table}{0}
-\renewcommand*{\thetable}{A\arabic{table}}
-```{r, results='asis'}
-
-appendix_items <- readRDS("data/appendix_items.Rds")
-library(psychTools)
-
-cat(fa2latex(appendix_items$`Factor analysis of economy`, heading = "Factor analysis of perception of economy", silent = TRUE, caption = "Factor analysis of perceptions of the economy."))
-```
-\setcounter{figure}{0}
-\renewcommand*{\thefigure}{A\arabic{figure}}
-```{r, fig.cap="Sensitivity to omitted variable bias for article-years (logged)", fig.height=8.5}
-library(sensemakr)
-appendix_items$"Fixed effect of non-interacted model" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
-                          eurosceptic + gender + excl_identity + factor(country), weights = EUI_final_DI_disc$weight, 
-                        data = EUI_final_DI_disc)
-
-#Run sensemakr
-appendix_items$"Sensitivity test of non-interacted model" <- sensemakr::sensemakr(appendix_items$`Fixed effect of non-interacted model`, 
-                                treatment = "article_disc_log")
-
-plot(appendix_items$"Sensitivity test of non-interacted model", type = "extreme")
-
-```
-
-```{r, fig.cap="Sensitivity to omitted variable bias for article-years (logged) X Euroscepticism", fig.height=8.5}
-
-appendix_items$"Fixed effect of interacted model" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
-                        eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + factor(country), weights = EUI_final_DI_disc$weight, 
-                      data = EUI_final_DI_disc)
-
-appendix_items$"Sensitivity test of interacted model" <- sensemakr(appendix_items$`Fixed effect of interacted model`, 
-                                  treatment = "article_disc_log:eurosceptic1")
-
-plot(appendix_items$`Sensitivity test of interacted model`, type = "extreme")
-
-```
-
-```{r, fig.cap="Sensitivity to omitted variable bias for article-years (logged) X Euroscepticism in Scandinavia", fig.height=8.5}
-nordic_data <- readRDS("data/nordic_data.Rds")
-CEE_data <- readRDS("data/CEE_data.Rds")
-
-appendix_items$"Fixed effect of interacted model (Nordics)" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
-                                                          eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + factor(country), weights = nordic_data$weight, 
-                                                        data = nordic_data)
-
-appendix_items$"Sensitivity test of interacted model (Nordics)" <- sensemakr(appendix_items$`Fixed effect of interacted model (Nordics)`, 
-                                                                   treatment = "article_disc_log:eurosceptic1")
-
-plot(appendix_items$`Sensitivity test of interacted model (Nordics)`, type = "extreme")
-```
-
-```{r, fig.cap="Sensitivity to omitted variable bias for article-years (logged) X Euroscepticism in Central and Eastern Europe", fig.height=8.5}
-library(sensemakr)
-appendix_items$"Fixed effect of interacted model (CEE)" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
-                                                                    eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + factor(country), weights = CEE_data$weight, 
-                                                                  data = CEE_data)
-
-appendix_items$"Sensitivity test of interacted model (CEE)" <- sensemakr(appendix_items$`Fixed effect of interacted model (CEE)`, 
-                                                                             treatment = "article_disc_log:eurosceptic1")
-
-plot(appendix_items$`Sensitivity test of interacted model (CEE)`, type = "extreme")
-
-```
-
-```{r}
-
-library(mice)
-library(miceadds)
-library(merTools)
-main_impute <- readRDS("data/main_impute.Rds")
-```
-
-```{r}
-coef_map_regions2 <- c('(Intercept)' = 'Constant',
-                          'article_disc_log:nordic_dummy' = 'Article-years (logged) X Scandinavia',
-                      'article_disc_log:CEE_dummy' = 'Article-years (logged) X CEE',
-                      "article_disc_log:eurosceptic1" = "Article-years (logged) X Eurosceptic",
-                          'article_disc_log' = 'Article-years (logged)',
-                          'market_europe' = 'Liberal economic values',
-                          'alternative econ' = 'Perception of economy',
-                          'gov_trust' = 'Support national government',
-                          'left_right' = 'Left-right',
-                          'left_right_sqr' = 'Left-right (sqr)',
-                          'eurosceptic1' = 'Eurosceptic',
-                          'gender' = 'Gender',
-                          'excl_identity' = 'Exclusively national identity', 
-                      "nordic_dummy" = "Scandinavia",
-                      "CEE_dummy" = "Central and Eastern Europe")
-
-modelsummary(main_impute, fmt = 3, output = "latex", coef_map = coef_map_regions2, 
-             stars = TRUE, title = "Imputed model of interaction between Euroscepticism and article-years (logged) and model with regional dummies")
-```
-
-```{r}
-regional_impute <- readRDS("data/regional_impute.Rds")
-
-coef_map_eurosceptic2 <- c('(Intercept)' = 'Constant',
-                          'article_disc_log:eurosceptic1' = 'Article-years (logged) X Euroscepticism',
-                          'article_disc_log' = 'Article-years (logged)',
-                          'market_europe' = 'Liberal economic values',
-                          'alternative econ' = 'Perception of economy',
-                          'gov_trust' = 'Support national government',
-                          'left_right' = 'Left-right',
-                          'left_right_sqr' = 'Left-right (sqr)',
-                          'eurosceptic1' = 'Eurosceptic',
-                          'gender' = 'Gender',
-                          'excl_identity' = 'Exclusively national identity')
-
-modelsummary::modelsummary(regional_impute, fmt = 3, output = "latex", coef_map = coef_map_eurosceptic2, 
-             stars = TRUE, title = "Imputed models showing interactions of Euroscepticism and article-years (logged) by region")
-```
-
-```{r}
-
-coef_map_alteurosceptic <- c('(Intercept)' = 'Constant',
-                          'article_disc_logeu_democ_satisf' = 'Article-years (logged) X Euroscepticism',
-                          'article_disc_log' = 'Article-years (logged)',
-                          'market_europe' = 'Liberal economic values',
-                          'alternative econ' = 'Perception of economy',
-                          'gov_trust' = 'Support national government',
-                          'left_right' = 'Left-right',
-                          'left_right_sqr' = 'Left-right (sqr)',
-                          'eu_democ_satisf' = 'Eurosceptic',
-                          'gender' = 'Gender',
-                          'excl_identity' = 'Exclusively national identity')
-
-modelsummary::modelsummary(appendix_items$`Alternative conceptualization of Euroscepticism`, fmt = 3, output = "latex", coef_map = coef_map_alteurosceptic, 
-             stars = TRUE, title = "Model showing alternative conceptualization of Euroscepticism")
-```
-
-```{r, fig.cap="Predicted values of support for constitutional differentiated integration as a function of an interaction between exposure to differentiated integration and satisfaction with EU democracy.
-Slopes for minimum, median and maximum values of satisfaction with EU democracy are shown.
-95% CIs", fig.height=8.5}
-
-pred_plot_interact_main <- plot_model(appendix_items$`Alternative conceptualization of Euroscepticism`, type = "pred", terms = c("article_disc_log", "eu_democ_satisf"), 
-                                      colors = "bw", legend.title = "EU democracy satisfaction")
-
-pred_plot_interact_main_final <-  pred_plot_interact_main + ggtitle("Support for constitutional differentiation", 
-                                                                    subtitle = "Interaction between EU democracy satisfaction and differentiation") + labs(x = "Differentiation (article-years (logged))",
-                                                                                                                                                           y = "Predicted value") + scale_linetype_manual(name = "EU democracy satisfaction",
-                           values = c("dashed", "solid", "dotted"),
-                           labels = c("Minimum", "Median", "Max")) + font_size(title = 15, labels.x = 15, labels.y = 15)
-
-pred_plot_interact_main_final
-```
-
-```{r, fig.cap="Number of differentiations from EU primary law 1952-2020 (log-transformed and recency-weighted article-years) by EU member state.", fig.height=8.5}
-bar_chart_articleyear <- ggplot(EUI_final_DI_disc, aes(x = factor(country), y = article_disc_log)) +
-  geom_col(stat = "identity", position = "dodge")
-
-bar_chart_articleyear + labs(x = "Country", y = "Discounted article-years (log-transformed)") +
-  scale_x_discrete(breaks = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14),
-                   labels = c("UK", "DK", "FI", "FR", "DE", "SE", "GR", "HU", "IT",
-                              "LT", "NL", "PL", "RO", "ES")) +
-  ggtitle(label = "Differentiated integration in the EU", 
-          subtitle = "Primary law differentiations (article-years log-transformed and weighted)") + font_size(title = 15, labels.x = 15, labels.y = 15)
-```
-
-```{r}
-southern_mod = list()
-
-southern_mod$"Southern Europe" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
-                           eurosceptic + gender + excl_identity + south_dummy + south_dummy*article_disc_log + (1|country), weights = EUI_final_DI_disc$weight, 
-                         data = EUI_final_DI_disc)
-
-coef_map_southern <- c('(Intercept)' = 'Constant',
-                          'article_disc_logsouthern_dummy1' = 'Article-years (logged) X Southern Europe',
-                          'article_disc_log' = 'Article-years (logged)',
-                          'market_europe' = 'Liberal economic values',
-                          'alternative econ' = 'Perception of economy',
-                          'gov_trust' = 'Support national government',
-                          'left_right' = 'Left-right',
-                          'left_right_sqr' = 'Left-right (sqr)',
-                          'eurosceptic1' = 'Eurosceptic',
-                          'gender' = 'Gender',
-                          'excl_identity' = 'Exclusively national identity', 
-                      "south_dummy1" = "Southern Europe")
-
-modelsummary::modelsummary(southern_mod, fmt = 3, output = "latex", coef_map = coef_map_alteurosceptic, 
-             stars = TRUE, title = "Model showing effect of exposure in Southern Europe")
-
-```
-
-```{r, fig.pos="H", fig.cap= "Support for differentiated integration in Southern Europe as an interaction between Euroscepticism and exposure to differentiated integration. 95% CIs.", fig.height=8.5}
-
-pred_plot_southern <- plot_model(southern_mod$`Southern Europe`, type = "pred", terms = c("article_disc_log", "south_dummy"), 
-                                 colors = "bw", legend.title = "Southern Europe")
-
-pred_plot_southern_final <-  pred_plot_southern + ggtitle("Support for constitutional differentiation (Southern Europe)", subtitle = "Support for differentiation among Southern and non-Southern Europeans") + labs(x = "Differentiation (article-years (logged))",
-                                                                                                                                 y = "Predicted value")
-pred_plot_southern_final + font_size(title = 15, labels.x = 15, labels.y = 15) + scale_linetype_manual(name = "Southern Europe",
-                           values = c("dashed", "solid"),
-                           labels = c("Non-Southern", "Southern"))
-```
-
-```{r }
-scandi_eurosceptic = list()
-
-scandi_eurosceptic$"Scandinavia (three-way interactions)" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
-                             eurosceptic + gender + excl_identity + nordic_dummy + article_disc_log*eurosceptic*nordic_dummy + (1|country), weights = EUI_final_DI_disc$weight, 
-                           data = EUI_final_DI_disc)
-
-coef_map_nordic <- c('(Intercept)' = 'Constant',
-                          'article_disc_logeurosceptic1nordic_dummy1' = 'Article-years (logged) X Eurosceptic X Scandinavia',
-                          'article_disc_log' = 'Article-years (logged)',
-                          'nordic_dummy1' = "Scandinavia",
-                          'market_europe' = 'Liberal economic values',
-                          'alternative econ' = 'Perception of economy',
-                          'gov_trust' = 'Support national government',
-                          'left_right' = 'Left-right',
-                          'left_right_sqr' = 'Left-right (sqr)',
-                          'eurosceptic1' = 'Eurosceptic',
-                          'gender' = 'Gender',
-                          'excl_identity' = 'Exclusively national identity')
-
-modelsummary::modelsummary(scandi_eurosceptic, fmt = 3, output = "latex", coef_map = coef_map_nordic,
-             stars = TRUE, title = "Model showing three-way interaction between Scandinavian dummy, Eurosceptic and article-years (logged)")
-```

### 2022-09-05 appendix.Rmd
Uploaded appendix Rmd file and revised main manuscript.

---
title: "APPENDIX: 'Does differentiation beget differentiation?'"
author:
- name: Martin Moland
  affiliation: ARENA, Centre for European Studies
geometry: margin=1in
date: "`r format(Sys.time(), '%B %d, %Y')`"
fontsize: 11pt
output: pdf_document
keywords: ""
endnotes: FALSE
# csl: inst/jpr.csl
mainfont: cochineal
# sansfont: Linux Biolinum O
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: ""
  removetitleabstract: ""
appendix: TRUE
appendixletter: "A"
pandocparas: FALSE
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
linkcolor: black
header-includes:
  - \usepackage{longtable}
  - \usepackage{csquotes}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \renewcommand{\thepage}{A-\arabic{page}}
  - \usepackage{booktabs}
  - \usepackage{siunitx}
  - \newcolumntype{d}{S[input-symbols = ()]}
---

```{r setup, include=FALSE}
is_docx <- knitr::pandoc_to("docx")
is_latex <- knitr::pandoc_to("latex")

table_format <- ifelse(is_docx, "huxtable", 'default')

# I don't know how Texas Instruments smart this is, but p-sure default DPI is 96.
# That's not a problem for LaTeX, but it looks not-so-great for Word.
# For Word, let's up that to 600. This should allow for cross-referencing 
# in {bookdown} while allowing for conditional DPI

if (is_latex) {
  conditional_dpi <- 96
  return(conditional_dpi)
} else {
  conditional_dpi <- 600
  return(conditional_dpi)
}

knitr::opts_chunk$set(echo = FALSE,
                      message=FALSE, warning=FALSE,
                      fig.path='doc/figs/',
                      cache.path = 'doc/_cache/',
                      fig.width = 8.5, dpi = conditional_dpi,
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      x2 = paste0(stringr::str_extract(x2, '.*/'), 
                                  "appendix-", 
                                  stringr::str_remove(x2, ".*/"))
                      if (file.rename(x, x2)) x2 else x
                      })

options(knitr.kable.NA = '')
```

\setcounter{table}{0}
\renewcommand*{\thetable}{A\arabic{table}}
```{r, results='asis'}
appendix_items <- readRDS("data/appendix_items.Rds")
EUI_final_DI_disc <- readRDS("data/EUI_final_DI_disc.Rds")

library(psychTools)
cat(fa2latex(appendix_items$`Factor analysis of economy`, heading = "Factor analysis of perception of economy", silent = TRUE, caption = "Factor analysis of perceptions of the economy."))
```

\setcounter{figure}{0}
\renewcommand*{\thefigure}{A\arabic{figure}}
```{r, fig.cap="Sensitivity to omitted variable bias for article-years (logged)", fig.height=8.5}
library(sensemakr)
appendix_items$"Fixed effect of non-interacted model" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
                          eurosceptic + gender + excl_identity + factor(country), weights = EUI_final_DI_disc$weight, 
                        data = EUI_final_DI_disc)
#Run sensemakr
appendix_items$"Sensitivity test of non-interacted model" <- sensemakr::sensemakr(appendix_items$`Fixed effect of non-interacted model`, 
                                treatment = "article_disc_log")
plot(appendix_items$"Sensitivity test of non-interacted model", type = "extreme")
```

```{r, fig.cap="Sensitivity to omitted variable bias for article-years (logged) X Euroscepticism", fig.height=8.5}
appendix_items$"Fixed effect of interacted model" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
                        eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + factor(country), weights = EUI_final_DI_disc$weight, 
                      data = EUI_final_DI_disc)
appendix_items$"Sensitivity test of interacted model" <- sensemakr(appendix_items$`Fixed effect of interacted model`, 
                                  treatment = "article_disc_log:eurosceptic1")
plot(appendix_items$`Sensitivity test of interacted model`, type = "extreme")
```

```{r, fig.cap="Sensitivity to omitted variable bias for article-years (logged) X Euroscepticism in Scandinavia", fig.height=8.5}
nordic_data <- readRDS("data/nordic_data.Rds")
CEE_data <- readRDS("data/CEE_data.Rds")
appendix_items$"Fixed effect of interacted model (Nordics)" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
                                                          eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + factor(country), weights = nordic_data$weight, 
                                                        data = nordic_data)
appendix_items$"Sensitivity test of interacted model (Nordics)" <- sensemakr(appendix_items$`Fixed effect of interacted model (Nordics)`, 
                                                                   treatment = "article_disc_log:eurosceptic1")
plot(appendix_items$`Sensitivity test of interacted model (Nordics)`, type = "extreme")
```

```{r, fig.cap="Sensitivity to omitted variable bias for article-years (logged) X Euroscepticism in Central and Eastern Europe", fig.height=8.5}
library(sensemakr)
appendix_items$"Fixed effect of interacted model (CEE)" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
                                                                    eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + factor(country), weights = CEE_data$weight, 
                                                                  data = CEE_data)
appendix_items$"Sensitivity test of interacted model (CEE)" <- sensemakr(appendix_items$`Fixed effect of interacted model (CEE)`, 
                                                                             treatment = "article_disc_log:eurosceptic1")
plot(appendix_items$`Sensitivity test of interacted model (CEE)`, type = "extreme")
```

```{r, results='asis'}
library(mice)
library(miceadds)
library(merTools)
main_impute <- readRDS("data/main_impute.Rds")

coef_map_regions2 <- c('(Intercept)' = 'Constant',
                          'article_disc_log:nordic_dummy' = 'Article-years (logged) X Scandinavia',
                      'article_disc_log:CEE_dummy' = 'Article-years (logged) X CEE',
                      "article_disc_log:eurosceptic1" = "Article-years (logged) X Eurosceptic",
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'excl_identity' = 'Exclusively national identity', 
                      "nordic_dummy" = "Scandinavia",
                      "CEE_dummy" = "Central and Eastern Europe")

modelsummary::modelsummary(main_impute, fmt = 3, output = "latex", coef_map = coef_map_regions2, 
             stars = TRUE, title = "Imputed model of interaction between Euroscepticism and article-years (logged) and model with regional dummies")
```

```{r, results='asis'}
regional_impute <- readRDS("data/regional_impute.Rds")
coef_map_eurosceptic2 <- c('(Intercept)' = 'Constant',
                          'article_disc_log:eurosceptic1' = 'Article-years (logged) X Euroscepticism',
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'excl_identity' = 'Exclusively national identity')
modelsummary::modelsummary(regional_impute, fmt = 3, output = "latex", coef_map = coef_map_eurosceptic2, 
             stars = TRUE, title = "Imputed models showing interactions of Euroscepticism and article-years (logged) by region")

coef_map_alteurosceptic <- c('(Intercept)' = 'Constant',
                          'article_disc_logeu_democ_satisf' = 'Article-years (logged) X Euroscepticism',
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eu_democ_satisf' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'excl_identity' = 'Exclusively national identity')

modelsummary::modelsummary(appendix_items$`Alternative conceptualization of Euroscepticism`, fmt = 3, output = "latex", coef_map = coef_map_alteurosceptic, 
             stars = TRUE, title = "Model showing alternative conceptualization of Euroscepticism")
```

```{r, fig.cap="Predicted values of support for constitutional differentiated integration as a function of an interaction between exposure to differentiated integration and satisfaction with EU democracy. Slopes for minimum, median and maximum values of satisfaction with EU democracy are shown. 95% CIs", fig.height=8.5}
library(sjPlot)
library(ggplot2)
pred_plot_interact_main <- plot_model(appendix_items$`Alternative conceptualization of Euroscepticism`, type = "pred", terms = c("article_disc_log", "eu_democ_satisf"), 
                                      colors = "bw", legend.title = "EU democracy satisfaction")
pred_plot_interact_main_final <-  pred_plot_interact_main + ggtitle("Support for constitutional differentiation", 
                                                                    subtitle = "Interaction between EU democracy satisfaction and differentiation") + labs(x = "Differentiation (article-years (logged))",
                                                                                                                                                           y = "Predicted value") + scale_linetype_manual(name = "EU democracy satisfaction",
                           values = c("dashed", "solid", "dotted"),
                           labels = c("Minimum", "Median", "Max")) + font_size(title = 15, labels.x = 15, labels.y = 15)
pred_plot_interact_main_final
```

```{r, fig.cap="Number of differentiations from EU primary law 1952-2020 (log-transformed and recency-weighted article-years) by EU member state.", fig.height=8.5}
bar_chart_articleyear <- ggplot(EUI_final_DI_disc, aes(x = factor(country), y = article_disc_log)) +
  geom_col(stat = "identity", position = "dodge")
bar_chart_articleyear + labs(x = "Country", y = "Discounted article-years (log-transformed)") +
  scale_x_discrete(breaks = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14),
                   labels = c("UK", "DK", "FI", "FR", "DE", "SE", "GR", "HU", "IT",
                              "LT", "NL", "PL", "RO", "ES")) +
  ggtitle(label = "Differentiated integration in the EU", 
          subtitle = "Primary law differentiations (article-years log-transformed and weighted)") + font_size(title = 15, labels.x = 15, labels.y = 15)
```

```{r, results='asis'}
southern_mod = list()
southern_mod$"Southern Europe" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                           eurosceptic + gender + excl_identity + south_dummy + south_dummy*article_disc_log + (1|country), weights = EUI_final_DI_disc$weight, 
                         data = EUI_final_DI_disc)
coef_map_southern <- c('(Intercept)' = 'Constant',
                          'article_disc_logsouthern_dummy1' = 'Article-years (logged) X Southern Europe',
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'excl_identity' = 'Exclusively national identity', 
                      "south_dummy1" = "Southern Europe")

modelsummary::modelsummary(southern_mod, fmt = 3, output = "latex", coef_map = coef_map_alteurosceptic, 
             stars = TRUE, title = "Model showing effect of exposure in Southern Europe")
```

```{r, fig.pos="H", fig.cap= "Support for differentiated integration in Southern Europe as an interaction between Euroscepticism and exposure to differentiated integration. 95% CIs.", fig.height=8.5}
pred_plot_southern <- plot_model(southern_mod$`Southern Europe`, type = "pred", terms = c("article_disc_log", "south_dummy"), 
                                 colors = "bw", legend.title = "Southern Europe")
pred_plot_southern_final <-  pred_plot_southern + ggtitle("Support for constitutional differentiation (Southern Europe)", subtitle = "Support for differentiation among Southern and non-Southern Europeans") + labs(x = "Differentiation (article-years (logged))",
                                                                                                                                 y = "Predicted value")
pred_plot_southern_final + font_size(title = 15, labels.x = 15, labels.y = 15) + scale_linetype_manual(name = "Southern Europe",
                           values = c("dashed", "solid"),
                           labels = c("Non-Southern", "Southern"))
```

```{r }
scandi_eurosceptic = list()
scandi_eurosceptic$"Scandinavia (three-way interactions)" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                             eurosceptic + gender + excl_identity + nordic_dummy + article_disc_log*eurosceptic*nordic_dummy + (1|country), weights = EUI_final_DI_disc$weight, 
                           data = EUI_final_DI_disc)
coef_map_nordic <- c('(Intercept)' = 'Constant',
                          'article_disc_logeurosceptic1nordic_dummy1' = 'Article-years (logged) X Eurosceptic X Scandinavia',
                          'article_disc_log' = 'Article-years (logged)',
                          'nordic_dummy1' = "Scandinavia",
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'excl_identity' = 'Exclusively national identity')
modelsummary::modelsummary(scandi_eurosceptic, fmt = 3, output = "latex", coef_map = coef_map_nordic,
             stars = TRUE, title = "Model showing three-way interaction between Scandinavian dummy, Eurosceptic and article-years (logged)")
```

### 2022-10-28 Manuscript_3_draft1.Rmd
Newest version of paper.

@@
 subtitle: Examining the link between implementation of and support for differentiation
-thanks: "This is a very early draft. All comments are welcome at martin.moland@arena.uio.no. This version was finalized at `r format(Sys.time(), '%B %d, %Y')`."
+thanks: "This is a very early draft. All comments are welcome at martin.moland@arena.uio.no. This version was finalized on `r format(Sys.time(), '%B %d, %Y')`."
 author:
@@
 abstract: "Questions about the public legitimacy of the EU’s flexible integration is emerging as an important question for European public opinion research.
-However, we still know very little about how attitudes towards the EU's differentiated integration are shaped by exposure to such differentiation in the past, and how this effect relates to individual Euroscepticism.
-Combining recent survey data with data showing country-level exposure to differentiated integration, I show that those who have been exposed to more differentiation are likelier to want a future EU that allows countries to opt out of undesirable policy integration.
-This effect is particularly strong among Eurosceptics.
-However, whether differentiated integration is voluntarily chosen or discriminatorily imposed seems to be less important.
-Altogether, my results suggest that exposure to the differentiated integration often used to integrate highly salient policy areas can solidify popular support for a less integrated EU.
-This potentially complicates the EU’s move towards an ‘ever closer Union’."
+However, we still know do not know whether the increased familiarity with differentiated integration that exposure to it brings leads to greater support a differentiated EU, and whether the mode of implementation and attitudes towards the EU moderates this effect.
+Combining recent survey data with data showing country-level exposure to differentiated integration, I show that familiarity with differentiation seems to breed greater support for it.
+This particularly applies to Eurosceptics.
+It also suggests that support for differentiated integration may be higher where governments frame it as a source of greater national autonomy.
+My results suggest that exposure to the differentiated integration used to integrate highly salient policy areas can solidify popular support for a less integrated EU.
+This can complicate the EU’s move towards ‘ever closer Union’."
 keywords: "differentiated integration, public opinion, Euroscepticism, European integration, contestation of the EU"
@@
   anonymous: ""
-  doublespacing: ""
+  doublespacing: TRUE
   removetitleabstract: ""
@@
 # Introduction
-Despite the EU being one of the most advanced political cooperations at the global stage, its integration has become increasingly fragmented after the ratification of the 1992 Maastricht Treaty.
-The differentiation of EU integration has primarily been driven by a desire to deepen political cooperation despite contestation of it in groups of countries [@leuffen_differentiated_2013; @schimmelfennig_ever_2020].
-Differentiation has typically taken two forms: First, some member states have opted out of cooperation to protect their own sovereignty.
-This is called “constitutional differentiation”.
-Second, “instrumental differentiation” has happened where countries with a diminished capacity for integration have been allowed to integrate at different schedules.
-Such differentiation can either happen voluntarily or because the EU's institutions impose differentiation as a precondition for membership [@schimmelfennig_instrumental_2014].
-Thus, ‘variable geography’ [@stubb_categorization_1996] is in many ways a better description of the EU's political development than the union's own goal of "ever closer Union".
 
-While increasing attention is being paid to what shapes popular preferences for such differentiation [@leuffen_public_2022; @telle_elite_2022], we know little about how actual exposure to it has impacted support for the goal of a uniformly integrated EU.
-While one of the few existing contributions examining the effect of differentiation on public opinion shows that allowing for differentiation may strengthen the perceived democratic legitimacy of the EU [@schraff_does_2020], important knowledge gaps remain: First, as mentioned, we know too little about how past exposure to differentiation shapes support for it.
-Second, our understanding of how this relationship is moderated by Eurosceptic sentiments.
-Lastly, we do not know whether the voluntarily chosen or involuntarily imposed nature of past differentiation impacts citizens' preferences for differentiation as a future model of EU integration.
+Despite the EU being one of the world's most advanced political cooperations, its integration has become increasingly fragmented after the ratification of the 1992 Maastricht Treaty.
+The differentiation of EU integration has primarily been driven by a desire to deepen political cooperation even where it may be contested by some of the EU's members [@leuffen_differentiated_2013; @schimmelfennig_ever_2020], and has typically taken two forms: First, some member states have opted out of cooperation to protect their own sovereignty.
+This is called "constitutional differentiation".
+Second, "instrumental differentiation" has happened where countries with a diminished capacity for integration have been allowed to integrate at different schedules.
+Such differentiation can either happen voluntarily or because the EU has imposed differentiation as a precondition for membership [@schimmelfennig_instrumental_2014].
+Thus, 'variable geography' [@stubb_categorization_1996] better describes the EU's political development than the EU's own goal of "ever closer Union".
 
-By combining survey data on citizens’ attitudes towards constitutional differentiated integration from 2020-2021 with data detailing country-level differentiations from the EU’s treaties, I find that support for constitutional differentiation increases with exposure to differentiations from EU law, and that the extent to which this is the case strongly depends on individual-level Euroscepticism.
-However, there does not seem to be a clear connection between how the differentiated integration citizens have been exposed to in the past have been implemented and support for a less uniformly integrated EU.
-The greater support for differentiation among Eurosceptics can be explained by how such differentiation is more likely to bring about the less integrated Europe that such individuals may want.
-However, more research is needed to understand the seeming lack of a diverging response to differentiation even where it has been implemented under very different circumstances.
+While increasing attention is being paid to what shapes popular preferences for such differentiation [@leuffen_public_2022; @telle_elite-mass_2022; @schuessler_mapping_2023], we know little about how actual exposure to it has shaped support for the goal of a uniformly integrated EU.
+While one of the few existing contributions examining the effect of differentiation on public opinion shows that allowing for differentiation may strengthen the perceived democratic legitimacy of the EU [@schraff_does_2020], important knowledge gaps remain.
+This article tests three theoretically informed hypotheses: First, it tests whether support for differentiation leads to greater support for it on average.
+Second, it tests whether those more critical of EU membership become more positively inclined to differentiated integration if exposed to more of it.
+Lastly, it tests whether those citizens exposed to differentiated integration that was voluntarily chosen by their governments are more favourably inclined towards future differentiation.
 
-In addition to expanding our knowledge of public support for differentiated EU integration, this article contributes to a broader literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009] While the EU’s legal and functional reach is much greater than that of other international organizations, many of the organizations constituting the global governance system have seen their legal and political powers expand in recent years [@zurn_theory_2018].
-Even if more research is needed to understand whether my results travel to contexts other than the EU, my results can help illuminate how the allowance of differentiations from an emerging international legal order can shape attitudes towards deepening transnational cooperation.
+By combining data from surveys fielded in 2020-2021 that measure citizens' attitudes towards constitutional differentiated integration in 13 EU member states with data detailing every country's exposure to differentiated integration, I find that support for constitutional differentiation is associated with greater exposure to differentiations from EU law, but that this effect is strongly contingent on individual-level Euroscepticism.
+I also find that support for it seems to be greater where differentiated integration has been framed as a tool for strengthening national sovereignty.
+This shows that familiarity with differentiated integration can lead to greater support for it in the future.
+However, more research is needed to understand what causes regional variations in support for differentiated integration.
 
-My findings are also important to the discussion about the EU’s future spawned by Brexit [@european_commission_white_2017]: Because EU integration has expanded into increasingly salient areas, differentiation has often been used to mitigate the challenges caused by public contestation [@schimmelfennig_ever_2020].
-By providing greater numbers of citizens with exposure to differentiation, the EU’s institutions may solidify opposition to further integration among greater numbers of citizens, potentially stalling future attempts at cooperation [@hooghe_postfunctionalist_2009].
+In addition to expanding our knowledge of public support for differentiated EU integration, this article contributes to a broader literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009].
+While the EU's legal and functional reach goes beyond that of most other international organization, many international organizations have seen their legal and political powers expand in recent years [@zurn_theory_2018].
+Even if more research is needed to understand whether my results travel to contexts beyond the EU, they help illuminate how allowing countries to opt out of international rule-making shapes attitudes towards the future organization of the international legal order.
+
+My findings are also important to the discussion about the EU's future spawned by Brexit [@european_commission_white_2017]: Because EU integration has expanded into increasingly salient areas, differentiation has often been used to reduce public contestation of integration [@schimmelfennig_ever_2020].
+By providing greater numbers of citizens with exposure to differentiation, the EU's institutions may solidify opposition to further integration among greater numbers of citizens, potentially stalling future attempts at cooperation [@hooghe_postfunctionalist_2009].
+In this sense, the EU's differentiated integration may be both a help and hindrance for future integration: On the one hand, it may help facilitate future integration of political powers traditionally considered core to statehood [@genschel_differentiated_2013].
+On the other, greater popular support for differentiated modes of integration can make it more difficult for the EU's institutions to get the public's buy-in to functionally necessary integration.
+The inevitable trade-off that this entails must be reckoned with by EU policy-makers deciding whether to allow for opt-outs from policy integration or to insist upon legal uniformity.
 
 # Public support for European integration - what we know and what we don't
-A large literature investigates individual preferences for European integration [@basile_sharing_2020; @bremer_junckers_2020; @de_vries_variable_2013; @eichenberg_post-maastricht_2007; @gabel_public_1998; @hobolt_public_2015; @hooghe_calculation_2005].
+
+A large literature investigates individual preferences for European integration [@basile_sharing_2020; @bremer_junckers_2020; @de_vries_variable_2013; @eichenberg_post-maastricht_2007; @hobolt_public_2015].
 The literature identifies three mechanisms that help shape attitudes towards the EU.
 The first is utilitarian, with support for the EU contingent on whether individuals or the social groups with which they identify stand to gain from integration [@foster_economic_2021; @gabel_public_1998].
-The second is cue-based, with citizens relying informational shortcuts like elite discourses, media messages, and national benchmarks to form opinions about European integration [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007; @vries_euroscepticism_2018].
-Lastly, strong identification with the nation-state potentially makes citizens more likely to oppose EU integration [@borzel_identity_2020; @hooghe_calculation_2005; @luedtke_european_2005].
+The second is based on cues and framings, with citizens relying on informational shortcuts like elite discourses, media messages, and national benchmarks to form opinions about European integration [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007; @de_vries_euroscepticism_2018].
+Lastly, strong identification with the nation-state makes citizens more likely to oppose EU integration [@borzel_identity_2020; @hooghe_calculation_2005; @luedtke_european_2005].
 
-An under-investigated issue is why some people support a more differentiated EU, in which the EU institutionally permits the divergence of the EU's legal order.
-Such differentiated integration is already comprehensively investigated conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004].
-The few contributions investigating individual support for differentiated integration has focused on liberal economic values [@leuffen_public_2022], Euroscepticism [de_blok_blessing_2021], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite_2022].
+An under-investigated question is why some people support a more differentiated EU, which renders the EU's integration either permanently or temporarily non-uniform.
+Differentiated integration is already investigated conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004].
+The few contributions investigating individual support for differentiated integration have focused on liberal economic values [@leuffen_public_2022], Euroscepticism [@de_blok_blessing_2023; @schuessler_mapping_2023], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite-mass_2022].
+Thus, with one notable exception, the effect of national contexts are under-studied.
+Such national contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; @harteveld_europe_2013].
+Here citizens' exposure to past differentiated integration plays a crucial role: Differentiations from EU law may create a disequilibrium between a country's actual level of integration and what its citizens would preferences.
+The same citizens may thus respond by changing their preferences [@busemeyer_beyond_2021; @wlezien_public_1995].
+While there have been many studies of the feedback loop between public opinion and the EU's ordinary policy-making [@bolstad_dynamics_2015; @de_bruycker_democratically_2020], we do not know how this feedback loop plays out when the policies being made are in essence alterations to the EU's institutional structure.
 
-Thus, with one notable exception, the effect of contextual factors are under-studied.
-Such national contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; harteveld_europe_2013].
-One potential source of support for differentiated integration may be the varying exposure that citizens of EU countries have had to it: A country’s number of differentiations from EU law may bring its level of integration in and out of alignment with what a citizen would prefer.
-If there is a misalignment between the actual and desired levels of integration, citizens may respond by either expressing greater or lesser support for a more differentiated EU that can bring about a political cooperation more aligned with their preferences [@busemeyer_beyond_2021; @wlezien_public_1995].
-Previous literature has primarily investigated the potential feedback loops between EU policy-making, such as directives and regulations, and public opinion [@bolstad_dynamics_2015; @de_bruycker_democratically_2020].
-This paper instead focuses on how alterations to the EU's legal structure, in the form of negotiated or imposed exemptions from the EU's treaties, shapes attitudes towards more or less uniform integration.
+It is reasonable to assume that the greater familiarity with differentiated integration that exposure to it will bring will lead to greater support for it in the future: Knowledge of EU policies is often, though not always, associated with greater support for EU integration [@marquart_knowing_2019; @verhaegen_does_2015].
+Similarly, familiarity with more differentiated ways of structuring EU memberships may lead to a stronger perception among citizens that EU memberships that allow countries to opt of policy integration may be less problematic.
 
-Greater exposure to differentiated integration is also likely to make citizens more familiar with EU differentiation, thus increasing their knowledge of how differentiated integration would work in the future.
-Such knowledge of EU policies is often, though not always, associated with greater support for EU integration [@marquart_knowing_2019; @verhaegen_does_2015].
-I also find it theoretically plausible that this relationship will depend on whether citizens support their countries' EU memberships or not: Because constitutional differentiated integration allows for permanent opt-outs from specific policy areas, Eurosceptics who favour leaving the EU are more likely to see differentiated integration as a viable path towards the less differentiated Europe that they primarily prefer.
-This yields two complementary hypotheses:
-
+This positive relationship is likely to be more strongly felt for those that oppose their country's EU membership.
+Since constitutional differentiated integration allows for permanent opt-outs from policy integration, Eurosceptics who favour leaving the EU may see differentiated integration as a potentially more realistic path towards the less integration Europe that they prefer.
+This yields two hypotheses:
 *H1a: Exposure to differentiated integration increases support for future differentiated integration.*  
-
 *H1b: Exposure will lead to greater support for differentiation among Eurosceptics than supporters of EU membership.*
 
-Responses to differentiated integration may also vary based on whether differentiation was imposed by the EU’s institutions or voluntarily chosen by member state governments.
-An important reason is that citizens may perceive such differentiation as the EU dominating their countries, by shutting them out of economically beneficial integration.
-As @schimmelfennig_eu_2014 shows, countries in Central and Eastern Europe were among those exposed to the greatest amount of discriminatory differentiated integration.
-Among other things, the new Eastern member states were for several years excluded from Schengen, and thus the possibility of easily moving for work purposes.
-The reason was that existing member states wanted to shield themselves from the economic destabilization believed to follow from the EU accession of countries like Poland, Estonia, and Hungary.
-As figure A6 shows, citizens of Central and Eastern Europe are among those in this sample exposed to the greatest amount of differentiated integration.
+Responses to differentiated integration may also vary depending on the situation in which differentiated integration was imposed.
+It is for instance likely that citizens may become more favourably inclined towards differentiated integration if the form of differentiated integration they have been exposed to was a result of a conscious decision to opt out on the part of their own governments.
+The reason is that these decisions may be more likely to be seen by citizens as an exercise of their own national sovereignty within the remits of an EU membership.
 
-Eurosceptics in these countries may also have more adverse reactions to differentiated integration than those supporting EU membership: Because citizens in these countries may predominantly be familiar with externally imposed and discriminatory differentiation, Eurosceptics may be more likely to identify a non-uniformly integrated EU that allows countries to opt out of integration with the potential for domination.
-Their concern for national sovereignty may again lead them to develop more strongly negative attitudes towards differentiated integration than what can be found among supporters of EU membership.
+Scandinavia offers an interesting case study of this assumption: Here differentiation has been chosen by national governments and framed as a tool for strengthening national sovereignty [@adler-nissen_opting_2014; @leruth_operationalizing_2015; @spendzharova_banking_2016], and has typically been decided as a result of a referendum.
+Previous literature has found that the renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy among Danish Eurosceptics [@schraff_does_2020].
+This shows that Scandinavian citizens may theoretically respond to differentiated integration, because of the particularly voluntary nature of how differentiated integration has been implemented in these regions, by wanting more differentiated integration in the future.
 
-Elites in Central and Eastern Europe, especially in Poland, have also framed differentiated integration as a path towards a divided EU, with a powerful core and a less powerful periphery [@cianciara_differentiated_2014 178-179].
-For example, Polish politicians opposed further integration of the Eurozone, a prominent example of differentiated integration, because they feared that the Central and Eastern member states outside the currency area would lose their influence over the EU's economic policy-making.
-As elite cues are particularly important for shaping attitudes towards less salient EU questions [@steenbergen_whos_2007], such as differentiated integration [@telle_elite_2022], citizens would be likely to respond to these discourses by expressing less support for differentiated integration.
-Thus, two different causal mechanisms could produce the same outcome.
+Both the way that differentiated integration has been framed and the particular national situation of the Scandinavian countries makes it likely that there will be an even stronger positive reaction to differentiation among Scandinavian Eurosceptics: Because their countries are both relatively small and highly dependent on European market access, Eurosceptic Scandinavians could see differentiation as a viable second-best alternative to the European disintegration that they otherwise favour.
+This effect is likely to be compounded by how Scandinavian elites have framed differentiated integration as a tool for greater sovereignty.
+I thus hypothesize:
 
-However, this history might be less relevant if current political elites do not use it to shape preferences towards differentiated.
-Instead, the increasingly Eurosceptic attitudes expressed by Central and Eastern European politicians [@borzel_identity_2020; @brack_towards_2020; @kriesi_politicization_2016] might form a more potent backdrop against which Central and Eastern European citizens might be expected to make up their minds.
-Exposure to such cues might, in contrast to what was stated before, make citizens more likely to express an increased demand for differentiated integration.
-This yields two competing hypotheses:
+*H2a: Exposure to differentiated integration will lead to greater support for it among citizens of Sweden and Denmark than in the EU as a whole.*\
+*H2b: Exposure to differentiated integration will lead to greater support for it among Scandinavian Eurosceptics than supporters of EU membership.*
 
-*H2a: Exposure to differentiated integration will lead to lower levels of support for future differentiated integration in Central and Eastern Europe than in the rest of the EU.*    
-*H2b: Exposure to differentiated integration in Central and Eastern Europe will lead to lower levels of support for it among Eurosceptics in these countries compared to supporters of EU membership.*
-*H3: Exposure to differentiated integration in Central and Eastern Europe will lead to higher levels of support for it than in the rest of the EU.*
-
-If exposure to discriminatory and externally imposed differentiation leads to less support for differentiated integration, especially among Eurosceptics, we are theoretically likely to find the opposite where differentiation has been voluntarily chosen.
-Scandinavia is an interesting case study of this theoretical assumption: Here differentiation has been chosen by national governments and justified by explicit references to how these differentiations enhanced national autonomy [@adler-nissen_opting_2014; @leruth_operationalizing_2015; @spendzharova_banking_2016].
-Previous literature has also found that the renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy among Danish Eurosceptics [@schraff_does_2020].
-In other words, it is likely that the familiarity that citizens in Scandinavia have with voluntarily chosen differentiation makes could make them more likely to support it as a future model of EU integration.
-
-This is also particularly likely to be true for Scandinavian Eurosceptics: Because their countries are both relatively small and highly dependent on European market access, Eurosceptic Scandinavians could be more likely to see differentiation as a viable second-best alternative to the European disintegration they would otherwise favour.
-It may also be the case that they, because of a generally critical attitude towards EU integration, are more likely to see differentiated integration as a solution that would better facilitate the less integrated Europe they otherwise favour.
-I hypothesize the following:
-
-*H4a: Exposure to differentiated integration will lead to greater levels of support for it among for citizens of Sweden and Denmark than in the EU as a whole.*
-
-*H4b: Exposure to differentiated integration will lead to greater support for it among Scandinavian Eurosceptics than supporters of EU membership.*
-
-My paper thus investigates the impact of exposure to differentiation on two levels: I first investigate the general effect of exposure to differentiated integration in the EU as a whole, as well as in Scandinavia and Central and Eastern Europe.
-I then test whether this relationship is moderated by pre-existing Eurosceptic sentiments.
-This lets me test both the general effect of exposure to differentiation as well as its interaction with pre-existing attitudes to European integration.
+My paper investigates the impact of exposure to differentiation on two levels: I first investigate the general effect of exposure to differentiated integration in the EU as a whole, and whether this relationship is moderated by pre-existing Euroscepticism.
+I then test whether the relationship differs according to whether citizens have been exposed primarily to voluntarily chosen or discriminatory differentiated integration.
 
 # Data and methods
+
 ```{r load data}
@@
 
-I investigate support for differentiated integration using multilevel linear models and data from Yougov surveys fielded in 2020-21 [@hemerijck_eui-yougov_2021].
+I investigate support for differentiation using OLS models with country fixed effects and data from Yougov surveys fielded in 2020-2021 [@hemerijck_eui-yougov_2021].
 The sample includes respondents from Denmark, Finland, France, Germany, Sweden, Greece, Hungary, Italy, Lithuania, the Netherlands, Poland, Romania, and Spain and has an *N* of `r nrow(EUI_final_DI_disc)`.
-The surveys include two questions about differentiated integration.
-The first asks whether respondents want an EU where countries may choose to opt out of certain policy integration.
-This is called constitutional differentiation.
+The sample is drawn from the Yougov online panel, with quotas used to make the sample broadly representative of the populations from which they are drawn [@noauthor_yougov_nodate] The surveys include two questions about differentiated integration.
+The first asks whether respondents want an EU where countries may choose to opt out of certain policy integration, so-called constitutional differentiation.
 The second asks whether the EU should allow countries to integrate at different speeds, while maintaining the goal of uniform integration.
 This is an instrumentally differentiated EU.
-This article focuses on popular support for constitutional differentiation, as its allowance for an EU that is permanently segmented into multiple policy communities represents arguably the sharpest departure from the EU’s normative goal of an “ever closer Union”.
+This article focuses on popular support for constitutional differentiation, whose allowance for what @schuessler_mapping_2023 define as sovereignty-enhancing differentiated integration represents the sharpest break with the EU's current goal of an 'ever closer Union'.
 
 The survey data are combined with a novel dataset [@schimmelfennig_eudiff1_2022] that features all differentiations from the EU treaties between 1952-2019.
-The data shows each differentiation a country has been exposed to, the number of articles it involved, and how long it was enforced.
-The data contain information about differentiations from EU treaties, or primary law, rather than differentiations from the less salient secondary law that constitutes the EU’s main legislative output [@duttle_opting_2017].
-I believe the greater salience of primary law differentiation in public debates will lead to a greater knowledge of them among EU citizens than what is the case for differentiations from secondary law.
-They are thus more likely to be important for respondents as they make up their minds about the desirability of differentiated integration.
+The data show each differentiation a country has been exposed to, the number of articles it involved, and how long it was enforced.
+I focus on differentiations from the EU treaties, so-called primary law, rather than exemptions from the less salient secondary law that constitutes the EU's main legislative output [@duttle_opting_2017].
+The reason is that the greater salience of primary law differentiation makes it more likely to be well-known by citizens than more technical secondary legislation.
+Citizens are thus more likely to use it to make up their minds about the desirability of differentiation.
 
 ## Dependent variable
+
 My dependent variable measures support for constitutional differentiated integration.
-The variable uses a five-unit Likert scale and asks respondents whether they support a constitutionally differentiated EU (“Please tell us how far you agree or disagree with the following statement: Member states should be allowed to opt out of specific areas of European integration.
+The variable uses a five-unit Likert scale and asks respondents whether they support a constitutionally differentiated EU ('Please tell us how far you agree or disagree with the following statement: Member states should be allowed to opt out of specific areas of European integration.
 This means that a member state can negotiate exceptions ("opt-out") for areas in which it does not wish to cooperate.
-For example, Denmark has opted out of the common currency, and Poland has opted out of the EU's Charter of Fundamental Rights.)".
-The response categories are recoded, so that 1 becomes Strongly disagree and 5 becomes Strongly agree.
-By not asking about “constitutional differentiated integration”, the question becomes easier to understand.
-It also frames differentiation as a tool for strengthening autonomy and clarifies that countries can choose to opt out.
+For example, Denmark has opted out of the common currency, and Poland has opted out of the EU's Charter of Fundamental Rights.').
+I code the response categories so that 1 is Strongly disagree and 5 is Strongly agree.
+This question frames differentiation as a tool for strengthening autonomy and clarifies that countries can choose to opt out.
+It thus primarily measures what @schuessler_mapping_2023 term sovereignty-enhancing differentiation.
 
 ## Variables of key interest
-To operationalize each country’s exposure to EU differentiation I multiply, for each differentiation, the number of articles that a country was exempt from with the number of years the differentiation was in force.
+
+To operationalize each country's exposure to EU differentiation I multiply, for each differentiation, the number of articles that a country was exempt from with the number of years the differentiation was in force.
 This is then weighted based on the length of time that has passed since the differentiation ceased to be enforced.
@@
 
-This operationalization has three benefits: First, it captures both the temporal and functional scope of a country’s exposure to differentiation.
-Second, the weighting scheme means that those instances of differentiation that were recently in force are given the most weight when investigating the impact of exposure to differentiated integration.
-This is also likely to correspond to how respondents relate their own experience of differentiation to support for it: While many respondents from Central and Eastern Europe may still remember the particulars of their countries’ exposure to externally imposed differentiation, possibly producing a cohort effect [@lauterbach_europe_2020], Southern European respondents are less likely to both remember that they were exposed to externally imposed differentiation several decades ago and use this as a backdrop against which they shape their attitudes towards it.
-Finally, the log-transformation reduces the influence of extreme values, caused by particularly large numbers of differentiations in certain countries.
+This operationalization has three benefits: First, it captures both the temporal and functional scope of a country's exposure to differentiation.
+Second, the weighting scheme ensures that treaty exemptions that were recently in force are given the most weight when correlating exposure to differentiation to support for it.
+This is also likely to correspond to how the survey respondents use their own experiences to form attitudes towards differentiation: While respondents from central and eastern Europe may still remember the particulars of their countries' exposure to externally imposed differentiation, potentially leading to a cohort effect [@lauterbach_europe_2020], southern European respondents are less likely to both remember that they were exposed to externally imposed differentiation several decades ago and relate this to their support for greater use of differentiation today.
+Finally, the log-transformation reduces the influence of outliers in the form of many differentiations from some countries.
 
@@
 I code all Swedish and Danish citizens with a 1 on the Scandinavian dummy.
-To create a Central and Eastern European dummy I do the same for everyone from Hungary, Lithuania, Poland, and Romania.
-These are used to investigate the interaction effects assumed by *H2-H4*.
-I also create a dummy for Euroscepticism, coding those who answer that they would vote yes to leaving the European Union in a hypothetical referendum with a 1 and those answering no with a 0.
+These are used to investigate the interaction effects assumed by *H2*.
+I lastly create a dummy for Euroscepticism, coding those who answer that they would vote yes to leaving the European Union in a hypothetical referendum with a 1 and those answering no with a 0.
+While this only measures what @de_vries_euroscepticism_2018 terms "exit scepticism", an alternative operationalization that uses a measure of satisfaction with EU democracy yields substantively the same results (see Robustness checks).
 
-I operationalize support for liberal economic policies, which has been found by previous literature to be a key determinant of support for differentiated integration [@leuffen_public_2022], through a question that asks whether respondents want to live in a Europe that “stresses economic integration, market competition and fiscal discipline”.
-This is contrasted to “a global Europe that acts as a leader on climate, human rights and global peace” or “a protective Europe that defends the European way of life and welfare against internal and external threats”.
+## Control variables
+
+I operationalize support for liberal economic policies, found by previous literature to be a key determinant of support for differentiated integration [@leuffen_public_2022], through a question that asks whether respondents want to live in a Europe that "stresses economic integration, market competition and fiscal discipline".
+This is contrasted to "a global Europe that acts as a leader on climate, human rights and global peace" or "a protective Europe that defends the European way of life and welfare against internal and external threats".
 I code those who say that they prefer to live in a primarily market-oriented Europe as 1 and others 0 on this variable.
-While the question is multi-faceted, by asking about both fiscal discipline that may be described as austerity policies and more general market integration, I believe it still measures a general affinity for economic liberalism.
+While the question is multi-faceted, by asking about both fiscal discipline that may be described as austerity policies and more general market integration, it appears to still measure a general affinity for liberal economic policies.
 
-I include an equally weighted index of questions measuring respondents’ evaluation of the national economy.
-A factor analysis finds that these questions, in contrast to one asking about the respondent’s perceptions of the local job market, measure the same underlying dimension (see table A1).
-Previous literature shows that positive sociotropic economic evaluations correlate with support for both uniform and differentiated integration (see for instance @gabel_public_1998; @harteveld_europe_2013; @hooghe_calculation_2005; @leuffen_public_2022).
-Egotropic evaluations have, however, been shown to have little relevance for explaining support for differentiated integration [@leuffen_public_2022].
-I measure support for national governments, relevant for explaining support for differentiated integration [@armingeon_loss_2014; @de_blok_blessing_2021; @harteveld_europe_2013], through a dummy asking respondents whether they believe their nation’s government is doing a good job.
+I include an equally weighted index of questions measuring respondents' evaluation of the national economy and their local areas' employment opportunities.
+A factor analysis finds that these questions, in contrast to one asking about the respondents' perceived income relative to others their age and financial security, all measure sociotropic economic evaluations (see table A1).
+Previous literature shows that such evaluations correlate with support for differentiated integration.
+The same does not apply to egotropic economic evaluations [@leuffen_public_2022].
+I measure support for national governments, relevant for explaining support for integration [@armingeon_loss_2014; @de_blok_blessing_2023; @harteveld_europe_2013], through a dummy asking respondents whether they believe their nation's government is doing a good job.
+As an indicator of income, I add an variable surveying respondents' perception of their own economic situation relative to others in their age group.
 
-I operationalize identity through a widely used question that asks respondents to rank the inclusiveness of their identities.
+I operationalize national identities through a widely used question that asks respondents to rank the inclusiveness of their identities.
 I construct a dummy where those answering that they identify solely with their nation-states are coded as 1 and everyone else as 0.
-While nested identities combining a national and European component are common [@risse_euro_2003; @starke_concept_2021], previous literature finds the distinction between exclusively national and partially inclusive identities to be more important for producing Eurosceptic sentiments than variations between different nested identities [@hooghe_calculation_2005].
+While mixed national/European identities are most common [@risse_euro_2003; @starke_concept_2021], previous literature finds the distinction between exclusively national and partially territorially inclusive identities to be more important for producing Eurosceptic sentiments than variations between different forms of mixed identities [@hooghe_calculation_2005].
 
-I also use ideology and gender as socio-political control variables.
+I also use ideology, age and gender as socio-political control variables.
 These are established predictors of support for both general and specific integration (see for example @carrubba_decision_2004; @hobolt_public_2015; @hooghe_calculation_2005; @schoen_identity_2008).
-I operationalise ideology through a 7-unit scale, with 0 representing far-left and 6 far-right positions.
+I operationalize ideology through a scale where 0 represents far-left and 7 far-right positions.
 As previous literature also finds evidence of a curvilinear correlation between ideology and support for the EU [@toshkov_beyond_2022; @van_elsas_changing_2015], I add a squared indicator of left-right ideology.
@@
 data_descriptives <- readRDS("data/Data_descriptives.Rds")
-data_descriptives <- data_descriptives %>%
-  rename(`Support constitutional DI` = opt_out_supp) %>%
-  rename(`Article-years (logged)` = article_disc_log) %>%
-  rename(`Liberal economic values` = market_europe) %>%
-  rename(`Perception of economy` = alternative_econ) %>%
-  rename(`Support national government` = gov_trust) %>%
-  rename(`Left-right` = left_right) %>%
-  rename(`Left-right (sqr.)` = left_right_sqr) %>%
-  rename(`Eurosceptic` = eurosceptic) %>%
-  rename(`Gender` = gender) %>%
-  rename(`Exclusive identity` = excl_identity)
 ```
@@
 ## Limitations
+
 A limitation of the study is that both surveys were fielded during the COVID-19 pandemic.
@@
 However, @cicchi_eu_2020 show that solidarity, as is common, still seemed to reside at the national level in the early phase of the crisis.
-This means that my results may generalize to periods beyond those in which the survey was fielded.
-To account for the different baseline levels of support that are likely to exist in each country, I use multilevel models with random country constants
+My results may thus generalize to periods beyond those in which the surveys were fielded.
+To account for both the different baseline levels of support that are likely to exist in each country and country-specific and unmodelled sources of variation, I use country fixed effects.
 
@@
 Missing data may bias the results if they are correlated with individual-level characteristics like ideology or gender.
-I test the effect of such bias by multiply imputing the data [@buuren_mice_2011].
-I show the results of this test in the section Robustness checks.
+I test the effect of such bias by multiply imputing the data [@buuren_mice_2011] and show the results in the section Robustness checks.
+I impute 25 data sets, and include the dependent variables in the imputation model.
+While there is a debate as to the desirability of imputing dependent variables, a recent simulation study [@kontopantelis_outcome-sensitive_2017] suggests that including dependent variables in the imputation procedure leads to approximately the same results as not doing so.
+It thus strengthens the statistical power of the analysis, without significantly altering the coefficient.
 
-A further limitation, general to all studies using observational survey data, is the challenge of omitted variable bias.
-Here omitted variable bias is particularly likely to stem from the absence of data on the respondents’ knowledge of the EU and education [@karp_know_2003; @marquart_knowing_2019].
+A further limitation, general to all studies using observational survey data, is the potential for omitted variable bias.
+Here omitted variable bias is particularly likely to stem from the absence of data on the respondents' knowledge of the EU and education [@karp_know_2003; @marquart_knowing_2019].
 I use sensitivity analysis [@cinelli_making_2020] to tentatively quantify the size of this bias.
 I show the results of these analyses in the section Robustness checks.
+Lastly, the observational data found in this survey cannot be used to make counter-factual claims in the vein of @pearl_causal_2010.
 
 # Results and discussion
+
 I first show the effect of exposure to differentiation in the broader population, and especially how this interacts with pre-existing Euroscepticism.
-I then discuss how the effect of exposure is moderated by Scandinavian or Central and Eastern European citizenship.
+I then discuss how the effect of exposure is moderated by Scandinavian or central and eastern European citizenship.
 
@@
 main_mod = list()
-main_mod$"Euroscepticism" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
-                                                         eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + (1|country), weights = EUI_final_DI_disc$weight, 
+main_mod$"Euroscepticism" <- lm(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
+                                                         eurosceptic + gender + income_scale + age + excl_identity + article_disc_log*eurosceptic + factor(country), weights = EUI_final_DI_disc$weight,
                                                          data = EUI_final_DI_disc)
 coef_map_eurosceptic <- c('(Intercept)' = 'Constant',
-                          'article_disc_logeurosceptic1' = 'Article-years (logged) X Euroscepticism',
-                          "article_disc_lognordic_dummy1" = "Article-years (logged) X Scandinavia", 
-                          "article_disc_logCEE_dummy1" = "Article-years (logged) X CEE",
+                          'article_disc_log:eurosceptic1' = 'Article-years (logged) X Euroscepticism',
+                          "article_disc_log:nordic_dummy1" = "Article-years (logged) X Scandinavia", 
+                          "article_disc_log:CEE_dummy1" = "Article-years (logged) X CEE",
                           "nordic_dummy1" = "Scandinavia",
@@
                           'gender' = 'Gender',
-                          'excl_identity' = 'Exclusively national identity')
+                          'age' = 'Age',
+                          'income_scale' = 'Income',
+                          'excl_identity1' = 'Exclusively national identity')
 modelsummary(main_mod, fmt = 3, coef_map = coef_map_eurosceptic, output = "latex", 
-             stars = TRUE, title = "Model showing interaction between Euroscepticism and exposure to DI")
+             stars = TRUE, title = "Model showing interaction between Euroscepticism and exposure to DI.
+Fixed country effects and SEs clustered at country level.", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
 ```
 
-Both table 2 and figure 1 strongly support *H1a* and *H1b*, by confirming that exposure to differentiated integration is associated with greater support for differentiation as a future model for the EU.
-We also find that exposure leads to greater levels of support for differentiated integration among Eurosceptics.
-The reason may be that Eurosceptic citizens in countries with greater exposure to differentiated integration are likelier to be familiar with the structure of a differentiated EU membership than those in countries not exposed to any differentiation.
-They may thus be the ones most likely to see such differentiation as a viable solution for achieving the goal of a less integrated EU.
-However, even if the effect is larger for Eurosceptic than non-Eurosceptic citizens, the results suggest that future demands for differentiated integration among those most strongly exposed to it are unlikely to be restricted only to Eurosceptics.
+Both table 2 and figure 1 show that exposure to differentiated integration is associated with greater support for differentiated integration as a future model of integration, and that this effect is particularly prominent among Eurosceptics.
 
-```{r, fig.pos="H", fig.cap="Predicted probability of support for constitutional differentiation as a function of the interaction between article-years (logged) and Euroscepticism. 95% CIs."}
+The large interaction effect between Euroscepticism and exposure to differentiation may be caused by Eurosceptics in countries with greater exposure to it being more familiar with a differentiated EU membership than those in countries without differentiations from the EU treaties.
+The effect of this familiarity is likely to be compounded by the fact that such differentiation may also bring about the less differentiated Europe they are likely to want.
+However, the results hint that the effect of becoming familiar with differentiated integration is not going to be confined to Eurosceptics.
+
+```{r, fig.pos="H", fig.cap="Predicted probability of support for constitutional differentiation as a function of the interaction between article-years (logged) and Euroscepticism.
+95% country-clustered CIs with country fixed effects."}
 library(sjPlot)
 library(ggplot2)
-pred_plot_eurosceptic <- plot_model(main_mod$`Euroscepticism`, type = "pred", terms = c("article_disc_log", "eurosceptic"), 
-                                    colors = "bw", legend.title = "Eurosceptic")
+pred_plot_eurosceptic <- plot_model(main_mod$`Euroscepticism`, type = "int", 
+                                    colors = "bw", vcov.fun = "vcovCL", vcov.args = list(cluster = EUI_final_DI_disc$country))
 pred_plot_eurosceptic_final <-  pred_plot_eurosceptic + ggtitle("Support for constitutional differentiation", 
@@
 
-A plausible counter-argument to my hypothesized causal mechanism could be that differentiation has primarily been chosen where underlying Euroscepticism, and potentially the demand for differentiation, was already high.
-Figure A6 offers some support for this assumption, as we find high levels of differentiated integration in Sweden and Denmark.
-These are countries where opt-outs from integration have been deployed as a response to a Euroscepticism shared by both political parties and voters [@leruth_operationalizing_2015; @spendzharova_banking_2016].
-However, citizens of countries as diverse as Lithuania, Hungary, Italy, and the Netherlands have also been exposed to differentiated integration.
-It is thus not a phenomenon restricted to the countries found in what @bolstad_dynamics_2015 identifies as a Eurosceptic periphery.
-This strengthens the case that the effect is not only driven by an endogenous relationship between Euroscepticism and differentiation.
+A reasonable counter-argument is that governments are more likely to opt for differentiated integration where underlying popular scepticism of integration may make citizens more likely to accept the idea of differentiated integration.
+Figure A6 offers some support for this assumption, as there are particularly high levels of differentiation in for instance Sweden and Denmark.
+These are countries where opt-outs from integration have been deployed as a response to broad Euroscepticism [@leruth_operationalizing_2015; @spendzharova_banking_2016].
+However, citizens of countries as diverse as Lithuania, Hungary, Italy and the Netherlands have also been exposed to differentiated integration.
+It is thus not a phenomenon restricted to the countries found in what @bolstad_dynamics_2015 identify as a Eurosceptic periphery.
+This suggests that the correlations do not solely reflect an endogenous relationship between Euroscepticism and differentiation.
 
-I next analyse how support for constitutional differentiation is shaped by whether differentiation is voluntarily chosen or involuntarily imposed.
-I do this by comparing the interaction effects of Scandinavian and Central and Eastern European citizenship.
+I next analyse whether support for differentiated integration is greater where it has been chosen by national governments and framed as a tool for strengthening autonomy, by comparing the interaction effects of Scandinavian and central and eastern European citizenship.
 
@@
 main_regional = list()
-main_regional$"Regional dummies" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
-                                               eurosceptic + gender + excl_identity + nordic_dummy + CEE_dummy + article_disc_log*nordic_dummy + article_disc_log*CEE_dummy +  (1|country), 
-                                               weights = EUI_final_DI_disc$weight, data = EUI_final_DI_disc)
+main_regional$"Article-years (logged) X regional dummies" <- lm(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
+                          eurosceptic + gender + excl_identity + income_scale + age + nordic_dummy + nordic_dummy*article_disc_log, weights = EUI_final_DI_disc$weight, data = EUI_final_DI_disc)
+
 library(modelsummary)
@@
 coef_map_regions <- c('(Intercept)' = 'Constant',
-                          'article_disc_lognordic_dummy1' = 'Article-years (logged) X Scandinavia',
-                      'article_disc_logCEE_dummy1' = 'Article-years (logged) X CEE',
+                          "article_disc_log:eurosceptic1" = "Article-years (logged) X Eurosceptic",
+                          'article_disc_log:nordic_dummy1' = 'Article-years (logged) X Scandinavia',
+                      'article_disc_log:CEE_dummy1' = 'Article-years (logged) X CEE',
                           'article_disc_log' = 'Article-years (logged)',
@@
                           'gender' = 'Gender',
-                          'excl_identity' = 'Exclusively national identity', 
-                      "article_disc_logeurosceptic1" = "Article-years (logged) X Eurosceptic")
+                          'age' = 'Age',
+                      'income_scale' = 'Income',
+                          'excl_identity1' = 'Exclusively national identity')
+
 modelsummary(main_regional, fmt = 3, output = "latex", coef_map = coef_map_regions, 
-             stars = TRUE, title = "Model showing regional interactions")
+             stars = TRUE, title = "Model showing regional interactions (Fixed effects omitted due to collinearity). SEs clustered at country level.", 
+             vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
 
@@
 
-Table 3 shows that the effect of the interaction between exposure to differentiated integration and Scandinavian citizenship is positive and statistically significant.
-However, the interaction effect between exposure to differentiated integration and Central and Eastern European citizenship is statistically insignificant.
+Table 3 shows that the correlation for the interaction between exposure to differentiation and Scandinavian citizenship is positive and statistically significant.
+This is also supported by the multiply imputed robustness tests (table A3).
 
@@
 #Scandinavia
-pred_plot_nordics <- plot_model(main_regional$`Regional dummies`, type = "pred", terms = c("article_disc_log", "nordic_dummy"), 
-                                legend.title = "Scandinavia", colors = "bw")
+pred_plot_nordics <- plot_model(main_regional$"Article-years (logged) X regional dummies", type = "pred", terms = c("article_disc_log", "nordic_dummy"), colors = "bw", 
+                                vcov.fun = "vcovCL", vcov.args = list(cluster= EUI_final_DI_disc$country))
+
 pred_plot_nordics_final <-  pred_plot_nordics + ggtitle("Scandinavia") + labs(x = "Differentiation (article-years (logged)",
@@
                            labels = c("Non-Scandinavian", "Scandinavian"))
-##CEE
-pred_plot_CEE <- plot_model(main_regional$`Regional dummies`, type = "pred", terms = c("article_disc_log", "CEE_dummy"), 
-                                    colors = "bw", legend.title = "CEE")
-pred_plot_CEE_final <-  pred_plot_CEE + ggtitle("Central and Eastern Europe") + labs(x = "Differentiation (article-years (logged))",
-                                                                                                                     y = "Predicted value") + scale_linetype_manual(name = "CEE",
-                           values = c("dashed", "solid"),
-                           labels = c("Non-CEE", "CEE"))
-pred_plot_CEE_final <- pred_plot_CEE_final + font_size(title = 15, labels.x = 15, labels.y = 15)
 ```
 
-```{r, fig.pos="H", fig.cap="Variations in support for differentiation as a function of exposure to DI (A = Scandinavia and B = CEE)", fig.height=8.5}
+```{r, fig.pos="H", fig.cap="Variations in support for differentiation as a function of exposure to DI (A = Scandinavia and B = CEE).
+95% country-clustered CIs.
+Fixed effects omitted due to multicollinearity.", fig.height=8.5}
 library(patchwork)
-regional_final <- pred_plot_nordics_final/pred_plot_CEE_final + plot_annotation(tag_levels = "A", title = "Support for constitutional differentiation", subtitle = "Predicted values for regional interactions of exposure to differentiation in Scandinavia and Central and Eastern Europe", theme = theme(plot.title = element_text(size = 16)))
-regional_final
+
+pred_plot_nordics_final
 ```
 
-Figure 2 supports this picture: Scandinavian citizens appear to become more favourably inclined towards differentiated integration as their exposure to it increases compared to those from other regions.
-However, this effect is only present for a certain range of the measure of exposure.
-Thus, I can confirm *H4a*.
-However, for a large range of the exposure variable the difference between Scandinavian and non-Scandinavian citizens is statistically indistinguishable from zero.
-
-However, the predicted values show no significant difference between support for differentiation in Central and Eastern Europe and the rest of the EU.
-This leads me to reject *H2a*.
-Neither do I find support for *H3*, as there is no statistically significant greater positive effect of exposure in Central and Eastern Europe.
+Figure 2 shows that Scandinavian citizens appear to become more favourably inclined towards differentiated integration as their exposure to it increases, in line with the assumption that voluntarily chosen differentiated integration will lead to a more strongly positive correlation with exposure to differentiated integration than what is found in rest of the EU.
+I thus confirm *H2a*.
 
 ```{r}
-region_mod <- readRDS("data/region_mod.Rds")
-modelsummary::modelsummary(region_mod, fmt = 3, output = "latex", coef_map = coef_map_eurosceptic, 
-             stars = TRUE, title = "Models showing interactions of Euroscepticism and region")
+region_mod = list()
+
+nordic_data <- EUI_final_DI_disc %>%
+  dplyr::filter(nordic_dummy == 1)
+
+CEE_data <- EUI_final_DI_disc %>%
+  dplyr::filter(CEE_dummy == 1)
+
+region_mod$"Scandinavia" <- lm(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr + income_scale + age +
+                                                                              eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + factor(country), 
+                                                                              weights = nordic_data$weight, data = nordic_data)
+
+modelsummary::modelsummary(region_mod, fmt = 3, output = "latex", coef_map = coef_map_regions, 
+             stars = TRUE, title = "Models showing interactions of Euroscepticism and region.
+Fixed country effects and SEs clustered at country-level", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
 ```
@@
                            labels = c("Non-Eurosceptic", "Eurosceptic"))
-##CEE
-pred_plot_CEE_eurosceptic <- plot_model(region_mod$CEE, type = "pred", terms = c("article_disc_log", "eurosceptic"), 
-                                legend.title = "Eurosceptic", colors = "bw")
-pred_plot_CEE_eurosceptic_final <-  pred_plot_CEE_eurosceptic + ggtitle("Central and Eastern Europe") + labs(x = "Differentiation (article-years (logged))",
-                                                                                                                           y = "Predicted value")
-pred_plot_CEE_eurosceptic_final <- pred_plot_CEE_eurosceptic_final +  scale_linetype_manual(name = "Eurosceptic",
-                           values = c("dashed", "solid"),
-                           labels = c("Non-Eurosceptic", "Eurosceptic"))
 ```
 
-```{r, fig.pos="H", fig.cap="Interaction between Euroscepticism and region. 95% CIs", fig.height=8.5}
+```{r, fig.pos="H", fig.cap="Interaction between Euroscepticism and region. 95% country-clustered CIs and country fixed effects.", fig.height=8.5}
 library(patchwork)
-regional_euroscepticism_final <- pred_plot_nordics_eurosceptic/pred_plot_CEE_eurosceptic_final + plot_annotation(tag_levels = "A", title = "Support for constitutional differentiation", subtitle = "Interaction of Euroscepticism and exposure to differentiation in CEE and Scandinavia", theme = theme(plot.title = element_text(size = 16)))
-regional_euroscepticism_final
+pred_plot_nordics_eurosceptic
 ```
 
-I also reject *H2b*, as the effect of the interaction between differentiated integration and Euroscepticism is statistically insignificant in the Central and Eastern Europe data.
-Lastly, I reject *H4b*, as there is no statistically significant difference between the effect of exposure for Eurosceptics and supporters of EU membership in Scandinavia.
-The lack of a lower level of support for differentiated integration among Central and Eastern Europeans, even if the discriminatory imposition of differentiation has been prevalent in these countries, suggests that historical experience with differentiated integration may shape support for it mainly when it is used as a part of elite discourse.
-This finding is important for our understanding of how attitudes towards greater differentiation can be shaped by political elites in the future.
+I also confirm *H2b*, as Scandinavian Eurosceptics appear more favourably inclined towards differentiated integration than those who support EU membership.
+This suggests that how differentiated integration is framed, and especially if it is framed as a source of greater national sovereignty, is crucial for whether it produces greater support or not.
 
-The fact that the surveys were fielded at the height of the COVID-19 pandemic means that we may be seeing lower levels of support for differentiated integration than we otherwise would: As @genschel_postfunctionalism_2021 argue, the COVID-19 pandemic “debordered” solidarity, making the prospect of mutual aid across member state borders more attractive to EU citizens.
+The fact that the surveys were fielded at the height of the COVID-19 pandemic means that we may be seeing lower levels of support for differentiated integration than would otherwise be the case: As @genschel_postfunctionalism_2021 argue, the COVID-19 pandemic potentially "debordered" solidarity, making the prospect of mutual aid across member state borders more attractive to EU citizens.
 In times where EU solidarity is framed as a net positive for its citizens, support for differentiation may be lower than it otherwise would have been.
@@
 # Robustness checks
-I use an approach outlined in @cinelli_making_2020 to test the sensitivity of several of my models to omitted variable bias (see figures A1-4).
-This quantifies the variance that a potential confounder would need to account for to nullify the effect of exposure to differentiated integration.
-A test of the sensitivity of the interaction between Euroscepticism and exposure to differentiation in Scandinavia (figure A3) shows that a variable explaining 50% of the remaining variance of support for constitutional differentiation would need to explain less than 10% of the remaining variance in real exposure to differentiation to nullify its effect.
+
+I use an approach outlined in @cinelli_making_2020 to test the sensitivity of several models to omitted variable bias (see figures A1-4).
+The test quantifies the variance that a confounding variable would need to account for to shrink the effect of exposure to differentiated integration to zero.
+A test of the sensitivity of the interaction between Euroscepticism and exposure to differentiation in Scandinavia (figure A3) shows that a variable explaining 50% of the remaining variance of support for constitutional differentiation would need to explain less than 10% of the remaining variance in real exposure to differentiation to nullify the effect of the variable.
 One important omitted variable is knowledge of the EU [@leuffen_public_2022].
 While knowledge of the EU is unlikely to be strongly connected with actual levels of differentiation, its strong connection to support for differentiation makes its omission particularly problematic.
-Omitted variable bias is thus likely to be an analytical concern throughout the entire study.
-Future studies must mitigate these concerns by including knowledge questions about current EU politics, to allow researchers to include a control similar to that found in other EU public opinion research.
+This shows that omitted variable bias is a real concern for this study.
+Future studies should mitigate this by explicitly measuring knowledge of EU politics, to allow researchers to include a control similar to that found in other EU public opinion research.
 
-I also use multiple imputation through chained equations [@buuren_mice_2011; @fjelstul_explaining_2022] to test how sensitive the results are to bias caused by potentially systematic data missingness.
-Multiple imputation better accounts for the uncertainty associated with data imputation than traditional mean or zero imputation methods, and provides more realistic standard errors.
-This is particularly beneficial when testing the statistical significance of multiply imputed estimates.
-As appendix tables A2-3 show, the multiply imputed models are substantively similar to those found in tables 2-4.
-This suggests that the effect sizes or directions are not driven by systematic data missingness.
+I also use multiple imputation [@buuren_mice_2011] to test how sensitive the results are to bias caused by potentially systematic data missingness.
+By pooling the estimates and standard errors from 25 data imputations according to "Rubin's rules" [@rubin_2004_multiple] I better account for the uncertainty associated with imputation of missing values.
+This is particularly beneficial when conducting significance test.
+The effects of the multiple imputation suggests that there may be reason to think that systematic missingness is at play: Table A2 suggests that the effect of the interaction between exposure to differentiation and Euroscepticism loses its significance when the data are imputed.
+Similarly, we find that there is no longer a significant interaction effect between exposure and Euroscepticism in Scandinavia (table A4).
+However, the interaction effect of exposure and Scandinavian citizenship appears strongly robust to imputation.
 
-I also test whether the substantial overlap between Scandinavian Eurosceptics and non-Eurosceptics is driven by the very low variability, with a standard deviation of `r round(sd(nordic_data$article_disc_log), digits = 2)`,  in the measure of exposure to differentiated integration found in data from this region.
+I also test whether the substantial overlap between Scandinavian Eurosceptics and non-Eurosceptics is driven by the very low variability, with a standard deviation of `r round(sd(nordic_data$article_disc_log), digits = 2)`, of the measure of exposure to differentiated integration found in the Scandinavian data.
 I do this by testing a three-way interaction between Scandinavian citizenship, exposure to differentiated integration and Euroscepticism in the full data.
-Appendix table A6 suggests that the results are substantially similar to the interaction effect of Scandinavian citizenship and Euroscepticism shown in table 4.
-The results thus suggest that the estimates are not unduly biased by the very small standard deviation found in this particular region.
+Appendix table A6 shows that the results are substantially similar to the interaction effect of Scandinavian citizenship and Euroscepticism shown in table 4.
+The results thus suggest that the estimates are not unduly biased by the small variability found in the Scandinavian data.
 
-Lastly, I test how robust the results are to an alternative operationalization of Euroscepticism: satisfaction with EU democracy.
-Dissatisfaction with the state of EU democracy may be seen as a form of "regime scepticism", characterized by a dissatisfaction with EU decision-making, rather than the "exit scepticism" expressed by a desire to leave the union altogether [@vries_euroscepticism_2018].
-Figure A5 (coefficients in table A4) shows similar trends for both operationalizations: Those least satisfied with EU democracy, and who thus express the greatest amount of regime scepticism, also express greater support for constitutional differentiated integration than those more satisfied with EU democracy.
-Their support for differentiation also increases more strongly as a function of exposure to previous differentiation than what is the case for those more satisfied with EU democracy.
-This strengthens the robustness of my original conclusions.
+Lastly, I test how robust the results are to an alternative operationalization of Euroscepticism: dissatisfaction with EU democracy.
+This may be seen as "regime scepticism", characterized by a dissatisfaction with the democratic quality of EU decision-making, rather than the "exit scepticism" expressed by a desire to leave the union altogether [@de_vries_euroscepticism_2018].
+Figure A5 shows that the results are broadly similar for both operationalizations: Those least satisfied with EU democracy also express greater support for constitutional differentiated integration than those satisfied with EU democracy.
+In other words, both strong and less strong rejection of the EU's democratic functioning leads to greater support for differentiated integration.
 
@@
 
-This paper establishes, by combining data surveying citizen preferences for differentiated integration with data on their countries’ actual exposure to differentiated integration, that there is a strong association between support for differentiated integration and past exposure to it.
-This may be because citizens more accustomed to differentiated integration may see it as a favourable way of combining the benefits of national membership of the EU with greater national autonomy.
-Importantly, this effect appears moderated by whether one opposes or supports EU membership.
-However, the results do not suggest that the mode of implementation of the differentiation that the same citizens have been exposed to play any large role in shaping attitudes towards a future differentiated EU.
+This paper establishes, by combining data surveying citizen preferences for differentiated integration with information about their countries' actual exposure to differentiation, that there may be a strong association between support for differentiated integration and past exposure to it.
+This may be because citizens more accustomed to differentiated integration have become more favourably inclined towards it.
+Importantly, the effect appears moderated by whether one opposes or supports EU membership.
+It also seems to be the case that exposure to frames depicting differentiated integration as a source of greater autonomy could be an important driver of support for differentiated integration
 
-What this suggests is that even if some differentiated integration could be perceived as a cartelization of the benefits of integration that also constrains the political autonomy of those subject to it, this does not necessarily translate to less support for differentiated integration as a policy.
-While my article does not establish why this is the case, it does strongly suggest that the positive or negative effect of historical exposure to differentiated integration on support for it in the future depends on whether political elites mobilize this discourse as a part of their efforts to shape attitudes towards the question.
-Understanding what kind of discourses are more likely to produce critical attitudes towards differentiated integration is thus an important question for future research.
+The fact that even Eurosceptics become more favourably inclined to differentiated integration the more they are exposed to it, despite a continued membership of the EU being out of line with what they actually want, suggests that differentiated integration may be seen as an acceptable "second-best" alternative to an EU exit even among those considering themselves opponents of membership.
+This furthermore strengthens the assumption that differentiated integration, as has been pointed to in previous literature [@bellamy_republican_2019; @schimmelfennig_ever_2020], can serve to mitigate the contestation of European integration brought by the EU's expansion into more important areas.
 
-Another question is whether the public is equally likely to prefer differentiation in all policy areas.
-Existing research [@de_blok_blessing_2021; @telle_elite_2022; @leuffen_public_2022] has largely focused on individual support for differentiated integration of the EU as a polity.
-It is both possible and probable that even those favouring an “ever closer Union” may have diverging preferences for integration at the level of individual policies.
-Testing where individual citizens may prefer more differentiation may also help theorize the ambivalent attitudes towards EU integration found in previous EU public opinion literature [@de_vries_variable_2013; @goldberg_one_2021; @moland_constraining_2022].
+Despite our increasing knowledge of what drives individual support for differentiated integration, we still do not know whether citizens equally likely to prefer differentiation in all policy areas.
+Existing research [@de_blok_blessing_2023; @telle_elite-mass_2022; @leuffen_public_2022] has largely focused on individual support for differentiated integration of the EU as a polity.
+It is both possible and probable that even those favouring an "ever closer Union" may perceive integration as more or less problematic depending on the policy.
+Testing where individual citizens may prefer more differentiation may also help theorize the ambivalent attitudes towards EU integration found in previous EU public opinion literature [@de_vries_variable_2013; @boomgaarden_mapping_2011; @moland_constraining_2022].
 
 My results have policy implications for both the EU and other international organizations seeking to deepen their own political integration.
-They suggest first that allowing for a greater institutionalization of differentiated integration may undermine the public legitimacy of the EU’s goal of ‘ever closer Union’.
-However, beyond the EU, the results also suggest that the public legitimacy of deeper political integration may be challenged if citizens gradually become accustomed to more differentiated and non-uniform forms of integration.
-While one should be wary of generalizing findings from the EU to other international organizations, my results suggest that allowing some countries to opt out of the legal order that is increasingly developing also beyond the EU [@zurn_theory_2018] may lead to greater popular opposition to the idea that countries should be members of entire frameworks of political cooperation, rather than more functionally limited partnerships dedicated to solving particular problems.
+They suggest first that allowing for a greater institutionalization of differentiated integration may undermine the public legitimacy of the EU's goal of 'ever closer Union'.
+Beyond the EU, the results also suggest that the public legitimacy of deeper political integration may be challenged if citizens gradually become accustomed to more differentiated deepening of political cooperation.
+While one should be wary of generalizing findings from the EU to other international organizations, my results suggest that allowing some countries to opt out of the legal order that is increasingly developing also beyond the EU may lead to greater popular opposition to the idea that countries should band together as coherent political orders rather than functionally demarcated intergovernmental organizations dedicated to maintaining traditional national sovereignty.
 
 # References
-<div id="refs"></div>
+
+::: {#refs}
+:::

### 2022-11-02 Manuscript_3_draft1.Rmd
Additions made to the appendix as well as well as rewrites of the main text.

@@
 Combining recent survey data with data showing country-level exposure to differentiated integration, I show that familiarity with differentiation seems to breed greater support for it.
-This particularly applies to Eurosceptics.
+This especially applies to Eurosceptics.
 It also suggests that support for differentiated integration may be higher where governments frame it as a source of greater national autonomy.
@@
 Despite the EU being one of the world's most advanced political cooperations, its integration has become increasingly fragmented after the ratification of the 1992 Maastricht Treaty.
-The differentiation of EU integration has primarily been driven by a desire to deepen political cooperation even where it may be contested by some of the EU's members [@leuffen_differentiated_2013; @schimmelfennig_ever_2020], and has typically taken two forms: First, some member states have opted out of cooperation to protect their own sovereignty.
+This fragmentation has been driven by a desire to deepen political cooperation even where it may be contested by some of the EU's members [@leuffen_differentiated_2013; @schimmelfennig_ever_2020], and has typically taken two forms: First, some member states have opted out of cooperation to protect their own sovereignty.
 This is called "constitutional differentiation".
-Second, "instrumental differentiation" has happened where countries with a diminished capacity for integration have been allowed to integrate at different schedules.
+Second, "instrumental differentiation" happens where countries with a diminished capacity for integration have been allowed to integrate at different schedules.
 Such differentiation can either happen voluntarily or because the EU has imposed differentiation as a precondition for membership [@schimmelfennig_instrumental_2014].
-Thus, 'variable geography' [@stubb_categorization_1996] better describes the EU's political development than the EU's own goal of "ever closer Union".
+Thus, 'variable geography' [@stubb_categorization_1996] better describes the EU's current development than its own normative goal of "ever closer Union".
 
-While increasing attention is being paid to what shapes popular preferences for such differentiation [@leuffen_public_2022; @telle_elite-mass_2022; @schuessler_mapping_2023], we know little about how actual exposure to it has shaped support for the goal of a uniformly integrated EU.
-While one of the few existing contributions examining the effect of differentiation on public opinion shows that allowing for differentiation may strengthen the perceived democratic legitimacy of the EU [@schraff_does_2020], important knowledge gaps remain.
-This article tests three theoretically informed hypotheses: First, it tests whether support for differentiation leads to greater support for it on average.
-Second, it tests whether those more critical of EU membership become more positively inclined to differentiated integration if exposed to more of it.
-Lastly, it tests whether those citizens exposed to differentiated integration that was voluntarily chosen by their governments are more favourably inclined towards future differentiation.
+While increasing attention is paid to what shapes popular preferences for such differentiation [@leuffen_public_2022; @telle_elite-mass_2022; @schuessler_mapping_2023; @de_blok_blessing_2023], we know little about how actual exposure to it has contributed to changing attitudes towards differentiation.
+While one of the few existing contributions examining the effect of differentiation on public opinion shows that allowing for differentiation may strengthen the democratic legitimacy of the EU [@schraff_does_2020], important knowledge gaps remain.
+This article tests two theoretically informed hypotheses: First, it tests whether exposure to differentiation leads to greater support for it on average, especially where the non-uniformity is voluntarily chosen.
+Second, it tests whether this effect differs by already existing attitudes towards EU membership.
 
-By combining data from surveys fielded in 2020-2021 that measure citizens' attitudes towards constitutional differentiated integration in 13 EU member states with data detailing every country's exposure to differentiated integration, I find that support for constitutional differentiation is associated with greater exposure to differentiations from EU law, but that this effect is strongly contingent on individual-level Euroscepticism.
+By combining data from surveys fielded in 13 EU member states in 2020-2021 measuring citizens' attitudes towards constitutional differentiated integration with data detailing their countries' exposure to differentiated integration, I find that support for constitutional differentiation is associated with greater exposure to differentiations from EU law, but that this effect is strongly contingent on individual-level Euroscepticism.
 I also find that support for it seems to be greater where differentiated integration has been framed as a tool for strengthening national sovereignty.
-This shows that familiarity with differentiated integration can lead to greater support for it in the future.
-However, more research is needed to understand what causes regional variations in support for differentiated integration.
+This suggests both that familiarity with differentiated integration can lead to greater support for it in the future, and that elite discourses matter for forming attitudes towards it.
 
-In addition to expanding our knowledge of public support for differentiated EU integration, this article contributes to a broader literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009].
-While the EU's legal and functional reach goes beyond that of most other international organization, many international organizations have seen their legal and political powers expand in recent years [@zurn_theory_2018].
-Even if more research is needed to understand whether my results travel to contexts beyond the EU, they help illuminate how allowing countries to opt out of international rule-making shapes attitudes towards the future organization of the international legal order.
+My findings are also important to the discussion about the EU's future spawned by Brexit [@european_commission_white_2017]: The results suggest that the EU's differentiated integration may facilitate future integration of political powers traditionally considered core to statehood [@genschel_differentiated_2013].
+This is made particularly clear by how even those who oppose EU membership appear particularly supportive of differentiated integration.
+That even those critical of EU membership may be favourably inclined towards the general principle of European integration as long as it allows countries to exercise autonomy when faced with the prospect of particularly controversial integration is an important finding for policy-makers seeking a way forward for the post-Brexit EU.
 
-My findings are also important to the discussion about the EU's future spawned by Brexit [@european_commission_white_2017]: Because EU integration has expanded into increasingly salient areas, differentiation has often been used to reduce public contestation of integration [@schimmelfennig_ever_2020].
-By providing greater numbers of citizens with exposure to differentiation, the EU's institutions may solidify opposition to further integration among greater numbers of citizens, potentially stalling future attempts at cooperation [@hooghe_postfunctionalist_2009].
-In this sense, the EU's differentiated integration may be both a help and hindrance for future integration: On the one hand, it may help facilitate future integration of political powers traditionally considered core to statehood [@genschel_differentiated_2013].
-On the other, greater popular support for differentiated modes of integration can make it more difficult for the EU's institutions to get the public's buy-in to functionally necessary integration.
+However, greater exposure to differentiated modes of integration can also complicate the EU's future integration if it leads to greater opposition to today's normative goal of 'ever closer Union'.
 The inevitable trade-off that this entails must be reckoned with by EU policy-makers deciding whether to allow for opt-outs from policy integration or to insist upon legal uniformity.
@@
 
-An under-investigated question is why some people support a more differentiated EU, which renders the EU's integration either permanently or temporarily non-uniform.
+An under-investigated question is why some people support differentiated integration, that renders the EU's integration permanently or temporarily non-uniform.
 Differentiated integration is already investigated conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004].
-The few contributions investigating individual support for differentiated integration have focused on liberal economic values [@leuffen_public_2022], Euroscepticism [@de_blok_blessing_2023; @schuessler_mapping_2023], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite-mass_2022].
-Thus, with one notable exception, the effect of national contexts are under-studied.
-Such national contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; @harteveld_europe_2013].
-Here citizens' exposure to past differentiated integration plays a crucial role: Differentiations from EU law may create a disequilibrium between a country's actual level of integration and what its citizens would preferences.
+The few contributions investigating individual support for differentiated integration focus on liberal economic values [@leuffen_public_2022], Euroscepticism [@de_blok_blessing_2023; @schuessler_mapping_2023], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite-mass_2022].
+However, we still do not know what impact exposure to differentiated integration in the past has on support for it as a future model of EU integration.
+Knowing whether exposure to differentiated integration has shaped attitudes towards it is important for two reasons: First, it can tell us whether the EU's current strategy of using differentiated integration to facilitate deeper EU integration [@schimmelfennig_ever_2020] has the counter-intuitive effect of solidifying support for a less integrated EU.
+Secondly, understanding whether exposure to differentiated integration leads to varying degrees of acceptance of it among opponents or supporters of EU membership is important because it can show whether differentiated integration can be used as a tool for reducing the contestation of European integration among citizens who are particularly critical of the EU.
+
+It is theoretically plausible that exposure to differentiated integration will shape public support for it: Differentiations from EU law may create a disequilibrium between a country's actual level of integration and what its citizens would prefer.
 The same citizens may thus respond by changing their preferences [@busemeyer_beyond_2021; @wlezien_public_1995].
-While there have been many studies of the feedback loop between public opinion and the EU's ordinary policy-making [@bolstad_dynamics_2015; @de_bruycker_democratically_2020], we do not know how this feedback loop plays out when the policies being made are in essence alterations to the EU's institutional structure.
+While the feedback loop between public opinion and the EU's ordinary policy-making has been extensively studied [@bolstad_dynamics_2015; @de_bruycker_democratically_2020], we do not know how this feedback loop plays out when the policies being made are alterations to the EU's institutional framework.
 
-It is reasonable to assume that the greater familiarity with differentiated integration that exposure to it will bring will lead to greater support for it in the future: Knowledge of EU policies is often, though not always, associated with greater support for EU integration [@marquart_knowing_2019; @verhaegen_does_2015].
-Similarly, familiarity with more differentiated ways of structuring EU memberships may lead to a stronger perception among citizens that EU memberships that allow countries to opt of policy integration may be less problematic.
+It is also reasonable to assume that the greater familiarity with differentiated integration that exposure to it will bring will lead to greater support for it in the future: Knowledge of EU policies is often, though not always, associated with greater support for EU integration, even if higher levels of general knowledge of EU politics do not always translate to either higher levels of knowledge of specific policies [@marquart_knowing_2019; @karp_know_2003; @verhaegen_does_2015].
+As @karp_know_2003 point out, the theoretical mechanism that we would expect to be at play in this case is that greater familiarity with the EU and its policies would lead citizens to perceive EU integration as less of a threat than if they had no knowledge of the policy in question.
+Thus, someone who supports EU membership without having been exposed to differentiated integration might find what can only be perceived as a deviation from the uniformity that is frequently touted as a normative goal of the EU's integration problematic, and wonder how it would impact their membership of the EU.
+However, someone familiar with differentiated integration might be less worried about how such differentiation could impact their countries, as they have first-hand experience with how differentiated integration can combine both greater autonomy with a similar level of influence within the EU as what they had before the implementation of differentiated integration.
 
 This positive relationship is likely to be more strongly felt for those that oppose their country's EU membership.
-Since constitutional differentiated integration allows for permanent opt-outs from policy integration, Eurosceptics who favour leaving the EU may see differentiated integration as a potentially more realistic path towards the less integration Europe that they prefer.
+For them, the perception that differentiated integration can better combine EU membership with national autonomy may be particularly salient, because they would be particularly likely to favour modes of integration that preserve a sovereignty they are particularly likely to believe to be important.
 This yields two hypotheses:
@@
 
-Responses to differentiated integration may also vary depending on the situation in which differentiated integration was imposed.
-It is for instance likely that citizens may become more favourably inclined towards differentiated integration if the form of differentiated integration they have been exposed to was a result of a conscious decision to opt out on the part of their own governments.
-The reason is that these decisions may be more likely to be seen by citizens as an exercise of their own national sovereignty within the remits of an EU membership.
+It is also theoretically reasonable to assume that citizens may become more favourably inclined towards differentiated integration if the differentiation was brought about as a result of a conscious decision to opt out of policy integration.
+This would especially be the case if the opt-out was brought about through political debate, as this would both make the issue of opt-outs salient for individual citizens and make it more likely that citizens would be exposed to discourses that frame differentiated integration as a tool for exercising greater autonomy.
 
-Scandinavia offers an interesting case study of this assumption: Here differentiation has been chosen by national governments and framed as a tool for strengthening national sovereignty [@adler-nissen_opting_2014; @leruth_operationalizing_2015; @spendzharova_banking_2016], and has typically been decided as a result of a referendum.
-Previous literature has found that the renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy among Danish Eurosceptics [@schraff_does_2020].
-This shows that Scandinavian citizens may theoretically respond to differentiated integration, because of the particularly voluntary nature of how differentiated integration has been implemented in these regions, by wanting more differentiated integration in the future.
+Scandinavia offers an interesting case study of this assumption: Here differentiation has been chosen by national governments and framed as a tool for strengthening national sovereignty [@adler-nissen_opting_2014; @leruth_operationalizing_2015], and has typically been decided as a result of a referendum following public debate.
+Previous literature has also found a causal effect of the act of recommitting to differentiated integration, as the renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy among Danish Eurosceptics [@schraff_does_2020].
+The fact that citizens of the region have responded positive to renewals of opt-outs
 
-Both the way that differentiated integration has been framed and the particular national situation of the Scandinavian countries makes it likely that there will be an even stronger positive reaction to differentiation among Scandinavian Eurosceptics: Because their countries are both relatively small and highly dependent on European market access, Eurosceptic Scandinavians could see differentiation as a viable second-best alternative to the European disintegration that they otherwise favour.
-This effect is likely to be compounded by how Scandinavian elites have framed differentiated integration as a tool for greater sovereignty.
+Both the way that differentiated integration has been framed and the particular national situation of the Scandinavian countries makes it likely that there will be an even stronger positive reaction to differentiation among Scandinavian Eurosceptics: Because their countries are both relatively small and highly dependent on European market access, Eurosceptic Scandinavians could be more likely to see differentiation as a viable second-best alternative to the European disintegration that they otherwise favour.
+This effect is likely to be compounded by how Scandinavian elites have framed differentiated integration as a tool for greater sovereignty [@adler-nissen_opting_2014].
 I thus hypothesize:
@@
 
-To analyse the regionally specific effects of differentiation, I create dummies for Scandinavian and Central and Eastern European citizenship.
+To analyse the regionally specific effects of differentiation, I create a dummy for Scandinavian citizenship.
 I code all Swedish and Danish citizens with a 1 on the Scandinavian dummy.
 These are used to investigate the interaction effects assumed by *H2*.
-I lastly create a dummy for Euroscepticism, coding those who answer that they would vote yes to leaving the European Union in a hypothetical referendum with a 1 and those answering no with a 0.
+I also create a dummy for Euroscepticism, coding those who answer that they would vote yes to leaving the European Union in a hypothetical referendum with a 1 and those answering no with a 0.
 While this only measures what @de_vries_euroscepticism_2018 terms "exit scepticism", an alternative operationalization that uses a measure of satisfaction with EU democracy yields substantively the same results (see Robustness checks).
@@
 library(modelsummary)
+library(flextable)
+library(dplyr)
 table_format <- ifelse(is_docx, "huxtable", 'default')
+
+#desc_tab <- modelsummary::datasummary_skim(data_descriptives, fmt = 3, histogram = FALSE, 
+                 #title = "Descriptive statistics", output = "flextable")
+#desc_tab %>%
+  #autofit()
+
 modelsummary::datasummary_skim(data_descriptives, fmt = 3, histogram = FALSE, 
@@
 I show the results of these analyses in the section Robustness checks.
-Lastly, the observational data found in this survey cannot be used to make counter-factual claims in the vein of @pearl_causal_2010.
+Lastly, the observational data found in this survey cannot be used to make counter-factual claims, as these are discussed in for instance @pearl_causal_2010.
 
@@
                           'excl_identity1' = 'Exclusively national identity')
+
+#analysis1_tab <- modelsummary(main_mod, fmt = 3, coef_map = coef_map_eurosceptic, output = "flextable", 
+ #            stars = TRUE, title = "Model showing interaction between Euroscepticism and exposure to DI.
+Fixed country effects and SEs clustered at country level.", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
+
+#analysis1_tab %>%
+ # autofit()
+
 modelsummary(main_mod, fmt = 3, coef_map = coef_map_eurosceptic, output = "latex", 
@@
 
-Both table 2 and figure 1 show that exposure to differentiated integration is associated with greater support for differentiated integration as a future model of integration, and that this effect is particularly prominent among Eurosceptics.
+Both table 2 and figure 1 show that exposure to differentiated integration is associated with greater support for differentiated integration, and that this effect is particularly prominent among Eurosceptics.
 
@@
 A reasonable counter-argument is that governments are more likely to opt for differentiated integration where underlying popular scepticism of integration may make citizens more likely to accept the idea of differentiated integration.
-Figure A6 offers some support for this assumption, as there are particularly high levels of differentiation in for instance Sweden and Denmark.
-These are countries where opt-outs from integration have been deployed as a response to broad Euroscepticism [@leruth_operationalizing_2015; @spendzharova_banking_2016].
+Figure A5 offers some support for this assumption, as there are particularly high levels of differentiation in for instance Sweden and Denmark.
+These are countries where opt-outs from integration have been deployed as a response to broad Euroscepticism [@leruth_operationalizing_2015; @adler-nissen_opting_2014].
 However, citizens of countries as diverse as Lithuania, Hungary, Italy and the Netherlands have also been exposed to differentiated integration.
@@
 
+#analysis2_tab <- modelsummary(main_regional, fmt = 3, output = "flextable", coef_map = coef_map_regions, 
+ #            stars = TRUE, title = "Model showing regional interactions (Fixed effects omitted due to collinearity). SEs clustered at country level.", 
+  #           vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
+
+#analysis2_tab %>%
+ # autofit()
+
 modelsummary(main_regional, fmt = 3, output = "latex", coef_map = coef_map_regions, 
@@
             vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
-
 ```
@@
 
-```{r, fig.pos="H", fig.cap="Variations in support for differentiation as a function of exposure to DI (A = Scandinavia and B = CEE).
+```{r, fig.pos="H", fig.cap="Variations in support for differentiation as a function of exposure to DI between Scandinavians and non-Scandinavians.
 95% country-clustered CIs.
-Fixed effects omitted due to multicollinearity.", fig.height=8.5}
+Fixed effects omitted due to multicollinearity.", fig.height=4}
 library(patchwork)
@@
 
+#nordic_analysis_tab <- modelsummary::modelsummary(region_mod, fmt = 3, output = "flextable", coef_map = coef_map_regions, 
+ #            stars = TRUE, title = "Models showing interactions of Euroscepticism and region.
+Fixed country effects and SEs clustered at country-level", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
+
+#nordic_analysis_tab %>%
+ # autofit()
+
 modelsummary::modelsummary(region_mod, fmt = 3, output = "latex", coef_map = coef_map_regions, 
-             stars = TRUE, title = "Models showing interactions of Euroscepticism and region.
-Fixed country effects and SEs clustered at country-level", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
+   stars = TRUE, title = "Models showing interactions of Euroscepticism and region. Fixed country effects and SEs clustered at country-level", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
 ```
@@
 
-I use an approach outlined in @cinelli_making_2020 to test the sensitivity of several models to omitted variable bias (see figures A1-4).
+I use an approach outlined in @cinelli_making_2020 to test the sensitivity of several models to omitted variable bias (see figures A1-3).
 The test quantifies the variance that a confounding variable would need to account for to shrink the effect of exposure to differentiated integration to zero.
@@
 I also use multiple imputation [@buuren_mice_2011] to test how sensitive the results are to bias caused by potentially systematic data missingness.
-By pooling the estimates and standard errors from 25 data imputations according to "Rubin's rules" [@rubin_2004_multiple] I better account for the uncertainty associated with imputation of missing values.
+By pooling the estimates and standard errors from 25 data imputations according to "Rubin's rules" [@rubin_2004_multiple] I better account for the uncertainty associated with imputing missing values.
 This is particularly beneficial when conducting significance test.
-The effects of the multiple imputation suggests that there may be reason to think that systematic missingness is at play: Table A2 suggests that the effect of the interaction between exposure to differentiation and Euroscepticism loses its significance when the data are imputed.
-Similarly, we find that there is no longer a significant interaction effect between exposure and Euroscepticism in Scandinavia (table A4).
+The effects of the multiple imputation suggests that systematic missingness may be an issue: Table A2 shows that the effect of the interaction between exposure to differentiation and Euroscepticism loses its significance when the data are imputed.
+Similarly, we no longer find a significant interaction effect between exposure and Euroscepticism in Scandinavia (table A4).
 However, the interaction effect of exposure and Scandinavian citizenship appears strongly robust to imputation.
 
-I also test whether the substantial overlap between Scandinavian Eurosceptics and non-Eurosceptics is driven by the very low variability, with a standard deviation of `r round(sd(nordic_data$article_disc_log), digits = 2)`, of the measure of exposure to differentiated integration found in the Scandinavian data.
+I also test whether the substantial overlap between Scandinavian Eurosceptics and non-Eurosceptics is driven by the low variability, with a standard deviation of `r round(sd(nordic_data$article_disc_log), digits = 2)`, of the exposure to differentiated integration found in the two Scandinavian countries.
 I do this by testing a three-way interaction between Scandinavian citizenship, exposure to differentiated integration and Euroscepticism in the full data.
-Appendix table A6 shows that the results are substantially similar to the interaction effect of Scandinavian citizenship and Euroscepticism shown in table 4.
-The results thus suggest that the estimates are not unduly biased by the small variability found in the Scandinavian data.
+Appendix table A7 shows that the results are substantially similar to the interaction effect of Scandinavian citizenship and Euroscepticism shown in table 4.
+The results thus suggest that the estimates are not unduly biased by the variability found in the Scandinavian data.
+
+To test the scope of the established effect, I restrict my analysis to differentiated integration among the policies @genschel_more_2016 term "core state powers", state powers that are essential to state functioning.
+A reasonable assumption might be that citizens could be especially inclined towards more differentiation if exposed to differentiations in the field of core state power integration, as these are areas where citizens might find it important to retain national sovereignty [@hooghe_postfunctionalist_2009; @de_wilde_can_2012].
+My results, visualized by predicted values (figure A7), suggest that policies in question matter for the effect of differentiation: While the effect of exposure to differentiated core state power integration seems largely flat among supporters of EU membership, it increases with exposure to differentiation among Eurosceptics.
+However, the effect seems largely comparable to those visualized in figure 1.
+While the reason for this may be that much differentiated integration has happened in the field of core state powers [@schimmelfennig_ever_2020], it also suggests that the effects are robust to alternative operationalizations of the measure of exposure to differentiated integration.
 
@@
 This paper establishes, by combining data surveying citizen preferences for differentiated integration with information about their countries' actual exposure to differentiation, that there may be a strong association between support for differentiated integration and past exposure to it.
-This may be because citizens more accustomed to differentiated integration have become more favourably inclined towards it.
+The reason may be that those accustomed to differentiated integration have become more favourably inclined towards it because they have learnt from experience how it can combine membership of the EU with greater exercise of national autonomy.
 Importantly, the effect appears moderated by whether one opposes or supports EU membership.
-It also seems to be the case that exposure to frames depicting differentiated integration as a source of greater autonomy could be an important driver of support for differentiated integration
+It also seems as if exposure to frames depicting differentiated integration as a source of greater autonomy could be an important driver of support for differentiated integration
 
-The fact that even Eurosceptics become more favourably inclined to differentiated integration the more they are exposed to it, despite a continued membership of the EU being out of line with what they actually want, suggests that differentiated integration may be seen as an acceptable "second-best" alternative to an EU exit even among those considering themselves opponents of membership.
-This furthermore strengthens the assumption that differentiated integration, as has been pointed to in previous literature [@bellamy_republican_2019; @schimmelfennig_ever_2020], can serve to mitigate the contestation of European integration brought by the EU's expansion into more important areas.
+The fact that Eurosceptics become more favourably inclined to differentiated integration the more they are exposed to it, even if they primarily want an exit from the European Union, suggests that differentiated integration may be seen as an acceptable "second-best" alternative to an EU exit even among those considering themselves opponents of membership.
+This strengthens the assumption that differentiated integration, as is pointed to in previous literature [@bellamy_republican_2019; @schimmelfennig_ever_2020], can mitigate the contestation of European integration brought by the EU's expansion into more important areas.
 
-Despite our increasing knowledge of what drives individual support for differentiated integration, we still do not know whether citizens equally likely to prefer differentiation in all policy areas.
-Existing research [@de_blok_blessing_2023; @telle_elite-mass_2022; @leuffen_public_2022] has largely focused on individual support for differentiated integration of the EU as a polity.
-It is both possible and probable that even those favouring an "ever closer Union" may perceive integration as more or less problematic depending on the policy.
-Testing where individual citizens may prefer more differentiation may also help theorize the ambivalent attitudes towards EU integration found in previous EU public opinion literature [@de_vries_variable_2013; @boomgaarden_mapping_2011; @moland_constraining_2022].
+Despite our increasing knowledge of what drives individual support for differentiated integration, we still do not know whether citizens are equally likely to prefer differentiation in all policy areas.
+Existing research [@de_blok_blessing_2023; @telle_elite-mass_2022; @leuffen_public_2022] has largely focused on individual support for differentiated EU integration.
+Even those favouring an "ever closer Union" may perceive integration as more or less problematic depending on the policy.
+Testing where individual citizens may prefer more differentiation may also help theorize the ambivalent attitudes towards EU integration found in previous EU public opinion literature [@de_vries_variable_2013; @boomgaarden_mapping_2011].
 
-My results have policy implications for both the EU and other international organizations seeking to deepen their own political integration.
+My results also have policy implications for the EU and other international organizations that seek deeper political integration.
 They suggest first that allowing for a greater institutionalization of differentiated integration may undermine the public legitimacy of the EU's goal of 'ever closer Union'.
 Beyond the EU, the results also suggest that the public legitimacy of deeper political integration may be challenged if citizens gradually become accustomed to more differentiated deepening of political cooperation.
-While one should be wary of generalizing findings from the EU to other international organizations, my results suggest that allowing some countries to opt out of the legal order that is increasingly developing also beyond the EU may lead to greater popular opposition to the idea that countries should band together as coherent political orders rather than functionally demarcated intergovernmental organizations dedicated to maintaining traditional national sovereignty.
+While one should not necessarily generalize findings from the EU context to other international organizations, my findings suggest that letting some countries opt out of international political integration may lead to greater opposition to the idea that countries should band together as transnational political orders rather than intergovernmental organizations built around traditional national sovereignty.
 

### 2022-11-02 appendix.Rmd
Additions made to the appendix as well as well as rewrites of the main text.

@@
 ---
-title: "APPENDIX: 'Does differentiation beget differentiation?'"
+title: "Online supplementary material: 'Does differentiation beget differentiation?'"
 author:
@@
 params:
-  anonymous: ""
+  anonymous: TRUE
   doublespacing: ""
@@
 nordic_data <- readRDS("data/nordic_data.Rds")
-CEE_data <- readRDS("data/CEE_data.Rds")
+
 appendix_items$"Fixed effect of interacted model (Nordics)" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
@@
 
-```{r, fig.cap="Sensitivity to omitted variable bias for article-years (logged) X Euroscepticism in Central and Eastern Europe", fig.height=8.5}
-library(sensemakr)
-appendix_items$"Fixed effect of interacted model (CEE)" <- lm(opt_out_supp ~ article_disc_log + alternative_econ + gov_trust + left_right + left_right_sqr +
-                                                                    eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + factor(country), weights = CEE_data$weight, 
-                                                                  data = CEE_data)
-appendix_items$"Sensitivity test of interacted model (CEE)" <- sensemakr(appendix_items$`Fixed effect of interacted model (CEE)`, 
-                                                                             treatment = "article_disc_log:eurosceptic1")
-plot(appendix_items$`Sensitivity test of interacted model (CEE)`, type = "extreme")
-```
-
 ```{r, results='asis'}
@@
 library(merTools)
+library(broom)
+library(broom.helpers)
 main_impute <- readRDS("data/main_impute.Rds")
@@
 coef_map_regions2 <- c('(Intercept)' = 'Constant',
-                          'article_disc_log:nordic_dummy' = 'Article-years (logged) X Scandinavia',
-                      'article_disc_log:CEE_dummy' = 'Article-years (logged) X CEE',
+                          'article_disc_log:nordic_dummy1' = 'Article-years (logged) X Scandinavia',
+                      'article_disc_log:CEE_dummy1' = 'Article-years (logged) X CEE',
                       "article_disc_log:eurosceptic1" = "Article-years (logged) X Eurosceptic",
@@
                           'left_right' = 'Left-right',
+                      'age' = "Age",
+                      'income_scale' = "Income",
                           'left_right_sqr' = 'Left-right (sqr)',
@@
 
-modelsummary::modelsummary(main_impute, fmt = 3, output = "latex", coef_map = coef_map_regions2, 
-             stars = TRUE, title = "Imputed model of interaction between Euroscepticism and article-years (logged) and model with regional dummies")
+modelsummary::modelsummary(main_impute$`Main model (Euroscepticism)`, fmt = 3, output = "latex", coef_map = coef_map_regions2, 
+             stars = TRUE, title = "Imputed model of interaction between Euroscepticism and article-years (logged). Fixed country effects and SEs clustered at country level.")
 ```
@@
                           'excl_identity' = 'Exclusively national identity')
-modelsummary::modelsummary(regional_impute, fmt = 3, output = "latex", coef_map = coef_map_eurosceptic2, 
-             stars = TRUE, title = "Imputed models showing interactions of Euroscepticism and article-years (logged) by region")
+modelsummary::modelsummary(regional_impute$`Article-years (logged) X regional dummies`, fmt = 3, output = "latex", coef_map = coef_map_regions2, 
+             stars = TRUE, title = "Imputed models showing interactions of regional dummies and article-years (logged) by region.
+Fixed effects omitted due to collinearity.
+SEs clustered at country level.")
+```
 
+```{r, results='asis'}
+region_mod <- readRDS("data/region_mod.Rds")
+coef_map_eurosceptic2 <- c('(Intercept)' = 'Constant',
+                          'article_disc_log:eurosceptic1' = 'Article-years (logged) X Euroscepticism',
+                          'article_disc_log' = 'Article-years (logged)',
+                          'market_europe' = 'Liberal economic values',
+                          'alternative econ' = 'Perception of economy',
+                          'gov_trust' = 'Support national government',
+                          'left_right' = 'Left-right',
+                          'age' = "Age", 
+                          'income_scale' = "Income",
+                          'left_right_sqr' = 'Left-right (sqr)',
+                          'eurosceptic1' = 'Eurosceptic',
+                          'gender' = 'Gender',
+                          'excl_identity' = 'Exclusively national identity')
+modelsummary::modelsummary(region_mod, fmt = 3, output = "latex", coef_map = coef_map_regions2, 
+             stars = TRUE, title = "Imputed models showing interactions of Euroscepticism and article-years (logged) by region") %>% kableExtra::kable_styling(latex_options = "scale_down")
+```
+
+```{r} 
 coef_map_alteurosceptic <- c('(Intercept)' = 'Constant',
-                          'article_disc_logeu_democ_satisf' = 'Article-years (logged) X Euroscepticism',
+                          'article_disc_log:eu_democ_satisf' = 'Article-years (logged) X Satisfaction with EU democracy',
                           'article_disc_log' = 'Article-years (logged)',
@@
                           'gender' = 'Gender',
+                          'age' = "Age", 
+                          'income_scale' = "Income",
                           'excl_identity' = 'Exclusively national identity')
@@
 modelsummary::modelsummary(appendix_items$`Alternative conceptualization of Euroscepticism`, fmt = 3, output = "latex", coef_map = coef_map_alteurosceptic, 
-             stars = TRUE, title = "Model showing alternative conceptualization of Euroscepticism")
+             stars = TRUE, title = "Model showing alternative conceptualization of Euroscepticism", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
 ```
@@
 pred_plot_interact_main <- plot_model(appendix_items$`Alternative conceptualization of Euroscepticism`, type = "pred", terms = c("article_disc_log", "eu_democ_satisf"), 
-                                      colors = "bw", legend.title = "EU democracy satisfaction")
+                                      colors = "bw", legend.title = "EU democracy satisfaction", vcov.fun = "vcovCL", vcov.args = list(cluster = EUI_final_DI_disc$country))
+
 pred_plot_interact_main_final <-  pred_plot_interact_main + ggtitle("Support for constitutional differentiation", 
@@
 ```{r, results='asis'}
-southern_mod = list()
-southern_mod$"Southern Europe" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
-                           eurosceptic + gender + excl_identity + south_dummy + south_dummy*article_disc_log + (1|country), weights = EUI_final_DI_disc$weight, 
+appendix_items$"Southern Europe" <- lm(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
+                           eurosceptic + gender + age + income_scale + excl_identity + south_dummy + south_dummy*article_disc_log, weights = EUI_final_DI_disc$weight, 
                          data = EUI_final_DI_disc)
+
 coef_map_southern <- c('(Intercept)' = 'Constant',
-                          'article_disc_logsouthern_dummy1' = 'Article-years (logged) X Southern Europe',
+                          'article_disc_log:south_dummy' = 'Article-years (logged) X Southern Europe',
                           'article_disc_log' = 'Article-years (logged)',
@@
                           'gender' = 'Gender',
+                       'age' = "Age", 
+                       'income_scale' = "Income",
                           'excl_identity' = 'Exclusively national identity', 
-                      "south_dummy1" = "Southern Europe")
+                      "south_dummy" = "Southern Europe")
 
-modelsummary::modelsummary(southern_mod, fmt = 3, output = "latex", coef_map = coef_map_alteurosceptic, 
-             stars = TRUE, title = "Model showing effect of exposure in Southern Europe")
+modelsummary::modelsummary(appendix_items$`Southern Europe`, fmt = 3, output = "latex", coef_map = coef_map_southern, 
+             stars = TRUE, title = "Model showing effect of exposure in Southern Europe (fixed effects omitted due to collinearity", vcov = sandwich::vcovCL(appendix_items$`Southern Europe`, ~country), gof_map = c("nobs", "adj.r.squared"))
 ```
@@
 ```{r, fig.pos="H", fig.cap= "Support for differentiated integration in Southern Europe as an interaction between Euroscepticism and exposure to differentiated integration. 95% CIs.", fig.height=8.5}
-pred_plot_southern <- plot_model(southern_mod$`Southern Europe`, type = "pred", terms = c("article_disc_log", "south_dummy"), 
-                                 colors = "bw", legend.title = "Southern Europe")
+pred_plot_southern <- plot_model(appendix_items$`Southern Europe`, type = "pred", terms = c("article_disc_log", "south_dummy"), 
+                                 colors = "bw", legend.title = "Southern Europe", vcov.fun = "vcovCL", vcov.args = list(cluster = EUI_final_DI_disc$country))
 pred_plot_southern_final <-  pred_plot_southern + ggtitle("Support for constitutional differentiation (Southern Europe)", subtitle = "Support for differentiation among Southern and non-Southern Europeans") + labs(x = "Differentiation (article-years (logged))",
@@
 scandi_eurosceptic = list()
-scandi_eurosceptic$"Scandinavia (three-way interactions)" <- lmer(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
-                             eurosceptic + gender + excl_identity + nordic_dummy + article_disc_log*eurosceptic*nordic_dummy + (1|country), weights = EUI_final_DI_disc$weight, 
+scandi_eurosceptic$"Scandinavia (three-way interactions)" <- lm(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
+                             eurosceptic + gender + excl_identity + age + income_scale + nordic_dummy + article_disc_log*eurosceptic*nordic_dummy + factor(country), weights = EUI_final_DI_disc$weight,
                            data = EUI_final_DI_disc)
+```
+
+```{r}
+
 coef_map_nordic <- c('(Intercept)' = 'Constant',
-                          'article_disc_logeurosceptic1nordic_dummy1' = 'Article-years (logged) X Eurosceptic X Scandinavia',
+                          'article_disc_log:eurosceptic1:nordic_dummy1' = 'Article-years (logged) X Eurosceptic X Scandinavia',
                           'article_disc_log' = 'Article-years (logged)',
                           'nordic_dummy1' = "Scandinavia",
+                          'market_europe' = 'Liberal economic values',
+                          'alternative econ' = 'Perception of economy',
+                          'gov_trust' = 'Support national government',
+                     'age' = "Age", 
+                     'income_scale' = "Income",
+                          'left_right' = 'Left-right',
+                          'left_right_sqr' = 'Left-right (sqr)',
+                          'eurosceptic1' = 'Eurosceptic',
+                          'gender' = 'Gender',
+                          'excl_identity' = 'Exclusively national identity')
+
+modelsummary::modelsummary(scandi_eurosceptic$"Scandinavia (three-way interactions)", fmt = 3, output = "latex", coef_map = coef_map_nordic,
+             stars = TRUE, title = "Model showing three-way interaction between Scandinavian dummy, Eurosceptic and article-years (logged)", vcov = sandwich::vcovCL(scandi_eurosceptic$"Scandinavia (three-way interactions)", ~country), gof_map = c("nobs", "adj.r.squared"))
+```
+
+```{r}
+
+main_mod_CSP = list()
+
+main_mod_CSP$"Euroscepticism" <- lm(opt_out_supp ~ CSP_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
+                                                         eurosceptic + gender + income_scale + age + excl_identity + CSP_disc_log*eurosceptic + factor(country), weights = EUI_final_DI_disc$weight, data = EUI_final_DI_disc)
+
+coef_map_CSP <- c('(Intercept)' = 'Constant',
+                          'CSP_disc_log:eurosceptic1' = 'Article-years (logged) X Euroscepticism',
+                          "CSP_disc_log:nordic_dummy1" = "Article-years (logged) X Scandinavia", 
+                          "CSP_disc_log:CEE_dummy1" = "Article-years (logged) X CEE",
+                          "nordic_dummy1" = "Scandinavia",
+                          "CEE_dummy1" = "Central and Eastern Europe",
+                          'CSP_disc_log' = 'Article-years (logged)',
                           'market_europe' = 'Liberal economic values',
@@
                           'gender' = 'Gender',
-                          'excl_identity' = 'Exclusively national identity')
-modelsummary::modelsummary(scandi_eurosceptic, fmt = 3, output = "latex", coef_map = coef_map_nordic,
-             stars = TRUE, title = "Model showing three-way interaction between Scandinavian dummy, Eurosceptic and article-years (logged)")
+                          'age' = 'Age',
+                          'income_scale' = 'Income',
+                          'excl_identity1' = 'Exclusively national identity')
+
+modelsummary::modelsummary(main_mod_CSP$"Euroscepticism", fmt = 3, output = "latex", coef_map = coef_map_CSP,
+             stars = TRUE, title = "Model showing interaction between Euroscepticism and DI in the field of core state powers", vcov = sandwich::vcovCL(main_mod_CSP$"Euroscepticism", ~country), gof_map = c("nobs", "adj.r.squared"))
 ```
+
+```{r, fig.pos="H", fig.cap="Predicted probability of support for constitutional differentiation as a function of the interaction between article-years (logged) in the field of core state powers and Euroscepticism.
+95% country-clustered CIs with country fixed effects."}
+library(sjPlot)
+library(ggplot2)
+pred_plot_eurosceptic_CSP <- plot_model(main_mod_CSP$`Euroscepticism`, type = "int", 
+                                    colors = "bw", vcov.fun = "vcovCL", vcov.args = list(cluster = EUI_final_DI_disc$country))
+pred_plot_eurosceptic_CSP_final <-  pred_plot_eurosceptic_CSP + ggtitle("Support for constitutional differentiation (exposure to CSP DI)", 
+                                                                subtitle = "Support for differentiation among Eurosceptics and non-Eurosceptics") + labs(x = "Differentiation (article-years (logged))",
+                                                                                                                     y = "Predicted value") + scale_linetype_manual(name = "Eurosceptic",
+                           values = c("dashed", "solid", "dotted"),
+                           labels = c("Non-Eurosceptic", "Eurosceptic"))
+pred_plot_eurosceptic_CSP_final <- pred_plot_eurosceptic_CSP_final + font_size(title = 15, labels.x = 15, labels.y = 15)
+pred_plot_eurosceptic_CSP_final
+```
